---
title: "apicalis"
author: "Annie"
date: "2024-06-20"
output: html_document
---

# This document uses the paleoview climate model and code written by July

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Libraries
```{r libraries,}
# Load necessary libraries for various tasks including data manipulation, spatial analysis, plotting, and machine learning
library(broom)
library(car)
library(DALEX)
library(dials)
library(dplyr)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
```

#Set the working directory
```{r Set the working directory}

knitr::opts_knit$set(root.dir =("C:/github/SNR_SDM/Data/processed"))
```

Read the 'apicalis.csv' file, print its contents for verification, and convert it to an sf (simple feature) object
Set the Coordinate Reference System (CRS) to EPSG:4326, which corresponds to GDA2020


```{r read file}
setwd("C:/github/SNR_SDM/Data/processed")
apicalis <- read_csv("apicalis.csv") %>%
  {print(.); .} %>%  # Print the data frame to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>% # Convert to sf object using longitude and latitude as coordinates
  st_set_crs(4326)   # Set the CRS to EPSG:4326 (GDA2020)

```

# Create an extent object for Australia using specific minimum and maximum X and Y coordinates (longitude and latitude)
# Object type: Extent (from the 'terra' package)

```{r}
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)
```

#Land masks
***Run Incorporate paleoview variables code file here***

##current land mask from paleoview
Create a single raster from the first layer of the 'masks' list = current landmask
```{r}
#PV_land_mask_t0 <- masks[[1]] #paleoview land mask
```


**Krapp2021**
##Create a land mask for the present time using the 'Krapp2021' dataset from the 'pastclim' package
# Object type: SpatRaster (spatial raster), where X = longitude, Y = latitude, and Z = time
```{r}
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>% 
  aggregate(fact = 5, fun = "mean", na.rm = T)
```
```{r}
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>%  # Crop the raster to the extent of Australia
  aggregate(fact = 5, fun = "mean", na.rm = T)  # Aggregate to reduce resolution for faster processing (mean values, ignoring NAs)
```

# Plot apicalis occurrence using ggplot2, with the current land mask as the background

# Load ggplot2 to visualize apicalis occurrence and current land mask

```{r Plot the data using ggplot2}
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask raster as the base layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask raster
  geom_sf(data = apicalis, aes(col = time_bp))  # Overlay apicalis occurrences, coloring points by 'time_bp' column
```
# Read in Paleoview climate variables from a NetCDF file
```{r read in palaoview variables}
climate_full <- sds("C:/github/SNR_SDM/Data/processed/paleoview/climate_full.nc")
```

# Generate pseudo-absences for apicalis data, focusing on occurrences between 20,000 and 90 years ago
```{r}
apicalis2 <- sample_pseudoabs_time(apicalis %>% 
                                     filter(time_bp == -1000), # Filter apicalis occurrences older than 20,000 years
                                           
                                   n_per_presence = 3, # Generate 3 pseudo-absences per presence point
                                   raster = climate_full[[1]], # Use the first layer of climate data as the reference raster
                                   time_col = "time_bp", # Specify the time column
                                   lubridate_fun = pastclim::ybp2date, # Convert years before present (ybp) to date
                                   method = c("dist_min", km2m(150))) # Ensure pseudo-absences are at least 150 km away
```
# Plot the generated pseudo-absences (apicalis2) with the current land mas
```{r ggplot}
ggplot() +
  geom_spatraster(data = land_mask_current) +  # Add the land mask raster
  scale_fill_terrain_c() +  # Use a terrain color scale for the raster
  geom_sf(data = apicalis2, aes(col = class))  # Overlay apicalis2 occurrences, coloring points by the 'class' column

```
#retrieve climate data for this region

```{r extract climate for region}
apicalis_df <- apicalis2 %>% # Convert apicalis2 to a data frame and extract the spatial coordinates
  dplyr::bind_cols(sf::st_coordinates(apicalis2)) %>% # Bind spatial coordinates (longitude, latitude) to the data
  mutate(time_bp = date2ybp(time_step)) %>%  # Convert the time_step to 'years before present' (ybp)
  as.data.frame() %>% # Convert the result to a standard data frame
  select(-geometry) # Drop the geometry column since it is no longer needed

# Extract climate data for the apicalis occurrences using the location_slice_from_region_series function
apicalis_df <- location_slice_from_region_series(apicalis_df,
                                                 region_series = climate_full) # Retrieve climate data from 'climate_full'
```

# Add climate reconstructions to the sf object and remove the time_step column
# Only include the relevant climate variables (bio05, bio06, bio12)
# Filter out any rows where bio05 is missing, as we need it for modeling
```{r add climate to sf}
apicalis4 <- apicalis2 %>%
  bind_cols(apicalis_df[, c("bio05", "bio06", "bio12")]) %>%# Add climate data columns to the apicalis sf object
  select(-time_step) %>% # Remove time_step as it's not needed for modeling
  filter(!is.na(.$bio05)) # Filter out rows with missing bio05 values
```

# Set up a recipe for modeling

 Next, we need to set up a recipe to define how to handle our dataset. We don’t want to transform our data, so we just need to define the formula (class is the outcome, all other variables are predictors; note that, for sf objects, geometry is automatically ignored as a predictor):


```{r make recipe}
apicalis_rec <- recipe(apicalis4, formula = class ~ .) # Define the recipe for the model
apicalis_rec # Check the recipe


# We can check the variables in the recipe with:

apicalis_rec$var_info # This will show us the variables included in the model
```
We now build a workflow_set of different models, defining which hyperparameters we want to tune. We will use glm, gam, random forest and boosted trees as our models, so only random forest and boosted trees have tunable hyperparameters. For the most commonly used models, tidysdm automatically chooses the most important parameters, but it is possible to fully customise model specifications.

```{r specify models}
apicalis_models <- workflow_set(
  preproc = list(default = apicalis_rec),  # Use the recipe we defined as the pre-processing step
  models = list(
    glm = sdm_spec_glm(),  # Standard GLM specs (no parameters to tune)
    gam = sdm_spec_gam(),  # Standard GAM specs (no parameters to tune)
    rf = sdm_spec_rf(),    # Random forest with tunable hyperparameters
    gbm = sdm_spec_boost_tree()  # Boosted tree model (gbm) with tunable hyperparameters
  ),
  cross = TRUE  # Create all combinations of pre-processing and models
) %>%
  # Update the workflow for GAMs, specifying the formula
  update_workflow_model("default_gam",
                        spec = sdm_spec_gam(),
                        formula = gam_formula(apicalis_rec)  # Specify the formula for GAMs
  ) %>%
  # Adjust the controls to store the necessary information for creating the ensemble model later
  option_add(control = control_ensemble_grid())


```
Note that gams are unusual, as we need to specify a formula to define to which variables we will fit smooths. By default, gam_formula() fits a smooth to every continuous predictor, but a custom formula can be provided instead.


We now want to set up a spatial block cross-validation scheme to tune and assess our models:

# Set up spatial block cross-validation with 5 folds (v = 5)
# This ensures that spatial structure is considered during model training and evaluation
 
```{r set up cross validation}
library(tidysdm)
set.seed(1005) # Set seed for reproducibility
apicalis_cv <- spatial_block_cv(apicalis4, v = 5) # Create 5 spatial blocks for cross-validation
autoplot(apicalis_cv)  # Visualize the spatial cross-validation blocks
```

We can now use the block CV folds to tune and assess the models:

```{r tune models}
set.seed(123) # Set seed for reproducibility
apicalis_models <- apicalis_models %>%
  workflow_map("tune_grid",                # Perform hyperparameter tuning using a grid search
               resamples = apicalis_cv,    # Use the spatial block CV folds for resampling
               grid = 5,                   # Use 5 points in the grid for hyperparameter tuning
               metrics = sdm_metric_set(), # Use the default set of SDM metrics
               verbose = TRUE)             # Print progress updates during tuning

# We can visualize the performance of the tuned models using autoplot:
autoplot(apicalis_models)  # Plot the model performance across different hyperparameters

```


# Create an ensemble by selecting the best set of parameters for each model
# The Boyce continuous index will be used to select the best-performing random forest and boosted tree models

```{r ensemble}
apicalis_ensemble <- simple_ensemble() %>%
  add_member(apicalis_models, metric = "boyce_cont")  # Add members to the ensemble based on the Boyce index

autoplot(apicalis_ensemble)  # Visualize the ensemble model and its performance

```
We can now make predictions with this ensemble (using the default option of taking the mean of the predictions from each model) for the Last Glacial Maximum (LGM, 21,000 years ago).

```{r create LGM}
# Create a climate raster for the LGM, including the climate variables bio05, bio06, and bio12
climate_lgm <- rast(list(climate_full[[1]][[1]], climate_full[[2]][[1]], climate_full[[3]][[1]]))  # Select the relevant layers from the climate data
names(climate_lgm) <- c("bio05", "bio06", "bio12")  # Rename the layers to match the climate variables used in the model
```

And predict using the ensemble:

```{r predict lgm}
# Use the ensemble model to make predictions for the LGM climate conditions
prediction_lgm <- predict_raster(apicalis_ensemble, climate_lgm)  # Predict the probability of occurrence for the LGM

# Visualize the prediction results using ggplot2
ggplot() +
  geom_spatraster(data = prediction_lgm, aes(fill = mean)) +  # Plot the predicted occurrence probabilities as a spatial raster
  scale_fill_terrain_c(name = "Probability\nof occurrence")  # Use a terrain color scale to represent the probability
```

Let's make a whole stack of predictions and write the maps to a folder.

```{r stack o maps}
# Make predictions for multiple time steps and save the maps to a specified folder
walk(1:1046, function(timestep) {
  # Extract the year before present (BP) by subtracting 2000 from the climate time index
  year_bp <- flatten_int(time(climate_full))[timestep] - 2000

  # Extract the climate data for the given timestep (bio05, bio06, bio12)
  climate_snapshot <- rast(list(climate_full[[1]][[timestep]], 
                                climate_full[[2]][[timestep]], 
                                climate_full[[3]][[timestep]]))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename the layers to match the predictors

  # Make predictions using the ensemble model for the current climate snapshot
  prediction1 <- predict_raster(apicalis_ensemble, climate_snapshot)

  # Plot the prediction using ggplot2, including a title with the corresponding year BP
  p <- ggplot() +
    geom_spatraster(data = prediction1, aes(fill = mean)) +
    scale_fill_terrain_c(name = "Probability\nof occurrence") +
    ggtitle(paste0("Paleoview model apicalis Years BP: ", year_bp))  # Add the title showing the year BP

  # Save the plot as a PNG file in the specified folder, naming it according to the year BP
  ggsave(path = "C:/github/SNR_SDM/Results/2apicalispaleoview",
         plot = p,
         filename = paste0("2apicalis_map_", year_bp, ".png"))
})
```
# Create an explainer object for the ensemble model
# This object will allow us to generate variable importance and partial dependence profiles
```{r PDP}
explainer_apicalis_ens <- explain_tidysdm(apicalis_ensemble)

# Compute variable importance using model_parts()
# This shows how much each variable contributes to the model predictions
vip_ensemble <- model_parts(explainer = explainer_apicalis_ens)
plot(vip_ensemble)  # Plot the variable importance results

# Generate a partial dependence profile for the variable bio12
# N specifies the number of observations used to compute the profile
pdp_bio12 <- model_profile(explainer_apicalis_ens, N = 500, variables = "bio12")
plot(pdp_bio12)  # Plot the partial dependence profile for bio12
```
```{r PDP bio05}
explainer_apicalis_ens <- explain_tidysdm(apicalis_ensemble) # Create an explainer object for the ensemble model

vip_ensemble <- model_parts(explainer = explainer_apicalis_ens)
plot(vip_ensemble) # Plot the variable importance results

pdp_bio05 <- model_profile(explainer_apicalis_ens, N = 500, variables = "bio05")
plot(pdp_bio05)  # Plot the partial dependence profile for bio05
```
```{r PDP bio06}
explainer_apicalis_ens <- explain_tidysdm(apicalis_ensemble)

vip_ensemble <- model_parts(explainer = explainer_apicalis_ens)
plot(vip_ensemble)

pdp_bio06 <- model_profile(explainer_apicalis_ens, N = 500, variables = "bio06")
plot(pdp_bio06)
```

  



```
