---
title: "apicalis"
author: "Annie"
date: "2024-06-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r libraries,}
# Load necessary libraries for various tasks including data manipulation, spatial analysis, plotting, and machine learning
library(broom)
library(car)
library(DALEX)
library(dials)
library(dplyr)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
```


```{r Set the working directory}

knitr::opts_knit$set(root.dir =("C:/github/SNR_SDM/Data/processed"))
```
Read in the 'apicalis.csv' file, print its contents, convert to sf, and set the CRS
```{r read file}
setwd("C:/github/SNR_SDM/Data/processed")
apicalis <- read_csv("apicalis.csv") %>%
  {print(.); .} %>%  # Print the data frame to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>%
  st_set_crs(4326)  # Set the Coordinate Reference System (CRS) to GDA2020, which has the EPSG code 4326

```

Create an extent object for Australia using s min and max X and Y coordinates
Object type: Extent (terra package)
```{r}
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)
```

______  LAND MASKS
```{r Import new landmask }
# Define the directory containing the landmask files and list all files that match the regular expression ".+ka_land.asc"
masks <- "../Data/raw/landmask/" %>% 
  
   # List all files in the specified directory that match the regular expression
  dir_ls(regexp = ".+ka_land.asc") %>%  
 
  # Sort the filenames in a human-readable order
  gtools::mixedsort() %>%  
  
  # Select the first 24 files from the sorted list
  .[1:24] %>%  
  
   # Load each selected file as a raster
  map(rast) %>% 
  
  # Set the CRS for each raster to WGS84
  map(~ project(.x, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")) %>%  
  
  # Reduce the resolution of each raster by a factor of 10
  map(aggregate, fact = 10) %>% 
  
  # Crop each raster to the extent of Australia defined by Aust_extent
  map(crop, y = Aust_extent)  

# Combine the individual rasters into a single SpatRaster object
SpatRast_land_mask <- rast(masks)

```
Create a land mask for the present time using the 'Krapp2021' dataset from the 'pastclim' package
Object type: SpatRaster where X =longitude, Y = latitude, Z = time
```{r creat current landmask for Austrlia}
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent))
```
Create a single raster from the first layer of the 'masks' list = current landmask
```{r creat current landmask}
# lower resolution option
land_mask_current2 <- masks[[1]]
```



___Use ggplots to plot apicalis occurrence with current land mask
```{r Plot the data using ggplot2}
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_sf(data = apicalis, aes(col = time_bp))  # Add the 'apicalis2' sf object as spatial features, colouring by the 'time_bp' column
```


```{r read in palaoview variables}
climate_full <- readRDS("C:/github/SNR_SDM/Data/processed/paleoview/climate_full.rds")
```
```{r}
apicalis <- sample_pseudoabs_time(apicalis |>
                                     #filter apicalis to within 21ka
                                     filter(time_bp > -21000, time_bp < -100),
                                   n_per_presence = 3,
                                   raster = climate_full,
                                   time_col = "time_bp",
                                   lubridate_fun = pastclim::ybp2date,
                                   method = c("dist_min", km2m(70)))

ggplot() +
  geom_spatraster(data = land_mask, aes(fill = land_mask_0)) +
  scale_fill_terrain_c() +
  geom_sf(data = apicalis, aes(col = class))

```




