---
title: "conditor"
author: "Annie"
date: "2024-06-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Libraries
```{r libraries,}
# Load necessary libraries for various tasks including data manipulation, spatial analysis, plotting, and machine learning
library(broom)
library(car)
library(DALEX)
library(dials)
library(dplyr)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
```

#Set the working directory
```{r Set the working directory}

knitr::opts_knit$set(root.dir =("C:/github/SNR_SDM/Data/processed"))
```

#Read conditor data
Read in the 'conditor.csv' file, print its contents, convert to sf, and set the CRS

```{r read file}
setwd("C:/github/SNR_SDM/Data/processed")
conditor <- read_csv("conditor.csv") %>%
  {print(.); .} %>%  # Print the data frame to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>%
  st_set_crs(4326)  # Set the Coordinate Reference System (CRS) to GDA2020, which has the EPSG code 4326

```

#Create Aust_extent
an extent object for Australia using s min and max X and Y coordinates Object type: Extent (terra package)

```{r}
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)
```

#Land masks
***Run Incorporate paleoview variables code file here***

##current land mask from paleoview
Create a single raster from the first layer of the 'masks' list = current landmask
```{r}
#PV_land_mask_t0 <- masks[[1]] #paleoview land mask
```


##Krapp2021
Create a land mask for the present time using the 'Krapp2021' dataset from the 'pastclim' package Object type: SpatRaster where X =longitude, Y = latitude, Z = time
```{r}
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>% 
  aggregate(fact = 5, fun = "mean", na.rm = T)
```


#plot conditor occurrence
Use ggplots to plot conditor occurrence with current land mask

```{r Plot the data using ggplot2}
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_sf(data = conditor, aes(col = time_bp))  # Add the 'conditor2' sf object as spatial features, colouring by the 'time_bp' column
```

```{r read in palaoview variables}
climate_full <- sds("C:/github/SNR_SDM/Data/processed/paleoview/climate_full.nc")
```

```{r}
conditor2 <- sample_pseudoabs_time(conditor %>% 
                                     filter(time_bp > -20000,
                                            time_bp < -90),
                                   n_per_presence = 3,
                                   raster = climate_full[[1]],
                                   time_col = "time_bp",
                                   lubridate_fun = pastclim::ybp2date,
                                   method = c("dist_min", km2m(150)))

ggplot() +
  geom_spatraster(data = land_mask_current) +
  scale_fill_terrain_c() +
  geom_sf(data = conditor2, aes(col = class))

```
#retrieve climate data for this region

```{r extract climate for region}
conditor_df <- conditor2 %>%
  dplyr::bind_cols(sf::st_coordinates(conditor2)) %>%
  mutate(time_bp = date2ybp(time_step)) %>% 
  as.data.frame() %>%
  select(-geometry)
# get climate
conditor_df <- location_slice_from_region_series(conditor_df,
                                                 region_series = climate_full)
```
add the climate reconstructions to the sf object, and remove the time_step
as we don't need it for modelling

```{r add climate to sf}
conditor4 <- conditor2 %>%
  bind_cols(conditor_df[, c("bio05", "bio06", "bio12")]) %>%
  select(-time_step) %>%
  filter(!is.na(.$bio05))
```

# Fit the model by crossvalidation

 Next, we need to set up a recipe to define how to handle our dataset. We don’t want to transform our data, so we just need to define the formula (class is the outcome, all other variables are predictors; note that, for sf objects, geometry is automatically ignored as a predictor):


```{r make recipe}
conditor_rec <- recipe(conditor4, formula = class ~ .)
conditor_rec
```

We can quickly check that we have the variables that we want with:

`conditor_rec$var_info`

We now build a workflow_set of different models, defining which hyperparameters we want to tune. We will use glm, gam, random forest and boosted trees as our models, so only random forest and boosted trees have tunable hyperparameters. For the most commonly used models, tidysdm automatically chooses the most important parameters, but it is possible to fully customise model specifications.

```{r specify models}
conditor_models <-
  # create the workflow_set
  workflow_set(
    preproc = list(default = conditor_rec),
    models = list(
      # the standard glm specs  (no params to tune)
      glm = sdm_spec_glm(),
      # the standard sdm specs (no params to tune)
      gam = sdm_spec_gam(),
      # rf specs with tuning
      rf = sdm_spec_rf(),
      # boosted tree model (gbm) specs with tuning
      gbm = sdm_spec_boost_tree()
    ),
    # make all combinations of preproc and models,
    cross = TRUE
  ) %>%
  # set formula for gams
  update_workflow_model("default_gam",
                        spec = sdm_spec_gam(),
                        formula = gam_formula(conditor_rec)
  ) %>%
  # tweak controls to store information needed later to create the ensemble
  option_add(control = control_ensemble_grid())
```
Note that gams are unusual, as we need to specify a formula to define to which variables we will fit smooths. By default, gam_formula() fits a smooth to every continuous predictor, but a custom formula can be provided instead.

We now want to set up a spatial block cross-validation scheme to tune and assess our models:

```{r set up cross validation}
library(tidysdm)
set.seed(1005)
conditor_cv <- spatial_block_cv(conditor4, v = 5)
autoplot(conditor_cv)
```

We can now use the block CV folds to tune and assess the models:

```{r tune models}
set.seed(123)
conditor_models <- conditor_models %>%
  workflow_map("tune_grid",
               resamples = conditor_cv, grid = 5,
               metrics = sdm_metric_set(), verbose = TRUE)
```

Note that workflow_set correctly detects that we have no tuning parameters for glm and gam. We can have a look at the performance of our models with:
autoplot(conditor_models)

Now let’s create an ensemble, selecting the best set of parameters for each model (this is really only relevant for the random forest, as there were not hype-parameters to tune for the glm and gam). We will use the Boyce continuous index as our metric to choose the best random forest and boosted tree. When adding members to an ensemble, they are automatically fitted to the full training dataset, and so ready to make predictions.

```{r ensemble}
conditor_ensemble <- simple_ensemble() %>% add_member(conditor_models, metric = "boyce_cont")

autoplot(conditor_ensemble)
```
We can now make predictions with this ensemble (using the default option of taking the mean of the predictions from each model) for the Last Glacial Maximum (LGM, 21,000 years ago).

```{r create LGM}
climate_lgm <- rast(list(climate_full[[1]][[1]], climate_full[[2]][[1]], climate_full[[3]][[1]]))
names(climate_lgm) <- c("bio05", "bio06", "bio12")
```

And predict using the ensemble:

```{r predict lgm}
prediction_lgm <- predict_raster(conditor_ensemble, climate_lgm)
ggplot() +
  geom_spatraster(data = prediction_lgm, aes(fill = mean)) +
  scale_fill_terrain_c(name = "Probability\nof occurrence")
```

Let's make a whole stack of predictions and write the maps to a folder.

```{r stack o maps}
walk(1:1046, function(timestep) {
  year_bp <- flatten_int(time(climate_full))[timestep] - 2000
  climate_snapshot <- rast(list(climate_full[[1]][[timestep]], 
                           climate_full[[2]][[timestep]], 
                           climate_full[[3]][[timestep]]))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")
  prediction1 <- predict_raster(conditor_ensemble, climate_snapshot)
  p <- ggplot() +
    geom_spatraster(data = prediction1, aes(fill = mean)) +
    scale_fill_terrain_c(name = "Probability\nof occurrence") +
    ggtitle(paste0("Paleoview model conditor Years BP: ", year_bp))
  ggsave(path= "C:/github/SNR_SDM/Results/conditorpaleoview",
         plot = p,
         filename = paste0("Paleoview_conditor_map_", year_bp, ".png"))
})
```
#partial dependance profile
explain importance of variables
```{r PDP}
explainer_conditor_ens <- explain_tidysdm(conditor_ensemble)

vip_ensemble <- model_parts(explainer = explainer_conditor_ens)
plot(vip_ensemble)

pdp_bio12 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio12")
plot(pdp_bio12)
```
```{r PDP bio05}
explainer_conditor_ens <- explain_tidysdm(conditor_ensemble)

vip_ensemble <- model_parts(explainer = explainer_conditor_ens)
plot(vip_ensemble)

pdp_bio05 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio05")
plot(pdp_bio05)
```
```{r PDP bio06}
explainer_conditor_ens <- explain_tidysdm(conditor_ensemble)

vip_ensemble <- model_parts(explainer = explainer_conditor_ens)
plot(vip_ensemble)

pdp_bio06 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio06")
plot(pdp_bio06)
```

  



```