---
title: "Krapp apicalis finished"
author: "Annie"
date: "2025-03-14"
output: html_document
---


# Libraries

```{r install_libraries, message=FALSE, warning=FALSE}

knitr::opts_knit$set(root.dir = "C:/github/SNR_SDM/Results")


# List of packages to check, install if needed, and load
packages <- c(
  "broom", "car", "DALEX", "dials", "dplyr", "ggplot2", "infer", 
  "modeldata", "oz", "parsnip", "pastclim", "patchwork", "purrr", 
  "readr", "recipes", "rsample", "rworldmap", "sf", "tidyr", 
  "tidymodels", "tidysdm", "tidyterra", "tibble", "tune", "terra", 
  "workflows", "workflowsets", "gtools", "fs", "USE", "viridis"
)


# Function to display a cat message
cat_message <- function(pkg, message_type) {
  if (message_type == "installed") {
    cat(paste0(pkg, " has been installed successfully!\n"))
  } else if (message_type == "loading") {
    cat(paste0(pkg, " is already installed and has been loaded!\n"))
  }
}

# Install missing packages and load them
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "pastclim") {
      devtools::install_local("C:/github/pastclim")
    } else if (pkg == "USE") {
      devtools::install_github("danddr/USE")
    } else {
        install.packages(pkg)
      }
    
    cat_message(pkg, "installed")
  } else {
    cat_message(pkg, "loading")
  }
  library(pkg, character.only = TRUE)
}

# If you are using renv, you can snapshot the renv after loading all the packages.

# renv::snapshot()
```

```{r libraries}
# Load necessary libraries for data manipulation, spatial analysis, plotting, and modeling
library(broom)
library(car)
library(DALEX)
library(dials)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
library(dplyr)
library(USE)
library(cowplot)

```

# Set the working directory

```{r Set the working directory}

# Set the working directory to the processed data folder
knitr::opts_knit$set(root.dir = "C:/github/SNR_SDM/Data/processed")

# Define the output directory for results
results_dir <- "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

# Define explainer output path (overwritten by future runs)
plot_path <- file.path(results_dir, "explainer.txt")

# Make plot_path available to the explainer script
assign("plot_path", plot_path, envir = .GlobalEnv)

# Run the explainer script
source("C:/github/create_explainer/create_explainer.R")

```

# Preparing your data

## Download Krapp2021 dataset

```{r download climate}
# Download the Krapp2021 dataset using pastclim package
pastclim::download_dataset("Krapp2021")
```

## Read apicalis data

```{r read file}
# Set working directory to the processed data folder
setwd("C:/github/SNR_SDM/Data/processed")

# Read in the 'apicalis_flat.csv' file, print its contents, convert to an sf object, and set the CRS
apicalis <- read_csv("apicalis_flat.csv") %>%
  {print(.); .} %>%  # Print the data to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>%  # Convert to a simple features (sf) object using latitude and longitude
  st_set_crs(4326)  # Set the Coordinate Reference System (CRS) to EPSG:4326 (GDA2020)
```

## Land Mask

### Create an extent for Australia

```{r}
# Define an extent object for Australia using specific longitude (X) and latitude (Y) coordinates
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)


```

### Krapp2021 Land Mask

```{r}
# Create a land mask for the current time using the 'Krapp2021' dataset from the 'pastclim' package
# This mask will restrict our area of interest to Australia
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>%  # Crop to the Australian extent
  aggregate(fact = 5, fun = "mean", na.rm = TRUE)  # Aggregate for better resolution (factor of 5), removing NAs
```

## Plot apicalis Occurrence Data

```{r Plot the data using ggplot2}
# Plot the apicalis occurrence points with the current land mask using ggplot2
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_sf(data = apicalis, aes(col = time_bp))  # Plot apicalis occurrence points, colored by time_bp
```

## Create Climate Time Series Using Krapp2021

```{r create climate time series}
# Create a time series of palaeoclimate reconstructions using the 'Krapp2021' dataset
# Variables used: bio05 (max temp), bio06 (min temp), bio12 (annual precipitation)
climate_vars <- c('bio05','bio06','bio12')
climate_full <- pastclim::region_series(
  bio_variables = climate_vars,  # Specify the variables we are interested in
  data = "Krapp2021",  # Use the Krapp2021 dataset
  crop = vect(Aust_extent)  # Crop the climate data to the extent of Australia
)


```

## Spatial Filtering

```{r spatial filtering}
set.seed(123)
apicalis <- thin_by_dist(apicalis,
  dist_min = km2m(50),

)
nrow(apicalis)
```


## Pseudo-absence Sampling

Stratified sampling in Environmental space

```{r build environmental space}

env_space_1950 <- c(climate_full[[1]][[800]], climate_full[[2]][[800]],
                    climate_full[[3]][[800]])
```



#```{r grid resolution optimisation}
# THIS TAKES AGES BUT ONLY NEEDS TO BE DONE ONCE


   # 1. Extract environmental values from the raster
      env_values <- terra::as.data.frame(env_space_1950, xy = TRUE, na.rm = TRUE)

   # 2. Perform PCA on the environmental variables
   #Check the column names in env_values 
   pca_res <- prcomp(env_values[, c("bio05_0", "bio06_0", "bio12_0")], scale. = TRUE)


    # 3. Create an sf object with the first two PC scores as the coordinates
    env_pca_sf <- st_as_sf(data.frame(PC1 = pca_res$x[, 1],
                                 PC2 = pca_res$x[, 2],
                                  x = env_values$x,
                                  y = env_values$y),
                       coords = c("PC1", "PC2"))

     opt_res <- optimRes(env_pca_sf, grid.res = seq(1, 100, by = 1), perc.thr = 10, cr = 1, showOpt = TRUE)

     optimal_grid_res <- opt_res[[2]]
     print(optimal_grid_res) 
#```


```{r Pseudo absence sampling}
library(scales)  # for alpha()
apicalis3 <- paSampling(
  env.rast = env_space_1950,
  pres = apicalis %>% filter(time_bp == -300),
  thres = 0.75, # you can tune this
  n.tr = 9,
  grid.res = 9 # calculated in optimisation step above, but entered manually 
) |> st_drop_geometry()





# Load world map and convert to sf object
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map)  # Do not filter, keeping all land



pseudoabsence_map <- ggplot() +
  geom_sf(data = australia_outline, 
          fill = alpha("#C4C4C4", 0.5), colour = "grey50") +

  # Pseudo Absences (circle with border)
  geom_point(data = apicalis3, 
             aes(x, y, shape = "Pseudo Absence", fill = "Pseudo Absence"), 
             colour = "black", size = 2.5, stroke = 0.3) +

  # Presences (triangle with border)
  geom_sf(data = apicalis |> filter(time_bp == -300), 
          aes(fill = "Presence", shape = "Presence"), 
          colour = "black", size = 2.5, stroke = 0.3) +

  # Manual shape and colour
  scale_fill_manual(
    values = c("Presence" = "#A100FA", "Pseudo Absence" = "#FF7E1D"),
    name = "Legend"
  ) +
  scale_shape_manual(
    values = c("Presence" = 24, "Pseudo Absence" = 21),
    name = "Legend"
  ) +

  # Set spatial limits and grid breaks
  coord_sf(
    xlim = c(110, 155),
    ylim = c(-45, -10),
    expand = FALSE,
    default_crs = NULL
  ) +
  scale_x_continuous(breaks = seq(110, 155, by = 5)) +
  scale_y_continuous(breaks = seq(-45, -10, by = 5)) +

  # Labels and theme
  labs(
    title = "Pseudo Absences in Environmental Space",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    panel.grid.major = element_line(colour = "#C4C4C4", linewidth = 0.3),
    panel.grid.minor = element_line(colour = "#C4C4C4", linewidth = 0.2),
    panel.border = element_rect(colour = "#C4C4C4", fill = NA, linewidth = 0.3)
  )

# Save the output
ggsave(
  filename = "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/pseudo_absences_envspace.png",
  plot = pseudoabsence_map,
  width = 6,
  height = 5,
  dpi = 300,
  bg = "white"
)



```

## Retrieve Climate Data for This Region

```{r extract climate for region}
# Convert apicalis to a data frame and extract spatial coordinates (longitude, latitude)
apicalis_df <- apicalis %>%
  dplyr::bind_cols(sf::st_coordinates(apicalis)) %>%  # Bind spatial coordinates to the data
  rename(x = X, y = Y) |>
  mutate(percP = "pres") |>
  bind_rows(apicalis3) |>
  mutate(time_bp = -300) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  dplyr::select(-geometry)  # Remove geometry column

# Retrieve climate data for the apicalis occurrence points from the climate series
apicalis_df <- location_slice_from_region_series(apicalis_df,
                                                 region_series = climate_full)  # Use the full climate series for extraction
```

## Add Climate Reconstructions to the sf Object

```{r add climate to sf}
apicalis4 <- apicalis %>%
  dplyr::bind_cols(sf::st_coordinates(apicalis)) %>%  # Bind spatial coordinates to the data
  rename(x = X, y = Y) |>
  mutate(percP = "pres") |>
  bind_rows(apicalis3) |>
  mutate(time_bp = -300) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  dplyr::select(-geometry) |># Remove geometry column
  bind_cols(apicalis_df[, c("bio05", "bio06", "bio12")]) %>%  # Bind the climate data to the sf object
  filter(!is.na(.$bio05)) |>
  mutate(percP = if_else(percP == "pabs", "pseudoabsence", "presence")) |>  # Mutate percP to the correct values
  st_as_sf(coords = c("x", "y")) # Convert back to sf object                               
```

# Fit the model by cross-validation

## Set Up a Recipe for Modeling

```{r make recipe}
# Define a recipe for the model where 'class' is the outcome and other variables are predictors
# Geometry is automatically ignored as a predictor for sf objects
apicalis_rec <- recipe(apicalis4, formula = percP ~ bio05 + bio06 + bio12)  # Define the model formula
apicalis_rec  # Check the recipe
```

## Build a Workflow Set for Model Tuning

```{r specify models}
# Create a workflow set of different models for tuning
# We will use glm, gam, random forest, and boosted trees as our models
apicalis_models <- workflow_set(
  preproc = list(default = apicalis_rec),  # Use the recipe for preprocessing
  models = list(
    glm = sdm_spec_glm(),  # Generalized Linear Model
    gam = sdm_spec_gam(),  # Generalized Additive Model
    rf = sdm_spec_rf(),    # Random Forest with tunable hyperparameters
    gbm = sdm_spec_boost_tree()  # Boosted Tree model with tunable hyperparameters
  ),
  cross = TRUE  # Cross all combinations of preprocessing and models
) %>%
  # Update the workflow model for GAMs with a formula specifying smooth terms for continuous predictors
  update_workflow_model("default_gam",
                        spec = sdm_spec_gam(),
                        formula = percP ~ s(bio05, k = 10) + s(bio06, k = 10) + s(bio12, k = 10)) %>%
  # Adjust controls to store information needed for the ensemble model
  option_add(control = control_ensemble_grid())
```

## Spatial Block Cross-Validation for Model Tuning

```{r set up cross validation}
# Set up spatial block cross-validation with 5 folds for tuning and assessing models
set.seed(1005)  # Set seed for reproducibility
apicalis_cv <- spatial_block_cv(apicalis4, v = 5)  # Create spatial block cross-validation with 5 folds
autoplot(apicalis_cv)  # Plot the cross-validation folds
```

## Tune and Assess Models

```{r tune models}
# Use block cross-validation folds to tune and assess models
set.seed(123)  # Set seed for reproducibility
apicalis_models <- apicalis_models %>%
  workflow_map("tune_grid",  # Tune hyperparameters using grid search
               resamples = apicalis_cv,  # Use the cross-validation resamples
               grid = 5,  # Use a grid size of 5 for tuning
               metrics = sdm_metric_set(),  # Use the default set of SDM metrics
               verbose = TRUE)  # Print verbose output
```

## Create an Ensemble Model

```{r ensemble}
library(ggplot2)
apicalis_ensemble <- simple_ensemble() %>%
  add_member(apicalis_models, metric = "boyce_cont")  # Add members to the ensemble using tss_max as the metric


# Label maps
workflow_labels <- c(
  default_gam = "GAM",
  default_gbm = "Boosted Tree",
  default_glm = "GLM",
  default_rf  = "Random Forest"
)

metric_labels <- c(
  boyce_cont = "Boyce Index",
  roc_auc = "Area Under Curve (AUC)",
  tss_max = "True Skill Statistic (TSS)"
)


model_colours <- c(
default_gam = "#FF61C3",     
      default_gbm = "#00B140",
      default_glm = "#00BFFF",
      default_rf  = "#BC43FF"
)

model_shapes <- c(
  default_gam = 16,
  default_gbm = 17,
  default_glm = 15,
  default_rf  = 3
)

# Final plot
autoplot(apicalis_ensemble) +
  scale_color_manual(
    name = "Model Type",
    values = model_colours,
    labels = workflow_labels
  ) +
  scale_shape_manual(
    name = "Model Type",
    values = model_shapes,
    labels = workflow_labels
  ) +
   labs(
    title = "Ensemble Model Performance Across Algorithms",
    x = "Fold",
    y = "Performance Metric"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "white", colour = NA),
    plot.background = element_rect(fill = "white", colour = NA),
    panel.grid.major = element_line(colour = "grey90"),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "bottom",          # <-- horizontal layout
    legend.direction = "horizontal",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )




```
```{r}
library(purrr)
library(dplyr)
library(ggplot2)
library(tidyr)

# 1. Define your model and metric labels
workflow_labels <- c(
  default_gam = "Generalised Additive Model",
  default_gbm = "Boosted Tree",
  default_glm = "Generalised Linear Model",
  default_rf  = "Random Forest"
)

model_order <- c(
  "Generalised Additive Model",
  "Boosted Tree",
  "Generalised Linear Model",
  "Random Forest"
)

metric_labels <- c(
  boyce_cont = "Boyce Index",
  roc_auc = "AUC",
  tss_max = "TSS"
)

colour_map <- c(
  "Generalised Additive Model" = "#FF61C3",
  "Boosted Tree"               = "#00B140",
  "Generalised Linear Model"   = "#00BFFF",
  "Random Forest"              = "#BC43FF"
)

# 2. Extract fold-level performance from tuned models
perf_data <- apicalis_models %>%
  mutate(metrics = map(result, collect_metrics, summarize = FALSE)) %>%
  select(wflow_id, metrics) %>%
  unnest(metrics)

# 3. Clean and label the data
perf_data_clean <- perf_data %>%
  filter(.metric %in% names(metric_labels)) %>%
  mutate(
    model = factor(workflow_labels[wflow_id], levels = model_order),
    metric_label = factor(metric_labels[.metric], levels = c("Boyce Index", "AUC", "TSS"))
  )

# 4. Plot using boxplot and jitter
ggplot(perf_data_clean, aes(x = model, y = .estimate, fill = model)) +
  geom_boxplot(width = 0.6, alpha = 0.9, outlier.shape = NA) +
  geom_jitter(
    aes(colour = model),
    size = 1.3,
    width = 0.15,
    alpha = 0.4,
    stroke = 0.3,
    shape = 21
  ) +
  scale_fill_manual(values = colour_map, name = "Model Type") +
  scale_colour_manual(values = rep("black", length(model_order)), guide = "none") +
  facet_wrap(~ metric_label, scales = "free_y") +
  labs(
    title = "Model Performance Across Algorithms (Fold-Level)",
    x = NULL,
    y = "Performance Metric"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "#EEEEEE", colour = NA),
    plot.background = element_rect(fill = "white", colour = NA),
    panel.grid.major = element_line(colour = "white"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 13),
    strip.text = element_text(margin = margin(b = 5))
  )


# 6. Save and generate metadata
ggsave("C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/ensemble/ensemble_performance_clean.png",
       width = 8, height = 5, dpi = 300)

plot_path <- "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/ensemble/ensemble_performance_clean.png"


# Add metadata
source("C:/github/create_explainer/create_explainer.R")

```
# Projecting to other times

## Make Predictions for the Last Glacial Maximum (LGM)

```{r create LGM}
# Create a climate raster for the Last Glacial Maximum (LGM) using the correct layer (779) for 21,000 BP
climate_lgm <- rast(list(climate_full[[1]][[538]], climate_full[[2]][[538]], climate_full[[3]][[538]]))  # Extract 20,000 BP layers
names(climate_lgm) <- c("bio05", "bio06", "bio12")  # Rename the layers to match the climate variables


```

## Predict Using the Ensemble Model for LGM

```{r predict lgm}
# Use the ensemble model to predict species occurrence probabilities for the Last Glacial Maximum (LGM)
apicalis_prediction_lgm <- predict_raster(apicalis_ensemble, climate_lgm)  # Predict LGM probabilities

# Plot the predicted probabilities using ggplot2
ggplot() +
  geom_spatraster(data = apicalis_prediction_lgm, aes(fill = mean)) +  # Add the prediction raster to the plot
  scale_fill_terrain_c(name = "Probability\nof occurrence")  # Use a terrain color scale to show probabilities


# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Define the point location (convert to a data frame for ggplot)
point_location <- data.frame(
  longitude = 118.965510,
  latitude = -24.270740  # South is negative in latitude
)

# Plot the predicted probabilities with an Australia outline and the red dot
ggplot() +
  geom_spatraster(data = apicalis_prediction_lgm, aes(fill = mean)) +  # Add the prediction raster
  scale_fill_terrain_c(name = "Probability\nof occurrence") +  # Use a terrain color scale
  geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +  # Add black outline
  geom_point(data = point_location, aes(x = longitude, y = latitude), color = "red", size = 1) +  # Add red dot
  coord_sf(xlim = c(112, 150), ylim = c(-40, -10))  # Crop to 10°S–40°S latitude and 110°E–150°E longitude




```

## save model as spatraster

```{r save_model_spatraster}

# Initialize an empty SpatRaster stack
apicalis_spatrast <- NULL

# Loop through 40 time steps (40ka in 1ka timesteps) to create and stack rasters
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present for each timestep
  
  # Create a climate snapshot for the current timestep
  climate_snapshot <- rast(list(
    climate_full[[1]][[timestep]], 
    climate_full[[2]][[timestep]], 
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename layers to match the climate variables

  # Predict species occurrence for the current climate snapshot
  apicalis_prediction_raster <- predict_raster(apicalis_ensemble, climate_snapshot)
  
  # Stack the prediction raster along the z-axis
  apicalis_spatrast <<- if (is.null(apicalis_spatrast)) {
    apicalis_prediction_raster
  } else {
    c(apicalis_spatrast, apicalis_prediction_raster)
  }
})

# Assign time steps to the z-axis of the stacked raster
time(apicalis_spatrast) <- time(climate_full[[1]])[760:800] - 1950

# Save the stacked SpatRaster to a file
output_file <- "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/predictions_stack.tif"
writeRaster(apicalis_spatrast, filename = output_file, overwrite = TRUE)
```

## stack o maps

```{r stack o maps}
library(magick)

# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Loop over 40 time steps (40 ka in 1 ka timesteps)
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present

  # Create climate snapshot
  climate_snapshot <- rast(list(
    climate_full[[1]][[timestep]], 
    climate_full[[2]][[timestep]], 
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")

  # Predict suitability
  apicalis_prediction1 <- predict_raster(apicalis_ensemble, climate_snapshot)

  # Create plot
  p <- ggplot() +
    geom_spatraster(data = apicalis_prediction1, aes(fill = mean)) +
    scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +
    coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
    ggtitle(bquote(italic("Leporillus apicalis") * " — Years BP: " * .(year_bp))) +
    theme_minimal() +
    theme(
   plot.background = element_rect(fill = "white", colour = NA),     
  panel.background = element_rect(fill = "#F2F2F2", colour = NA),
  panel.grid.major = element_line(colour = "#C4C4C4"),
  panel.grid.minor = element_line(colour = "#C4C4C4"),
      plot.margin = margin(0, 0, 0, 0, unit = "pt")
    )

  # Define output path
  plot_path <- paste0(
    "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/model/apicalis_map_",
    year_bp, ".png"
  )

  # Save PNG
  ggsave(
    filename = plot_path,
    plot = p,
    width = 7,
    height = 7,
    dpi = 300,
    bg = "transparent"
  )
  
  assign("plot_path", plot_path, envir = .GlobalEnv)
source("C:/github/create_explainer/create_explainer.R")

  # Trim and add white border
  img <- image_read(plot_path)
  img_trimmed <- image_trim(img)
  img_bordered <- image_border(img_trimmed, color = "white", geometry = "30x30")
  image_write(img_bordered, path = plot_path)
})




```

```{r 6 image panel}
library(ggplot2)
library(patchwork)
library(terra)
library(purrr)
library(dplyr)

# Extract raster time steps
raster_times <- terra::time(apicalis_spatrast)

# Define target slices
target_times <- c(-40000, -26000, -21000, -12000, -6000, 0)
matched_indices <- match(target_times, raster_times)
panel_tags <- letters[1:length(target_times)]

# Define colours
bg_colour <- "#DCF3FC"
outline_colour <- "black"

# Create each panel (no middens)
panel_plots <- pmap(
  list(idx = matched_indices, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- apicalis_spatrast[[idx]]

    ggplot() +
      geom_spatraster(data = current_prediction, aes(fill = mean)) +
      scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, colour = "grey20", linewidth = 0.3) +
      coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(year_bp))) +
      theme_minimal(base_size = 10) +
      theme(
        panel.background = element_rect(fill = bg_colour, colour = NA),
        plot.background = element_rect(fill = "white", colour = NA),
        panel.grid.major = element_line(colour = "white"),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 8),
        axis.title = element_blank(),
        plot.margin = margin(2, 2, 2, 2),
        legend.position = "right",
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 11)
      ) +
      guides(
        fill = guide_colourbar(
          title.theme = element_text(size = 11),
          label.theme = element_text(size = 11)
        )
      )
  }
)

# Combine into multi-panel figure
final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +
  plot_annotation(
    title = expression(italic("Leporillus apicalis") * " SDM model predictions"),
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )
  )

# Save the panel
ggsave("C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/apicalis_model_panel.png",
       plot = final_panel, width = 10, height = 9, dpi = 300)

```



##Centroid
```{r}
# Calculate suitability-weighted centroid for each timestep in apicalis_spatrast

apicalis_centroids <- lapply(1:nlyr(apicalis_spatrast), function(i) {
  r <- apicalis_spatrast[[i]]

  # Extract suitability values and matching coordinates (including NAs)
  vals <- terra::values(r)[, 1]
  coords <- terra::crds(r, na.rm = FALSE)

  # Filter to cells with non-NA and positive suitability values
  valid <- !is.na(vals) & vals > 0
  if (sum(valid) == 0) return(NULL)  # Skip if no valid cells

  weights <- vals[valid]
  coords <- coords[valid, ]

  # Calculate weighted centroid
  centroid_x <- sum(coords[, 1] * weights) / sum(weights)
  centroid_y <- sum(coords[, 2] * weights) / sum(weights)

  data.frame(
    time_bp = time(apicalis_spatrast)[i],
    longitude = centroid_x,
    latitude = centroid_y
  )
}) %>% bind_rows()


# Rainbow colours: red = recent, violet = old
viridis_colours <- viridis(41, option = "H")



ggplot(apicalis_centroids, aes(x = longitude, y = latitude)) +
  geom_path(colour = "grey50", linewidth = 1) +
  geom_point(aes(colour = time_bp), size = 2) +
 scale_colour_gradientn(
  colours = viridis_colours,
  name = "Years BP",
  limits = c(-40000, 0),
  breaks = c(-40000, -30000, -20000, -10000, 0),
  labels = c("40k", "30k", "20k", "10k", "0")
) +
  coord_fixed(xlim = c(129, 131), ylim = c(-28, -24.8)) +  # Consistent extent
  labs(
    title = "Centroid shift of predicted apicalis habitat through time",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
   theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    panel.grid.major = element_line(colour = "#C4C4C4", linewidth = 0.3),
    panel.grid.minor = element_line(colour = "#C4C4C4", linewidth = 0.2),
    panel.border = element_rect(colour = "#C4C4C4", fill = NA, linewidth = 0.3)
  )


# Save output with clean background
ggsave(
  filename = "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/apicalis_centroid_shift_plot.png",
  plot = last_plot(),
  width = 6, height = 6, dpi = 300,
  bg = "white"
)

# Identify the timestep with the northernmost centroid
northernmost <- apicalis_centroids %>% filter(latitude == max(latitude))
print(northernmost)

```
```{r}
# Load geosphere if not already loaded
if (!requireNamespace("geosphere", quietly = TRUE)) {
  install.packages("geosphere")
}
library(geosphere)
library(dplyr)
library(ggplot2)

# Ensure apicalis_centroids are ordered by time (from oldest to most recent)
apicalis_centroids <- apicalis_centroids %>%
  arrange(desc(time_bp)) %>%
  mutate(
    lead_lon = lead(longitude),
    lead_lat = lead(latitude),
    azimuth = geosphere::bearing(cbind(longitude, latitude), cbind(lead_lon, lead_lat)),
    shift_km = geosphere::distGeo(cbind(longitude, latitude), cbind(lead_lon, lead_lat)) / 1000,
    time_diff = lead(time_bp) - time_bp,
    rate_kmpky = shift_km / (abs(time_diff) / 1000)
  )

# Remove last row with NA values
apicalis_centroids_summary <- apicalis_centroids %>% filter(!is.na(shift_km))

# Stepwise summary stats
total_shift <- sum(apicalis_centroids_summary$shift_km)
mean_rate <- mean(apicalis_centroids_summary$rate_kmpky)

# Compute net azimuth from first to last centroid
start_point <- c(apicalis_centroids$longitude[nrow(apicalis_centroids)], apicalis_centroids$latitude[nrow(apicalis_centroids)])
end_point <- c(apicalis_centroids$longitude[1], apicalis_centroids$latitude[1])
net_azimuth <- geosphere::bearing(start_point, end_point) %% 360

# Helper function: convert azimuth to compass direction
get_compass_direction <- function(azimuth_deg) {
  directions <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")
  breaks <- seq(-22.5, 360, by = 45)
  azimuth_deg <- azimuth_deg %% 360
  direction <- directions[findInterval(azimuth_deg, breaks, rightmost.closed = TRUE)]
  return(direction)
}
net_direction <- get_compass_direction(net_azimuth)

# Print summary
cat("Total shift:", round(total_shift, 1), "km\n")
cat("Average shift per 1,000 years:", round(mean_rate, 2), "km\n")
cat("Net direction of centroid movement:", round(net_azimuth, 1), "degrees (", net_direction, ")\n")

# Optional: plot histogram of stepwise azimuths
ggplot(apicalis_centroids_summary, aes(x = azimuth %% 360)) +
  geom_histogram(binwidth = 15, fill = "#4682B4", colour = "white") +
  coord_polar(start = -pi/2) +
  scale_x_continuous(breaks = seq(0, 360, by = 45)) +
  labs(
    title = "Stepwise azimuths of range shift",
    x = "Direction (degrees from north)",
    y = "Frequency"
  ) +
  theme_minimal()




```
# **Model ends here.**

# Additional steps for validation and analysis

# True skills statistic using midden data

This step is an additional validation step using the midden data set.


## Load and organise midden data set
```{r load midden data set}
#Midden data set is used for additional validation

# Load the independent dataset
dated_middens <- read.csv("C:/github/SNR_SDM/Data/processed/datedmiddens.csv")

# Convert to sf object (assuming it contains longitude and latitude columns)
middens_sf <- dated_middens %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # Adjust column names if different

# Reproject to match the CRS of the apicalis_spatrast
middens_sf <- st_transform(middens_sf, crs = crs(apicalis_spatrast))

# Convert sf object to SpatVector for compatibility with terra::extract
middens_vect <- terra::vect(middens_sf)

# Extract predicted values from the raster stack for presence locations
extracted_values <- terra::extract(apicalis_spatrast, middens_vect)

# Bind the extracted predicted values back to the original sf object
middens_sf$predicted <- extracted_values[, 2]  # Assuming the second column contains predictions

# Extract timesteps from the raster stack
raster_timesteps <- time(apicalis_spatrast)

# Round time_bp to the nearest raster timestep
middens_sf <- middens_sf %>%
  mutate(
    binned_time_bp = sapply(time_bp, function(x) {
      raster_timesteps[which.min(abs(raster_timesteps - x))]
    }),
    lubridate_time = ybp2date(time_bp)
  )

# Create a data frame with raster timesteps
timesteps_df <- data.frame(
  raster_timestep = raster_timesteps  # Timesteps from the raster
)

# Count occurrences per timestep
organized_data <- timesteps_df %>%
  left_join(
    middens_sf %>%
      group_by(binned_time_bp) %>%
      summarise(
        count = n(),                        # Number of occurrences
        mean_value = mean(predicted, na.rm = TRUE)  # Mean predicted suitability (example)
      ),
    by = c("raster_timestep" = "binned_time_bp")
  )

# Replace NAs with 0 for counts (if no data in some bins)
organized_data <- organized_data %>%
  mutate(
    count = replace_na(count, 0),
    mean_value = replace_na(mean_value, NA)  # Leave NA for mean_value where no data exists
  )

# View the organized data
print(organized_data)



```
```{r plot middens}
library(rworldxtra)

# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Extract raster timesteps
raster_timesteps <- time(apicalis_spatrast)

# Loop through each timestep
walk(seq_along(raster_timesteps), function(i) {
  current_year_bp <- raster_timesteps[i]

  # Filter middens within ±500 years of the timestep
  current_middens <- middens_sf %>%
    filter(abs(time_bp - current_year_bp) <= 500)

  # Get prediction layer
  current_prediction <- apicalis_spatrast[[i]]

  # Create plot
  p <- ggplot() +
    geom_spatraster(data = current_prediction, aes(fill = mean)) +
    scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +
    geom_sf(data = current_middens, shape = 21, fill = "#FF3399", colour = "black", size = 2, stroke = 0.4)  +    coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
    ggtitle(bquote(italic("Leporillus apicalis") * " SDM with Middens — Years BP: " * .(current_year_bp))) +
    theme_minimal(base_size = 11) +
theme(
  plot.background = element_rect(fill = "white", colour = NA),     
  panel.background = element_rect(fill = "#F2F2F2", colour = NA),
  panel.grid.major = element_line(colour = "#C4C4C4"),
  panel.grid.minor = element_line(colour = "#C4C4C4"),
  plot.margin = margin(0, 0, 0, 0, "pt")
    )

  # Define save path
  plot_path <- paste0(
    "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/midden maps/apicalis_middens_",
    current_year_bp, ".png"
  )

  # Save plot with white background
  ggsave(
    filename = plot_path,
    plot = p,
    width = 7,
    height = 7,
    dpi = 300,
    bg = "white"
  )
  assign("plot_path", plot_path, envir = .GlobalEnv)
source("C:/github/create_explainer/create_explainer.R")


  # Trim and add white border
  img <- magick::image_read(plot_path)
  img_trimmed <- magick::image_trim(img)
  img_bordered <- image_border(img_trimmed, color = "white", geometry = "30x30")
  magick::image_write(img_bordered, path = plot_path)
})

```





#Make 6 panel plot

```{r make 6 panel plot}

# Extract raster time slices
raster_times <- terra::time(apicalis_spatrast)

# Select target time steps
target_times <- c(-28000, -15000, -12000, -7000, -4000, 0)
matched_indices <- match(target_times, raster_times)
panel_tags <- letters[1:length(target_times)]

# Define colours
point_colour <- "#FF3399"
outline_colour <- "black"

# Create individual panels
panel_plots <- pmap(
  list(idx = matched_indices, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- apicalis_spatrast[[idx]]
    current_middens <- middens_sf %>%
      filter(abs(time_bp - year_bp) <= 500) %>%
      mutate(type = "Midden")

    ggplot() +
      geom_spatraster(data = current_prediction, aes(fill = mean)) +
      scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, colour = outline_colour, linewidth = 0.5) +
      geom_sf(data = current_middens,
              aes(colour = type, shape = type),
              fill = point_colour, size = 2, stroke = 0.4, show.legend = TRUE) +
      scale_colour_manual(name = NULL, values = c("Midden" = "black")) +
      scale_shape_manual(name = NULL, values = c("Midden" = 21)) +
      coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(abs(year_bp)))) +
      theme(
        plot.background = element_rect(fill = "white", colour = NA),
        panel.background = element_rect(fill = "#E8E8E8", colour = NA),
        panel.grid.major = element_line(colour = "#C4C4C4"),
        panel.grid.minor = element_line(colour = "#C4C4C4"),
        axis.text = element_text(size = 8),
        axis.title = element_blank(),
        plot.margin = margin(2, 2, 2, 2),
        legend.position = "bottom",
        legend.box = "horizontal",
        legend.justification = "center",
        legend.spacing.x = unit(1.2, "cm"),
        legend.spacing.y = unit(0.4, "cm"),
        legend.key = element_rect(fill = "transparent", colour = NA),
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 11)
      ) +
      guides(
  fill = guide_colourbar(
    title = "Probability of occurrence",
    title.position = "top",
    title.theme = element_text(size = 11),
    label.theme = element_text(size = 10),
    direction = "horizontal",
    barwidth = unit(5, "cm"),
    barheight = unit(0.4, "cm")
  ),
  colour = guide_legend(
    override.aes = list(size = 2.5),
    title.theme = element_text(size = 11),
    label.theme = element_text(size = 11)
  ),
  shape = guide_legend(
    title.theme = element_text(size = 11),
    label.theme = element_text(size = 11)
  )
)
  }
)

# Combine panels and apply shared legend theme correctly
final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +
  plot_annotation(
    title = "Leporillus apicalis SDM with middens",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )
  ) &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.spacing.x = unit(1.2, "cm"),
    legend.spacing.y = unit(0.4, "cm"),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )

# Save figure
ggsave("C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/midden_panel_plot.png",
       plot = final_panel, width = 10, height = 9, dpi = 300)

```
# Calculate TSS 
The model output is a continuous suitability gradient. for TSS to work we need to define suitability binary. To do this we need to define a threshold - above which is suitable and below is unsuitable. We use MasSS to calculate and define the optimal threshold.

```{r  true skills statistic using midden data MaxSS}

# -----------------------------------------------------------------
# Step 1: Initial Binary Conversion & Pseudo-absence Sampling with Default Threshold
# -----------------------------------------------------------------

# Use a default threshold (0.5) to create a preliminary binary raster.
threshold_init <- 0.27
apicalis_binary_init <- app(x = apicalis_spatrast, fun = \(i) i > threshold_init)
time(apicalis_binary_init) <- time(apicalis_spatrast) |> pastclim::ybp2date()
apicalis_rs_init <- sds(apicalis_binary_init)
names(apicalis_rs_init) <- "pres_abs"  

# Sample pseudo-absences using the default binary raster.
pseudoabs_init <- sample_pseudoabs_time(
  data = dated_middens,
  raster = apicalis_rs_init,
  n_per_presence = 3,         # 3 pseudo-absences per presence
  time_col = "time_bp",
  lubridate_fun = pastclim::ybp2date,
  method = c("dist_min", km2m(200)),
  time_buffer = 365000 * 2
)

# -----------------------------------------------------------------
# Step 2: Extract Continuous Predictions for Presence & Pseudo-Absence Points
# -----------------------------------------------------------------

library(sf)
library(terra)
# Convert presence (midden) data to an sf object.
presence_sf <- st_as_sf(dated_middens, coords = c("longitude", "latitude"), crs = 4326)
# (Reproject if needed to match the raster CRS)
presence_sf <- st_transform(presence_sf, crs(apicalis_spatrast))

# Extract continuous predictions from the first layer (or your chosen layer) of the prediction raster.
presence_vals <- terra::extract(apicalis_spatrast[[1]], presence_sf)[,2]

# Convert pseudoabsence data to an sf object.
pseudoabs_sf <- st_as_sf(pseudoabs_init, coords = c("longitude", "latitude"), crs = 4326)
pseudoabs_sf <- st_transform(pseudoabs_sf, crs(apicalis_spatrast))
pseudoabs_vals <- terra::extract(apicalis_spatrast[[1]], pseudoabs_sf)[,2]

# Combine predicted values and create the observed vector.
predicted_values <- c(presence_vals, pseudoabs_vals)
observed <- c(rep(1, length(presence_vals)), rep(0, length(pseudoabs_vals)))

# -----------------------------------------------------------------
# Step 3: Calculate the Optimal Threshold Using maxSSS
# -----------------------------------------------------------------

thresholds <- seq(0, 1, by = 0.01)
results <- data.frame(threshold = thresholds, TSS = NA)

for(i in seq_along(thresholds)) {
  thresh <- thresholds[i]
  pred_binary <- ifelse(predicted_values > thresh, 1, 0)
  
  # Confusion matrix components:
  a <- sum(pred_binary == 1 & observed == 1)  # True Positives
  b <- sum(pred_binary == 1 & observed == 0)  # False Positives
  c <- sum(pred_binary == 0 & observed == 1)  # False Negatives
  d <- sum(pred_binary == 0 & observed == 0)  # True Negatives
  
  # Skip thresholds where either class is missing.
  if ((a + c) == 0 | (b + d) == 0) next
  
  sensitivity <- a / (a + c)
  specificity <- d / (b + d)
  
  results$TSS[i] <- sensitivity + specificity - 1
}

opt_threshold <- results$threshold[which.max(results$TSS)]
cat("Optimal Threshold (maxSSS):", opt_threshold, "\n")

# -----------------------------------------------------------------
# Step 4: Re-create Binary Raster & Pseudo-Absences Using the Optimal Threshold
# -----------------------------------------------------------------

# Recompute the binary raster using the optimal threshold.
apicalis_binary <- app(x = apicalis_spatrast, fun = \(i) i > opt_threshold)
time(apicalis_binary) <- time(apicalis_spatrast) |> pastclim::ybp2date()
apicalis_rs <- sds(apicalis_binary)
names(apicalis_rs) <- "pres_abs"  

# Now re-sample pseudo-absences using the updated binary raster.
pseudoabs <- sample_pseudoabs_time(
  data = dated_middens,
  raster = apicalis_rs,
  n_per_presence = 3,         # 3 pseudo-absences per presence
  time_col = "time_bp",
  lubridate_fun = pastclim::ybp2date,
  method = c("dist_min", km2m(200)),
  time_buffer = 365000 * 2
)

# Plot the pseudo-absence + presence points with the current land mask.
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +
  scale_fill_terrain_c() +
  geom_point(data = pseudoabs, aes(col = time_step, x = longitude, y = latitude)) +
  geom_point(data = dated_middens, aes(x = longitude, y = latitude), color = "red")

# -----------------------------------------------------------------
# Step 5: Create Data Frame and Evaluate the Model with the Updated Binary Predictions
# -----------------------------------------------------------------

# Create data frame in correct format.
pres_abs_observed <- pseudoabs %>% 
  mutate(time_bp = date2ybp(time_step)) %>%
  select(-class, -time_step) %>% 
  rowid_to_column("name")

# Extract presence and absence points from SDM.
# (Note: location_slice_from_region_series is a custom function on your system.)
pres_abs_pred <- location_slice_from_region_series(
  x = pres_abs_observed,
  region_series = apicalis_rs
)
pres_abs_pred$class <- pseudoabs$class

# Calculate confusion matrix components from the extracted predictions.
a <- nrow(filter(pres_abs_pred, class == "presence", pres_abs == 1))
b <- nrow(filter(pres_abs_pred, class == "pseudoabs", pres_abs == 1))
c <- nrow(filter(pres_abs_pred, class == "presence", pres_abs == 0))
d <- nrow(filter(pres_abs_pred, class == "pseudoabs", pres_abs == 0))

tpr <- a / (a + c)
tnr <- d / (b + d)
tss <- tpr + tnr - 1
sorensen <- 2 * (a / (c + (2 * a) + d))
f2score <- 2 * (((a / (a + b)) * tpr) / ((a / (a + b)) + tpr))

print(paste0("True Skill Statistic: ", tss))
print(paste0("Sorensen: ", sorensen))
print(paste0("F2 score: ", f2score))

```

# Partial Dependence Profiles and Variable Importance

```{r PDP}
library(DALEX)
library(tidysdm)
library(ggplot2)
library(dplyr)

# Create explainer
apicalis_explainer <- explain_tidysdm(apicalis_ensemble)

# Set output path
apicalis_outdir <- "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/variable_importance"

# Get variable importance
apicalis_vip <- model_parts(explainer = apicalis_explainer)

# Filter out baseline and full model
apicalis_vip_filtered <- apicalis_vip %>%
  filter(!variable %in% c("_baseline_", "_full_model_"))

# Summarise to get mean dropout loss for bars
apicalis_vip_summary <- apicalis_vip_filtered %>%
  group_by(variable) %>%
  summarise(mean_dropout_loss = mean(dropout_loss, na.rm = TRUE)) %>%
  ungroup()

# Plot with two layers: bars first, then boxplots
apicalis_vip_plot <- ggplot() +
  # Light purple bars for mean
  geom_col(data = apicalis_vip_summary, 
           aes(x = reorder(variable, -mean_dropout_loss), y = mean_dropout_loss),
           fill = "#BC43FF", colour = "#740CB4", width = 0.7) +
  # Dark purple boxplots fully filled
  geom_boxplot(data = apicalis_vip_filtered,
               aes(x = variable, y = dropout_loss),
               width = 0.4,
               colour = "#740CB4", fill = "#740CB4", alpha = 1,
               outlier.shape = NA) +
  coord_flip() +
  labs(
    title = "Leporillus apicalis – Variable Importance",
    y = "Drop-out Loss",
    x = "Environmental Variable"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    panel.grid.major = element_line(colour = "#C4C4C4", linewidth = 0.3),
    panel.grid.minor = element_line(colour = "#C4C4C4", linewidth = 0.2),
    panel.border = element_rect(colour = "#C4C4C4", fill = NA, linewidth = 0.3),
    plot.title = element_text(face = "italic", size = 13),
    axis.text = element_text(size = 12)  # <- fix axis label size
  ) +
  annotate("text", x = Inf, y = Inf, label = "(b)", hjust = 1.1, vjust = 1.5, size = 5, fontface = "bold")

  


# Save
ggsave(
  filename = file.path(apicalis_outdir, "apicalis_variable_importance_combined.png"),
  plot = apicalis_vip_plot,
  width = 6, height = 4, dpi = 300
)

# 2. FUNCTION TO CREATE & SAVE PDP PLOTS
apicalis_plot_and_save_pdp <- function(var, label) {
  apicalis_pdp <- model_profile(apicalis_explainer, N = 500, variables = var)

  apicalis_pdp_plot <- ggplot(apicalis_pdp$agr_profiles, aes(x = `_x_`, y = `_yhat_`)) +
    geom_line(colour = "darkblue", linewidth = 1) +
    labs(
      title = paste("Leporillus apicalis – Partial Dependence on", var),
      x = label,
      y = "Predicted Suitability"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "italic", size = 13),
      legend.position = "none"
    )

  ggsave(
    filename = file.path(apicalis_outdir, paste0("apicalis_pdp_", var, ".png")),
    plot = apicalis_pdp_plot,
    width = 6, height = 4, dpi = 300
  )
}

# 3. GENERATE & SAVE ALL PDPs
apicalis_plot_and_save_pdp("bio05", "Maximum temperature of warmest month (°C)")
apicalis_plot_and_save_pdp("bio06", "Minimum temperature of coldest month (°C)")
apicalis_plot_and_save_pdp("bio12", "Annual precipitation (mm)")


```

##Kernel density estimate 
```{r}
# Load required libraries
# Load necessary libraries
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)

# Load and prepare data
dated_middens_2 <- read.csv("C:/github/SNR_SDM/Data/processed/datedmiddens.csv")

# Convert to sf
middens_sf_2 <- st_as_sf(dated_middens_2, coords = c("longitude", "latitude"), crs = 4326)

# Extract coordinates
middens_df_2 <- middens_sf_2 %>%
  mutate(
    longitude = st_coordinates(.)[, 1],
    latitude = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry()

# Australia map
aus_outline <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "Australia")

# Dummy bbox for consistent framing
bbox_sf <- st_as_sfc(st_bbox(c(
  xmin = 108, xmax = 152,
  ymin = -46, ymax = -8
), crs = 4326)) %>% st_sf()

# Check if there are at least a few dozen distinct coordinates
print(head(middens_df_2))
print(nrow(middens_df_2))

# Add dummy points at map edges to force KDE range
pad_points <- data.frame(
  longitude = c(110, 150, 150, 110),
  latitude = c(-10, -10, -45, -45)
)

# Combine with midden data
middens_df_padded <- bind_rows(middens_df_2, pad_points)

# Plot with KDE using padded data
p <- ggplot() +
  geom_sf(data = aus_outline, fill = scales::alpha("#C4C4C4", 0.5), colour = "black") +
  stat_density_2d(
    data = middens_df_padded,
    aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    colour = NA,
    adjust = 0.4,
    alpha = 0.8,
    n = 200
  ) +
  scale_fill_viridis_c(option = "magma", name = "Density", direction = -1) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +
  labs(
    title = "Spatial Density of Dated Midden Samples",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    panel.grid.major = element_line(colour = "#C4C4C4"),
    panel.grid.minor = element_line(colour = "#C4C4C4"),
    
    # Legend styling for in-panel placement
    legend.position = c(0.03, 0.20),  # ⬅️ top-left corner
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", colour = "grey70", linewidth = 0.3),
    legend.key = element_rect(fill = "white", colour = NA),
    legend.margin = margin(6, 12, 6, 9),  # ⬅️ wider horizontal padding
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
      ) +
    guides(
      fill = guide_colourbar(
        title = "Density",
        title.position = "top",
        title.theme = element_text(size = 10),
        label.theme = element_text(size = 9),
        direction = "horizontal",
        barwidth = unit(4, "cm"),
        barheight = unit(0.4, "cm"),
        ticks = TRUE,
        ticks.colour = "black",
        frame.colour = "grey70",
        frame.linewidth = 0.3,
        label.position = "bottom",
        breaks = c(0, 0.04, 0.08)  # ⬅️ This limits to 3 ticks
      )
    )


ggsave(
  filename = "C:/github/SNR_SDM/Results/stratified enviromental sampling/midden_sampling_density.png",
  plot = p,
  width = 7,
  height = 6,
  dpi = 300,
  bg = "white"
)




```
```{r endnote library}
# Ensure working directory is set to Results folder
setwd("C:/github/SNR_SDM/Results")

# Run EndNote export script to create citation library in this folder
source("C:/github/create_endnote_library/create_endnote_library.r")

```

```{r}
lapply(c("tidysdm", "tidymodels", "workflowsets", "workflows", "tune", "parsnip", "recipes", "yardstick", "rsample", 
         "spatialsample", "dials", "dplyr", "tidyr", "purrr", "tibble", "ggplot2", "cowplot", "scales", "viridis", 
         "viridisLite", "sf", "terra", "tidyterra", "pastclim", "oz", "rworldmap", "DALEX", "ecospat", "broom", 
         "infer", "readr", "fs", "gtools", "RefManageR"), 
       function(pkg) { cat("\n---", pkg, "---\n"); print(citation(pkg)) })

```