---
title: "Krapp apicalis finished"
author: "Annie"
date: "2025-03-14"
output: html_document
---


# Libraries

```{r install_libraries, message=FALSE, warning=FALSE}

# List of packages to check, install if needed, and load
packages <- c(
  "broom", "car", "DALEX", "dials", "dplyr", "ggplot2", "infer", 
  "modeldata", "oz", "parsnip", "pastclim", "purrr", "readr", 
  "recipes", "rsample", "rworldmap", "sf", "tidyr", "tidymodels", 
  "tidysdm", "tidyterra", "tibble", "tune", "terra", "workflows", 
  "workflowsets", "gtools", "fs", "USE"
)

# Function to display a cat message
cat_message <- function(pkg, message_type) {
  if (message_type == "installed") {
    cat(paste0(pkg, " has been installed successfully!\n"))
  } else if (message_type == "loading") {
    cat(paste0(pkg, " is already installed and has been loaded!\n"))
  }
}

# Install missing packages and load them
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "pastclim") {
      devtools::install_local("C:/github/pastclim")
    } else if (pkg == "USE") {
      devtools::install_github("danddr/USE")
    } else {
        install.packages(pkg)
      }
    
    cat_message(pkg, "installed")
  } else {
    cat_message(pkg, "loading")
  }
  library(pkg, character.only = TRUE)
}

# If you are using renv, you can snapshot the renv after loading all the packages.

# renv::snapshot()
```

```{r libraries}
# Load necessary libraries for data manipulation, spatial analysis, plotting, and modeling
library(broom)
library(car)
library(DALEX)
library(dials)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
library(dplyr)
library(USE)
```

# Set the working directory

```{r Set the working directory}
# Set the working directory to the processed data folder
knitr::opts_knit$set(root.dir =("C:/github/SNR_SDM/Data/processed"))
```

# Preparing your data

## Download Krapp2021 dataset

```{r download climate}
# Download the Krapp2021 dataset using pastclim package
pastclim::download_dataset("Krapp2021")
```

## Read apicalis data

```{r read file}
# Set working directory to the processed data folder
setwd("C:/github/SNR_SDM/Data/processed")

# Read in the 'apicalis.csv' file, print its contents, convert to an sf object, and set the CRS
apicalis <- read_csv("apicalis.csv") %>%
  {print(.); .} %>%  # Print the data to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>%  # Convert to a simple features (sf) object using latitude and longitude
  st_set_crs(4326)  # Set the Coordinate Reference System (CRS) to EPSG:4326 (GDA2020)
```

## Land Mask

### Create an extent for Australia

```{r}
# Define an extent object for Australia using specific longitude (X) and latitude (Y) coordinates
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)
```

### Krapp2021 Land Mask

```{r}
# Create a land mask for the current time using the 'Krapp2021' dataset from the 'pastclim' package
# This mask will restrict our area of interest to Australia
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>%  # Crop to the Australian extent
  aggregate(fact = 5, fun = "mean", na.rm = TRUE)  # Aggregate for better resolution (factor of 5), removing NAs
```

## Plot apicalis Occurrence Data

```{r Plot the data using ggplot2}
# Plot the apicalis occurrence points with the current land mask using ggplot2
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_sf(data = apicalis, aes(col = time_bp))  # Plot apicalis occurrence points, colored by time_bp
```

## Create Climate Time Series Using Krapp2021

```{r create climate time series}
# Create a time series of palaeoclimate reconstructions using the 'Krapp2021' dataset
# Variables used: bio05 (max temp), bio06 (min temp), bio12 (annual precipitation)
climate_vars <- c('bio05','bio06','bio12')
climate_full <- pastclim::region_series(
  bio_variables = climate_vars,  # Specify the variables we are interested in
  data = "Krapp2021",  # Use the Krapp2021 dataset
  crop = vect(Aust_extent)  # Crop the climate data to the extent of Australia
)


```

## Spatial Filtering

```{r spatial filtering}
set.seed(123)
apicalis <- thin_by_dist(apicalis,
  dist_min = km2m(50),

)
nrow(apicalis)
```


## Pseudo-absence Sampling

Stratified sampling in Environmental space

```{r build environmental space}

env_space_1950 <- c(climate_full[[1]][[800]], climate_full[[2]][[800]],
                    climate_full[[3]][[800]])
```



#```{r grid resolution optimisation}
# THIS TAKES AGES BUT ONLY NEEDS TO BE DONE ONCE


   # 1. Extract environmental values from the raster
      env_values <- terra::as.data.frame(env_space_1950, xy = TRUE, na.rm = TRUE)

   # 2. Perform PCA on the environmental variables
   #Check the column names in env_values 
   pca_res <- prcomp(env_values[, c("bio05_0", "bio06_0", "bio12_0")], scale. = TRUE)


    # 3. Create an sf object with the first two PC scores as the coordinates
    env_pca_sf <- st_as_sf(data.frame(PC1 = pca_res$x[, 1],
                                 PC2 = pca_res$x[, 2],
                                  x = env_values$x,
                                  y = env_values$y),
                       coords = c("PC1", "PC2"))

     opt_res <- optimRes(env_pca_sf, grid.res = seq(1, 100, by = 1), perc.thr = 10, cr = 1, showOpt = TRUE)

     optimal_grid_res <- opt_res[[2]]
     print(optimal_grid_res) 
#```


```{r Pseudo absence sampling}

apicalis3 <- paSampling(
  env.rast = env_space_1950,
  pres = apicalis %>% filter(time_bp == -300),
  thres = 0.75, # you can tune this
  n.tr = 9,
  grid.res = 9 # calculated in optimisation step above, but entered manually 
) |> st_drop_geometry()

# Load world map and convert to sf object
world_map <- rworldmap::getMap(resolution = "high")
land_outline <- sf::st_as_sf(world_map)  # Do not filter, keeping all land



# Define the plot
ggplot() +
  geom_sf(data = australia_outline, fill = "#F9E3D9", color = "#F1CFCF") +  # Use proper landmass outline
  
  # Pseudo-absences (circles) with thin black outline
  geom_point(data = apicalis3, aes(x, y, color = "Pseudo Absence", shape = "Pseudo Absence"), 
             size = 2, stroke = 0) +  
  
  # Presences (triangles) with thin black outline
  geom_sf(data = apicalis |> filter(time_bp == -300), aes(color = "Presence", shape = "Presence"), 
          size = 2, stroke = 0) +  
  
  scale_color_manual(values = c("Presence" = "#D689FF", "Pseudo Absence" = "#00D19B")) +  # Define colors
  scale_shape_manual(values = c("Presence" = 17, "Pseudo Absence" = 16)) +  # Triangles (17) for Presence, Circles (16) for Pseudo Absence
  
  labs(
    title = "Pseudo Absences in Environmental Space",
    x = "Longitude",
    y = "Latitude",
    color = "Legend",
    shape = "Legend"  # Ensure legend applies to both color & shape
  ) +
  coord_sf(xlim = c(110, 150), ylim = c(-40, -10)) +  # Crop correctly
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#DCF3FC", color = NA),  # Background color for map
    plot.background = element_rect(fill = "white", color = NA),  # Keep plot background white
    panel.grid.major = element_line(color = "white"),  # White major gridlines
    panel.grid.minor = element_line(color = "white")   # White minor gridlines
  )




```

## Retrieve Climate Data for This Region

```{r extract climate for region}
# Convert apicalis to a data frame and extract spatial coordinates (longitude, latitude)
apicalis_df <- apicalis %>%
  dplyr::bind_cols(sf::st_coordinates(apicalis)) %>%  # Bind spatial coordinates to the data
  rename(x = X, y = Y) |>
  mutate(percP = "pres") |>
  bind_rows(apicalis3) |>
  mutate(time_bp = -300) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  dplyr::select(-geometry)  # Remove geometry column

# Retrieve climate data for the apicalis occurrence points from the climate series
apicalis_df <- location_slice_from_region_series(apicalis_df,
                                                 region_series = climate_full)  # Use the full climate series for extraction
```

## Add Climate Reconstructions to the sf Object

```{r add climate to sf}
apicalis4 <- apicalis %>%
  dplyr::bind_cols(sf::st_coordinates(apicalis)) %>%  # Bind spatial coordinates to the data
  rename(x = X, y = Y) |>
  mutate(percP = "pres") |>
  bind_rows(apicalis3) |>
  mutate(time_bp = -300) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  dplyr::select(-geometry) |># Remove geometry column
  bind_cols(apicalis_df[, c("bio05", "bio06", "bio12")]) %>%  # Bind the climate data to the sf object
  filter(!is.na(.$bio05)) |>
  mutate(percP = if_else(percP == "pabs", "pseudoabsence", "presence")) |>  # Mutate percP to the correct values
  st_as_sf(coords = c("x", "y")) # Convert back to sf object                               
```

# Fit the model by cross-validation

## Set Up a Recipe for Modeling

```{r make recipe}
# Define a recipe for the model where 'class' is the outcome and other variables are predictors
# Geometry is automatically ignored as a predictor for sf objects
apicalis_rec <- recipe(apicalis4, formula = percP ~ bio05 + bio06 + bio12)  # Define the model formula
apicalis_rec  # Check the recipe
```

## Build a Workflow Set for Model Tuning

```{r specify models}
# Create a workflow set of different models for tuning
# We will use glm, gam, random forest, and boosted trees as our models
apicalis_models <- workflow_set(
  preproc = list(default = apicalis_rec),  # Use the recipe for preprocessing
  models = list(
    glm = sdm_spec_glm(),  # Generalized Linear Model
    gam = sdm_spec_gam(),  # Generalized Additive Model
    rf = sdm_spec_rf(),    # Random Forest with tunable hyperparameters
    gbm = sdm_spec_boost_tree()  # Boosted Tree model with tunable hyperparameters
  ),
  cross = TRUE  # Cross all combinations of preprocessing and models
) %>%
  # Update the workflow model for GAMs with a formula specifying smooth terms for continuous predictors
  update_workflow_model("default_gam",
                        spec = sdm_spec_gam(),
                        formula = percP ~ s(bio05, k = 10) + s(bio06, k = 10) + s(bio12, k = 10)) %>%
  # Adjust controls to store information needed for the ensemble model
  option_add(control = control_ensemble_grid())
```

## Spatial Block Cross-Validation for Model Tuning

```{r set up cross validation}
# Set up spatial block cross-validation with 5 folds for tuning and assessing models
set.seed(1005)  # Set seed for reproducibility
apicalis_cv <- spatial_block_cv(apicalis4, v = 5)  # Create spatial block cross-validation with 5 folds
autoplot(apicalis_cv)  # Plot the cross-validation folds
```

## Tune and Assess Models

```{r tune models}
# Use block cross-validation folds to tune and assess models
set.seed(123)  # Set seed for reproducibility
apicalis_models <- apicalis_models %>%
  workflow_map("tune_grid",  # Tune hyperparameters using grid search
               resamples = apicalis_cv,  # Use the cross-validation resamples
               grid = 5,  # Use a grid size of 5 for tuning
               metrics = sdm_metric_set(),  # Use the default set of SDM metrics
               verbose = TRUE)  # Print verbose output
```

## Create an Ensemble Model

```{r ensemble}
# Create an ensemble by selecting the best set of parameters for each model
# tss_max will be used as the metric for selecting the best random forest and boosted tree models
apicalis_ensemble <- simple_ensemble() %>%
  add_member(apicalis_models, metric = "boyce_cont")  # Add members to the ensemble using tss_max as the metric

autoplot(apicalis_ensemble)  # Plot the ensemble model performance
```

# Projecting to other times

## Make Predictions for the Last Glacial Maximum (LGM)

```{r create LGM}
# Create a climate raster for the Last Glacial Maximum (LGM) using the correct layer (779) for 21,000 BP
climate_lgm <- rast(list(climate_full[[1]][[538]], climate_full[[2]][[538]], climate_full[[3]][[538]]))  # Extract 20,000 BP layers
names(climate_lgm) <- c("bio05", "bio06", "bio12")  # Rename the layers to match the climate variables


```

## Predict Using the Ensemble Model for LGM

```{r predict lgm}
# Use the ensemble model to predict species occurrence probabilities for the Last Glacial Maximum (LGM)
prediction_lgm <- predict_raster(apicalis_ensemble, climate_lgm)  # Predict LGM probabilities

# Plot the predicted probabilities using ggplot2
ggplot() +
  geom_spatraster(data = prediction_lgm, aes(fill = mean)) +  # Add the prediction raster to the plot
  scale_fill_terrain_c(name = "Probability\nof occurrence")  # Use a terrain color scale to show probabilities


# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Define the point location (convert to a data frame for ggplot)
point_location <- data.frame(
  longitude = 118.965510,
  latitude = -24.270740  # South is negative in latitude
)

# Plot the predicted probabilities with an Australia outline and the red dot
ggplot() +
  geom_spatraster(data = prediction_lgm, aes(fill = mean)) +  # Add the prediction raster
  scale_fill_terrain_c(name = "Probability\nof occurrence") +  # Use a terrain color scale
  geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +  # Add black outline
  geom_point(data = point_location, aes(x = longitude, y = latitude), color = "red", size = 1) +  # Add red dot
  coord_sf(xlim = c(112, 150), ylim = c(-40, -10))  # Crop to 10°S–40°S latitude and 110°E–150°E longitude




```

## save model as spatraster

```{r save_model_spatraster}

# Initialize an empty SpatRaster stack
apicalis_spatrast <- NULL

# Loop through 40 time steps (40ka in 1ka timesteps) to create and stack rasters
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present for each timestep
  
  # Create a climate snapshot for the current timestep
  climate_snapshot <- rast(list(
    climate_full[[1]][[timestep]], 
    climate_full[[2]][[timestep]], 
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename layers to match the climate variables

  # Predict species occurrence for the current climate snapshot
  prediction_raster <- predict_raster(apicalis_ensemble, climate_snapshot)
  
  # Stack the prediction raster along the z-axis
  apicalis_spatrast <<- if (is.null(apicalis_spatrast)) {
    prediction_raster
  } else {
    c(apicalis_spatrast, prediction_raster)
  }
})

# Assign time steps to the z-axis of the stacked raster
time(apicalis_spatrast) <- time(climate_full[[1]])[760:800] - 1950

# Save the stacked SpatRaster to a file
output_file <- "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/predictions_stack.tif"
writeRaster(apicalis_spatrast, filename = output_file, overwrite = TRUE)
```

## Create a Stack of Predictions and Plot for All Time Steps

```{r stack o maps}
# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Loop over 40 time steps (40ka in 1ka timesteps) to create and save a map of predictions for each time step
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present for each timestep
  
  # Create a climate snapshot for the current timestep
  climate_snapshot <- rast(list(climate_full[[1]][[timestep]], 
                                climate_full[[2]][[timestep]], 
                                climate_full[[3]][[timestep]]))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename layers to match the climate variables

  # Predict species occurrence for the current climate snapshot
  prediction1 <- predict_raster(apicalis_ensemble, climate_snapshot)

  # Plot the predictions using ggplot2
  p <- ggplot() +
    geom_spatraster(data = prediction1, aes(fill = mean)) +  # Add the prediction raster
    scale_fill_terrain_c(name = "Probability\nof occurrence") +  # Use a terrain color scale
    geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +  # Add black Australia outline
    coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +  # Crop to 10°S–40°S latitude, 110°E–150°E longitude
    ggtitle(bquote("Predicted distribution of " * italic("Leporillus apicalis") * " (Krapp2021)  Years BP: " * .(year_bp)))

  # Save the plot to a folder with filenames indicating the year BP
  ggsave(filename = paste0("C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/model/apicalis_map_",
                           year_bp, ".png"),
         plot = p,
         width = 7, height = 7)
})

```

# **Model ends here.**

# Additional steps for validation and analysis

# True skills statistic using midden data

This step is an additional validation step using the midden data set.


## Load and organise midden data set
```{r load midden data set}
#Midden data set is used for additional validation

# Load the independent dataset
dated_middens <- read.csv("C:/github/SNR_SDM/Data/processed/datedmiddens.csv")

# Convert to sf object (assuming it contains longitude and latitude columns)
middens_sf <- dated_middens %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # Adjust column names if different

# Reproject to match the CRS of the apicalis_spatrast
middens_sf <- st_transform(middens_sf, crs = crs(apicalis_spatrast))

# Convert sf object to SpatVector for compatibility with terra::extract
middens_vect <- terra::vect(middens_sf)

# Extract predicted values from the raster stack for presence locations
extracted_values <- terra::extract(apicalis_spatrast, middens_vect)

# Bind the extracted predicted values back to the original sf object
middens_sf$predicted <- extracted_values[, 2]  # Assuming the second column contains predictions

# Extract timesteps from the raster stack
raster_timesteps <- time(apicalis_spatrast)

# Round time_bp to the nearest raster timestep
middens_sf <- middens_sf %>%
  mutate(
    binned_time_bp = sapply(time_bp, function(x) {
      raster_timesteps[which.min(abs(raster_timesteps - x))]
    }),
    lubridate_time = ybp2date(time_bp)
  )

# Create a data frame with raster timesteps
timesteps_df <- data.frame(
  raster_timestep = raster_timesteps  # Timesteps from the raster
)

# Count occurrences per timestep
organized_data <- timesteps_df %>%
  left_join(
    middens_sf %>%
      group_by(binned_time_bp) %>%
      summarise(
        count = n(),                        # Number of occurrences
        mean_value = mean(predicted, na.rm = TRUE)  # Mean predicted suitability (example)
      ),
    by = c("raster_timestep" = "binned_time_bp")
  )

# Replace NAs with 0 for counts (if no data in some bins)
organized_data <- organized_data %>%
  mutate(
    count = replace_na(count, 0),
    mean_value = replace_na(mean_value, NA)  # Leave NA for mean_value where no data exists
  )

# View the organized data
print(organized_data)



```
```{r plot middens}
install.packages("rworldxtra")
library(rworldxtra)
# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Define the point location (convert to a data frame for ggplot)
point_location <- data.frame(
  longitude = 118.965510,
  latitude = -24.270740  # South is negative in latitude
)

# Extract raster timesteps
raster_timesteps <- time(apicalis_spatrast)

# Loop through each timestep in your stacked raster predictions
walk(seq_along(raster_timesteps), function(i) {
  
  current_year_bp <- raster_timesteps[i]

  # Filter middens relevant to the current raster timestep (rounded to nearest 1000 years)
  current_middens <- middens_sf %>%
    filter(abs(time_bp - current_year_bp) <= 500) # within ±500 years of raster timestep

  # Get the raster layer for the current timestep
  current_prediction <- apicalis_spatrast[[i]]

  # Plot SDM prediction with midden points overlayed and purple reference dot
  p <- ggplot() +
    geom_spatraster(data = current_prediction, aes(fill = mean)) +  # Add the prediction raster
    scale_fill_terrain_c(name = "Probability\nof occurrence") +  # Use a terrain color scale
    geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +  # Add black Australia outline
    geom_sf(data = current_middens, colour = "#FF3399", size = 2, shape = 19) +  # Midden locations (pink)
    geom_point(data = point_location, aes(x = longitude, y = latitude), color = "purple", size = 1) +  # Purple reference dot
    coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +  # Crop to 10°S–40°S latitude, 110°E–150°E longitude
    ggtitle(bquote(italic("Leporillus apicalis") * " SDM with Middens — Years BP: " * .(current_year_bp)))

  # Save the plot to a folder with filenames indicating the year BP
  ggsave(
    filename = paste0(
      "C:/github/SNR_SDM/Results/stratified enviromental sampling/apicalis/midden maps/apicalis_middens_",
      current_year_bp, ".png"
    ),
    plot = p,
    width = 7,
    height = 7
  )
})


```
# Calculate TSS 
The model output is a continuous suitability gradient. for TSS to work we need to define suitability binary. To do this we need to define a threshold - above which is suitable and below is unsuitable. We use MasSS to calculate and define the optimal threshold.

```{r  true skills statistic using midden data MaxSS}

# -----------------------------------------------------------------
# Step 1: Initial Binary Conversion & Pseudo-absence Sampling with Default Threshold
# -----------------------------------------------------------------

# Use a default threshold (0.5) to create a preliminary binary raster.
threshold_init <- 0.35
apicalis_binary_init <- app(x = apicalis_spatrast, fun = \(i) i > threshold_init)
time(apicalis_binary_init) <- time(apicalis_spatrast) |> pastclim::ybp2date()
apicalis_rs_init <- sds(apicalis_binary_init)
names(apicalis_rs_init) <- "pres_abs"  

# Sample pseudo-absences using the default binary raster.
pseudoabs_init <- sample_pseudoabs_time(
  data = dated_middens,
  raster = apicalis_rs_init,
  n_per_presence = 3,         # 3 pseudo-absences per presence
  time_col = "time_bp",
  lubridate_fun = pastclim::ybp2date,
  method = c("dist_min", km2m(200)),
  time_buffer = 365000 * 2
)

# -----------------------------------------------------------------
# Step 2: Extract Continuous Predictions for Presence & Pseudo-Absence Points
# -----------------------------------------------------------------

library(sf)
library(terra)
# Convert presence (midden) data to an sf object.
presence_sf <- st_as_sf(dated_middens, coords = c("longitude", "latitude"), crs = 4326)
# (Reproject if needed to match the raster CRS)
presence_sf <- st_transform(presence_sf, crs(apicalis_spatrast))

# Extract continuous predictions from the first layer (or your chosen layer) of the prediction raster.
presence_vals <- terra::extract(apicalis_spatrast[[1]], presence_sf)[,2]

# Convert pseudoabsence data to an sf object.
pseudoabs_sf <- st_as_sf(pseudoabs_init, coords = c("longitude", "latitude"), crs = 4326)
pseudoabs_sf <- st_transform(pseudoabs_sf, crs(apicalis_spatrast))
pseudoabs_vals <- terra::extract(apicalis_spatrast[[1]], pseudoabs_sf)[,2]

# Combine predicted values and create the observed vector.
predicted_values <- c(presence_vals, pseudoabs_vals)
observed <- c(rep(1, length(presence_vals)), rep(0, length(pseudoabs_vals)))

# -----------------------------------------------------------------
# Step 3: Calculate the Optimal Threshold Using maxSSS
# -----------------------------------------------------------------

thresholds <- seq(0, 1, by = 0.01)
results <- data.frame(threshold = thresholds, TSS = NA)

for(i in seq_along(thresholds)) {
  thresh <- thresholds[i]
  pred_binary <- ifelse(predicted_values > thresh, 1, 0)
  
  # Confusion matrix components:
  a <- sum(pred_binary == 1 & observed == 1)  # True Positives
  b <- sum(pred_binary == 1 & observed == 0)  # False Positives
  c <- sum(pred_binary == 0 & observed == 1)  # False Negatives
  d <- sum(pred_binary == 0 & observed == 0)  # True Negatives
  
  # Skip thresholds where either class is missing.
  if ((a + c) == 0 | (b + d) == 0) next
  
  sensitivity <- a / (a + c)
  specificity <- d / (b + d)
  
  results$TSS[i] <- sensitivity + specificity - 1
}

opt_threshold <- results$threshold[which.max(results$TSS)]
cat("Optimal Threshold (maxSSS):", opt_threshold, "\n")

# -----------------------------------------------------------------
# Step 4: Re-create Binary Raster & Pseudo-Absences Using the Optimal Threshold
# -----------------------------------------------------------------

# Recompute the binary raster using the optimal threshold.
apicalis_binary <- app(x = apicalis_spatrast, fun = \(i) i > opt_threshold)
time(apicalis_binary) <- time(apicalis_spatrast) |> pastclim::ybp2date()
apicalis_rs <- sds(apicalis_binary)
names(apicalis_rs) <- "pres_abs"  

# Now re-sample pseudo-absences using the updated binary raster.
pseudoabs <- sample_pseudoabs_time(
  data = dated_middens,
  raster = apicalis_rs,
  n_per_presence = 3,         # 3 pseudo-absences per presence
  time_col = "time_bp",
  lubridate_fun = pastclim::ybp2date,
  method = c("dist_min", km2m(200)),
  time_buffer = 365000 * 2
)

# Plot the pseudo-absence + presence points with the current land mask.
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +
  scale_fill_terrain_c() +
  geom_point(data = pseudoabs, aes(col = time_step, x = longitude, y = latitude)) +
  geom_point(data = dated_middens, aes(x = longitude, y = latitude), color = "red")

# -----------------------------------------------------------------
# Step 5: Create Data Frame and Evaluate the Model with the Updated Binary Predictions
# -----------------------------------------------------------------

# Create data frame in correct format.
pres_abs_observed <- pseudoabs %>% 
  mutate(time_bp = date2ybp(time_step)) %>%
  select(-class, -time_step) %>% 
  rowid_to_column("name")

# Extract presence and absence points from SDM.
# (Note: location_slice_from_region_series is a custom function on your system.)
pres_abs_pred <- location_slice_from_region_series(
  x = pres_abs_observed,
  region_series = apicalis_rs
)
pres_abs_pred$class <- pseudoabs$class

# Calculate confusion matrix components from the extracted predictions.
a <- nrow(filter(pres_abs_pred, class == "presence", pres_abs == 1))
b <- nrow(filter(pres_abs_pred, class == "pseudoabs", pres_abs == 1))
c <- nrow(filter(pres_abs_pred, class == "presence", pres_abs == 0))
d <- nrow(filter(pres_abs_pred, class == "pseudoabs", pres_abs == 0))

tpr <- a / (a + c)
tnr <- d / (b + d)
tss <- tpr + tnr - 1
sorensen <- 2 * (a / (c + (2 * a) + d))
f2score <- 2 * (((a / (a + b)) * tpr) / ((a / (a + b)) + tpr))

print(paste0("True Skill Statistic: ", tss))
print(paste0("Sorensen: ", sorensen))
print(paste0("F2 score: ", f2score))

```

# Partial Dependence Profiles and Variable Importance

```{r PDP}
# PDP for bio12
# Explain the ensemble model to generate partial dependence profiles and variable importance
explainer_apicalis_ens <- explain_tidysdm(apicalis_ensemble)  # Create an explainer object for the ensemble model

# Compute variable importance using the DALEX package
vip_ensemble <- model_parts(explainer = explainer_apicalis_ens)  # Calculate variable importance
plot(vip_ensemble)  # Plot the variable importance results

# Generate a partial dependence profile for bio12
pdp_bio12 <- model_profile(explainer_apicalis_ens, N = 500, variables = "bio12")  # Generate PDP for bio12
plot(pdp_bio12)  # Plot the partial dependence profile for bio12


#PDP for bio05

# Generate and plot a partial dependence profile for bio05
pdp_bio05 <- model_profile(explainer_apicalis_ens, N = 500, variables = "bio05")
plot(pdp_bio05)


#$ PDP for bio06
 
# Generate and plot a partial dependence profile for bio06
pdp_bio06 <- model_profile(explainer_apicalis_ens, N = 500, variables = "bio06")
plot(pdp_bio06)
```
