---
title: "Krapp apicalis finished"
author: "Annie"
date: "2025-03-14"
output: html_document
---


# Libraries

```{r install_libraries, message=FALSE, warning=FALSE}




# List of packages to check, install if needed, and load
packages <- c(
  "broom", "car", "DALEX", "dials", "dplyr", "ggplot2", "infer", 
  "modeldata", "oz", "parsnip", "pastclim", "patchwork", "purrr", 
  "readr", "recipes", "rsample", "rworldmap", "sf", "tidyr", 
  "tidymodels", "tidysdm", "tidyterra", "tibble", "tune", "terra", 
  "workflows", "workflowsets", "gtools", "fs", "USE", "viridis", "here"
)


# Function to display a cat message
cat_message <- function(pkg, message_type) {
  if (message_type == "installed") {
    cat(paste0(pkg, " has been installed successfully!\n"))
  } else if (message_type == "loading") {
    cat(paste0(pkg, " is already installed and has been loaded!\n"))
  }
}

# Install missing packages and load them
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "pastclim") {
      devtools::install_local("C:/github/pastclim")
    } else if (pkg == "USE") {
      devtools::install_github("danddr/USE")
    } else {
        install.packages(pkg)
      }
    
    cat_message(pkg, "installed")
  } else {
    cat_message(pkg, "loading")
  }
  library(pkg, character.only = TRUE)
}

# If you are using renv, you can snapshot the renv after loading all the packages.

# renv::snapshot()
```

```{r libraries}
# Load necessary libraries for data manipulation, spatial analysis, plotting, and modeling
library(broom)
library(car)
library(DALEX)
library(dials)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
library(dplyr)
library(USE)
library(cowplot)
library (here)

i_am ("Analyes/Krapp apicalis finished.Rmd")

# Directories
results_dir   <- here("Results/stratified enviromental sampling", "apicalis")
processed_dir <- here("Data", "processed")

# Ensure results_dir exists
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

```

# Set the working directory

```{r Set the working directory}


# Define explainer output path (overwritten by future runs)
plot_path <- file.path(results_dir, "explainer.txt")

# Make plot_path available to the explainer script
assign("plot_path", plot_path, envir = .GlobalEnv)

# Run the explainer script
source("C:/github/create_explainer/create_explainer.R")

```

# Preparing your data

## Download Krapp2021 dataset

```{r download climate}
# Download the Krapp2021 dataset using pastclim package
pastclim::download_dataset("Krapp2021")
```

## Read apicalis data

```{r read file}

# Read in the 'apicalis_flat.csv' file, print its contents, convert to an sf object, and set the CRS
apicalis <- read_csv(file.path(processed_dir, "apicalis_flat.csv")) %>%
  {print(.); .} %>%  # Print the data to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>%  # Convert to a simple features (sf) object using latitude and longitude
  st_set_crs(4326)  # Set the Coordinate Reference System (CRS) to EPSG:4326 (GDA2020)
```

## Land Mask

### Create an extent for Australia

```{r}
# Define an extent object for Australia using specific longitude (X) and latitude (Y) coordinates
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)


```

### Krapp2021 Land Mask

```{r}
# Create a land mask for the current time using the 'Krapp2021' dataset from the 'pastclim' package
# This mask will restrict our area of interest to Australia
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>%  # Crop to the Australian extent
  aggregate(fact = 5, fun = "mean", na.rm = TRUE)  # Aggregate for better resolution (factor of 5), removing NAs
```

## Plot apicalis Occurrence Data

```{r Plot the data using ggplot2}
# Plot the apicalis occurrence points with the current land mask using ggplot2
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_sf(data = apicalis, aes(col = time_bp))  # Plot apicalis occurrence points, colored by time_bp
```

## Create Climate Time Series Using Krapp2021

```{r create climate time series}
# Create a time series of palaeoclimate reconstructions using the 'Krapp2021' dataset
# Variables used: bio05 (max temp), bio06 (min temp), bio12 (annual precipitation)
climate_vars <- c('bio05','bio06','bio12')
climate_full <- pastclim::region_series(
  bio_variables = climate_vars,  # Specify the variables we are interested in
  data = "Krapp2021",  # Use the Krapp2021 dataset
  crop = vect(Aust_extent)  # Crop the climate data to the extent of Australia
)


```

## Spatial Filtering

```{r spatial filtering}
set.seed(123)
apicalis <- thin_by_dist(apicalis,
  dist_min = km2m(50),

)
nrow(apicalis)
```


## Pseudo-absence Sampling

Stratified sampling in Environmental space

```{r build environmental space}

env_space_1950 <- c(climate_full[[1]][[800]], climate_full[[2]][[800]],
                    climate_full[[3]][[800]])
```



#```{r grid resolution optimisation}
# THIS TAKES AGES BUT ONLY NEEDS TO BE DONE ONCE


   # 1. Extract environmental values from the raster
      env_values <- terra::as.data.frame(env_space_1950, xy = TRUE, na.rm = TRUE)

   # 2. Perform PCA on the environmental variables
   #Check the column names in env_values 
   pca_res <- prcomp(env_values[, c("bio05_0", "bio06_0", "bio12_0")], scale. = TRUE)


    # 3. Create an sf object with the first two PC scores as the coordinates
    env_pca_sf <- st_as_sf(data.frame(PC1 = pca_res$x[, 1],
                                 PC2 = pca_res$x[, 2],
                                  x = env_values$x,
                                  y = env_values$y),
                       coords = c("PC1", "PC2"))

     opt_res <- optimRes(env_pca_sf, grid.res = seq(1, 100, by = 1), perc.thr = 10, cr = 1, showOpt = TRUE)

     optimal_grid_res <- opt_res[[2]]
     print(optimal_grid_res) 
#```


```{r Pseudo absence sampling}
library(scales)  # for alpha()
apicalis3 <- paSampling(
  env.rast = env_space_1950,
  pres = apicalis %>% filter(time_bp == -300),
  thres = 0.75, # you can tune this
  n.tr = 9,
  grid.res = 9 # calculated in optimisation step above, but entered manually 
) |> st_drop_geometry()





# Load world map and convert to sf object
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map)  # Do not filter, keeping all land



pseudoabsence_map <- ggplot() +
  geom_sf(data = australia_outline, 
          fill = alpha("#C4C4C4", 0.5), colour = "grey50") +

  # Pseudo Absences (circle with border)
  geom_point(data = apicalis3, 
             aes(x, y, shape = "Pseudo Absence", fill = "Pseudo Absence"), 
             colour = "black", size = 2.5, stroke = 0.3) +

  # Presences (triangle with border)
  geom_sf(data = apicalis |> filter(time_bp == -300), 
          aes(fill = "Presence", shape = "Presence"), 
          colour = "black", size = 2.5, stroke = 0.3) +

  # Manual shape and colour
  scale_fill_manual(
    values = c("Presence" = "#A100FA", "Pseudo Absence" = "#FF7E1D"),
    name = "Legend"
  ) +
  scale_shape_manual(
    values = c("Presence" = 24, "Pseudo Absence" = 21),
    name = "Legend"
  ) +

  # Set spatial limits and grid breaks
  coord_sf(
    xlim = c(110, 155),
    ylim = c(-45, -10),
    expand = FALSE,
    default_crs = NULL
  ) +
  scale_x_continuous(breaks = seq(110, 155, by = 5)) +
  scale_y_continuous(breaks = seq(-45, -10, by = 5)) +

  # Labels and theme
  labs(
    title = "Pseudo Absences in Environmental Space",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    panel.grid.major = element_line(colour = "#C4C4C4", linewidth = 0.3),
    panel.grid.minor = element_line(colour = "#C4C4C4", linewidth = 0.2),
    panel.border = element_rect(colour = "#C4C4C4", fill = NA, linewidth = 0.3)
  )

# Save the output
ggsave(
  filename = file.path(results_dir, "pseudo_absences_envspace.png"),
  plot = pseudoabsence_map,
  width = 6,
  height = 5,
  dpi = 300,
  bg = "white"
)


```

## Retrieve Climate Data for This Region

```{r extract climate for region}
# Convert apicalis to a data frame and extract spatial coordinates (longitude, latitude)
apicalis_df <- apicalis %>%
  dplyr::bind_cols(sf::st_coordinates(apicalis)) %>%  # Bind spatial coordinates to the data
  rename(x = X, y = Y) |>
  mutate(percP = "pres") |>
  bind_rows(apicalis3) |>
  mutate(time_bp = -300) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  dplyr::select(-geometry)  # Remove geometry column

# Retrieve climate data for the apicalis occurrence points from the climate series
apicalis_df <- location_slice_from_region_series(apicalis_df,
                                                 region_series = climate_full)  # Use the full climate series for extraction
```

## Add Climate Reconstructions to the sf Object

```{r add climate to sf}
apicalis4 <- apicalis %>%
  dplyr::bind_cols(sf::st_coordinates(apicalis)) %>%  # Bind spatial coordinates to the data
  rename(x = X, y = Y) |>
  mutate(percP = "pres") |>
  bind_rows(apicalis3) |>
  mutate(time_bp = -300) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  dplyr::select(-geometry) |># Remove geometry column
  bind_cols(apicalis_df[, c("bio05", "bio06", "bio12")]) %>%  # Bind the climate data to the sf object
  filter(!is.na(.$bio05)) |>
  mutate(percP = if_else(percP == "pabs", "pseudoabsence", "presence")) |>  # Mutate percP to the correct values
  st_as_sf(coords = c("x", "y")) # Convert back to sf object                               
```

# Fit the model by cross-validation

## Set Up a Recipe for Modeling

```{r make recipe}
# Define a recipe for the model where 'class' is the outcome and other variables are predictors
# Geometry is automatically ignored as a predictor for sf objects
apicalis_rec <- recipe(apicalis4, formula = percP ~ bio05 + bio06 + bio12)  # Define the model formula
apicalis_rec  # Check the recipe
```

## Build a Workflow Set for Model Tuning

```{r specify models}
# Create a workflow set of different models for tuning
# We will use glm, gam, random forest, and boosted trees as our models
apicalis_models <- workflow_set(
  preproc = list(default = apicalis_rec),  # Use the recipe for preprocessing
  models = list(
    glm = sdm_spec_glm(),  # Generalized Linear Model
    gam = sdm_spec_gam(),  # Generalized Additive Model
    rf = sdm_spec_rf(),    # Random Forest with tunable hyperparameters
    gbm = sdm_spec_boost_tree()  # Boosted Tree model with tunable hyperparameters
  ),
  cross = TRUE  # Cross all combinations of preprocessing and models
) %>%
  # Update the workflow model for GAMs with a formula specifying smooth terms for continuous predictors
  update_workflow_model("default_gam",
                        spec = sdm_spec_gam(),
                        formula = percP ~ s(bio05, k = 10) + s(bio06, k = 10) + s(bio12, k = 10)) %>%
  # Adjust controls to store information needed for the ensemble model
  option_add(control = control_ensemble_grid())
```

## Spatial Block Cross-Validation for Model Tuning

```{r set up cross validation}
# Set up spatial block cross-validation with 5 folds for tuning and assessing models
set.seed(1005)  # Set seed for reproducibility
apicalis_cv <- spatial_block_cv(apicalis4, v = 5)  # Create spatial block cross-validation with 5 folds
autoplot(apicalis_cv)  # Plot the cross-validation folds
```

## Tune and Assess Models

```{r tune models}
# Use block cross-validation folds to tune and assess models
set.seed(123)  # Set seed for reproducibility
apicalis_models <- apicalis_models %>%
  workflow_map("tune_grid",  # Tune hyperparameters using grid search
               resamples = apicalis_cv,  # Use the cross-validation resamples
               grid = 5,  # Use a grid size of 5 for tuning
               metrics = sdm_metric_set(),  # Use the default set of SDM metrics
               verbose = TRUE)  # Print verbose output
```

## Create an Ensemble Model


```{r}
library(purrr)
library(dplyr)
library(ggplot2)
library(tidyr)

# 1. Define your model and metric labels
workflow_labels <- c(
  default_gam = "Generalised Additive Model",
  default_gbm = "Boosted Tree",
  default_glm = "Generalised Linear Model",
  default_rf  = "Random Forest"
)

model_order <- c(
  "Generalised Additive Model",
  "Boosted Tree",
  "Generalised Linear Model",
  "Random Forest"
)

metric_labels <- c(
  boyce_cont = "Boyce Index",
  roc_auc = "AUC",
  tss_max = "TSS"
)

colour_map <- c(
  "Generalised Additive Model" = "#FF61C3",
  "Boosted Tree"               = "#00B140",
  "Generalised Linear Model"   = "#00BFFF",
  "Random Forest"              = "#BC43FF"
)

# 2. Extract fold-level performance from tuned models
perf_data <- apicalis_models %>%
  mutate(metrics = map(result, collect_metrics, summarize = FALSE)) %>%
  select(wflow_id, metrics) %>%
  unnest(metrics)

# 3. Clean and label the data
perf_data_clean <- perf_data %>%
  filter(.metric %in% names(metric_labels)) %>%
  mutate(
    model = factor(workflow_labels[wflow_id], levels = model_order),
    metric_label = factor(metric_labels[.metric], levels = c("Boyce Index", "AUC", "TSS"))
  )

# 4. Plot using boxplot and jitter
ggplot(perf_data_clean, aes(x = model, y = .estimate, fill = model)) +
  geom_boxplot(width = 0.6, alpha = 0.9, outlier.shape = NA) +
  geom_jitter(
    aes(colour = model),
    size = 1.3,
    width = 0.15,
    alpha = 0.4,
    stroke = 0.3,
    shape = 21
  ) +
  scale_fill_manual(values = colour_map, name = "Model Type") +
  scale_colour_manual(values = rep("black", length(model_order)), guide = "none") +
  facet_wrap(~ metric_label, scales = "free_y") +
  labs(
    title = "Model Performance Across Algorithms (Fold-Level)",
    x = NULL,
    y = "Performance Metric"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "#EEEEEE", colour = NA),
    plot.background = element_rect(fill = "white", colour = NA),
    panel.grid.major = element_line(colour = "white"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "right",
    legend.direction = "vertical",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 13),
    strip.text = element_text(margin = margin(b = 5))
  )


# 6. Save and generate metadata
plot_path <- file.path(results_dir, "ensemble", "ensemble_performance_clean.png")

ggsave(
  filename = plot_path,
  width = 8,
  height = 5,
  dpi = 300
)


# Add metadata
source("C:/github/create_explainer/create_explainer.R")

```
# Projecting to other times

## Make Predictions for the Last Glacial Maximum (LGM)

```{r create LGM}
# Create a climate raster for the Last Glacial Maximum (LGM) using the correct layer (779) for 21,000 BP
climate_lgm <- rast(list(climate_full[[1]][[538]], climate_full[[2]][[538]], climate_full[[3]][[538]]))  # Extract 20,000 BP layers
names(climate_lgm) <- c("bio05", "bio06", "bio12")  # Rename the layers to match the climate variables


```

## Predict Using the Ensemble Model for LGM

```{r predict lgm}
# Use the ensemble model to predict species occurrence probabilities for the Last Glacial Maximum (LGM)
apicalis_prediction_lgm <- predict_raster(apicalis_ensemble, climate_lgm)  # Predict LGM probabilities

# Plot the predicted probabilities using ggplot2
ggplot() +
  geom_spatraster(data = apicalis_prediction_lgm, aes(fill = mean)) +  # Add the prediction raster to the plot
  scale_fill_terrain_c(name = "Probability\nof occurrence")  # Use a terrain color scale to show probabilities


# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Define the point location (convert to a data frame for ggplot)
point_location <- data.frame(
  longitude = 118.965510,
  latitude = -24.270740  # South is negative in latitude
)

# Plot the predicted probabilities with an Australia outline and the red dot
ggplot() +
  geom_spatraster(data = apicalis_prediction_lgm, aes(fill = mean)) +  # Add the prediction raster
  scale_fill_terrain_c(name = "Probability\nof occurrence") +  # Use a terrain color scale
  geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +  # Add black outline
  geom_point(data = point_location, aes(x = longitude, y = latitude), color = "red", size = 1) +  # Add red dot
  coord_sf(xlim = c(112, 150), ylim = c(-40, -10))  # Crop to 10°S–40°S latitude and 110°E–150°E longitude




```

## save model as spatraster

```{r save_model_spatraster}

# Initialize an empty SpatRaster stack
apicalis_spatrast <- NULL

# Loop through 40 time steps (40ka in 1ka timesteps) to create and stack rasters
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present for each timestep
  
  # Create a climate snapshot for the current timestep
  climate_snapshot <- rast(list(
    climate_full[[1]][[timestep]], 
    climate_full[[2]][[timestep]], 
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename layers to match the climate variables

  # Predict species occurrence for the current climate snapshot
  apicalis_prediction_raster <- predict_raster(apicalis_ensemble, climate_snapshot)
  
  # Stack the prediction raster along the z-axis
  apicalis_spatrast <<- if (is.null(apicalis_spatrast)) {
    apicalis_prediction_raster
  } else {
    c(apicalis_spatrast, apicalis_prediction_raster)
  }
})

# Assign time steps to the z-axis of the stacked raster
time(apicalis_spatrast) <- time(climate_full[[1]])[760:800] - 1950

# Save the stacked SpatRaster to a file
output_file <- file.path(results_dir, "predictions_stack.tif")
writeRaster(apicalis_spatrast, filename = output_file, overwrite = TRUE)

```

## stack o maps

```{r stack o maps}
library(magick)

# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Loop over 40 time steps (40 ka in 1 ka timesteps)
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present

  # Create climate snapshot
  climate_snapshot <- rast(list(
    climate_full[[1]][[timestep]], 
    climate_full[[2]][[timestep]], 
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")

  # Predict suitability
  apicalis_prediction1 <- predict_raster(apicalis_ensemble, climate_snapshot)

  # Create plot
  p <- ggplot() +
    geom_spatraster(data = apicalis_prediction1, aes(fill = mean)) +
    scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +
    coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
    ggtitle(bquote(italic("Leporillus apicalis") * " — Years BP: " * .(year_bp))) +
    theme_minimal() +
    theme(
   plot.background = element_rect(fill = "white", colour = NA),     
  panel.background = element_rect(fill = "#F2F2F2", colour = NA),
  panel.grid.major = element_line(colour = "#C4C4C4"),
  panel.grid.minor = element_line(colour = "#C4C4C4"),
      plot.margin = margin(0, 0, 0, 0, unit = "pt")
    )

# Define output path
plot_path <- file.path(
  results_dir, "model",
  paste0("apicalis_map_", year_bp, ".png")
)

# Save PNG
ggsave(
  filename = plot_path,
  plot = p,
  width = 7,
  height = 7,
  dpi = 300,
  bg = "transparent"
)

# Keep plot_path available globally
assign("plot_path", plot_path, envir = .GlobalEnv)

# Add metadata
source(here("create_explainer", "create_explainer.R"))


  # Trim and add white border
  img <- image_read(plot_path)
  img_trimmed <- image_trim(img)
  img_bordered <- image_border(img_trimmed, color = "white", geometry = "30x30")
  image_write(img_bordered, path = plot_path)
})




```

```{r 6 image panel}
library(ggplot2)
library(patchwork)
library(terra)
library(purrr)
library(dplyr)

# Extract raster time steps
raster_times <- terra::time(apicalis_spatrast)

# Define target slices
target_times <- c(-40000, -26000, -21000, -12000, -6000, 0)
matched_indices <- match(target_times, raster_times)
panel_tags <- letters[1:length(target_times)]

# Define colours
bg_colour <- "#DCF3FC"
outline_colour <- "black"

# Create each panel (no middens)
panel_plots <- pmap(
  list(idx = matched_indices, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- apicalis_spatrast[[idx]]

    ggplot() +
      geom_spatraster(data = current_prediction, aes(fill = mean)) +
      scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, colour = "grey20", linewidth = 0.3) +
      coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(year_bp))) +
      theme_minimal(base_size = 10) +
      theme(
        panel.background = element_rect(fill = bg_colour, colour = NA),
        plot.background = element_rect(fill = "white", colour = NA),
        panel.grid.major = element_line(colour = "white"),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 8),
        axis.title = element_blank(),
        plot.margin = margin(2, 2, 2, 2),
        legend.position = "right",
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 11)
      ) +
      guides(
        fill = guide_colourbar(
          title.theme = element_text(size = 11),
          label.theme = element_text(size = 11)
        )
      )
  }
)

# Combine into multi-panel figure
final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +
  plot_annotation(
    title = expression(italic("Leporillus apicalis") * " SDM model predictions"),
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )
  )

# Save the panel
ggsave(
  filename = file.path(results_dir, "apicalis_model_panel.png"),
  plot = final_panel,
  width = 10,
  height = 9,
  dpi = 300
)

```



##Centroid
```{r}
# Calculate suitability-weighted centroid for each timestep in apicalis_spatrast

apicalis_centroids <- lapply(1:nlyr(apicalis_spatrast), function(i) {
  r <- apicalis_spatrast[[i]]

  # Extract suitability values and matching coordinates (including NAs)
  vals <- terra::values(r)[, 1]
  coords <- terra::crds(r, na.rm = FALSE)

  # Filter to cells with non-NA and positive suitability values
  valid <- !is.na(vals) & vals > 0
  if (sum(valid) == 0) return(NULL)  # Skip if no valid cells

  weights <- vals[valid]
  coords <- coords[valid, ]

  # Calculate weighted centroid
  centroid_x <- sum(coords[, 1] * weights) / sum(weights)
  centroid_y <- sum(coords[, 2] * weights) / sum(weights)

  data.frame(
    time_bp = time(apicalis_spatrast)[i],
    longitude = centroid_x,
    latitude = centroid_y
  )
}) %>% bind_rows()


# Rainbow colours: red = recent, violet = old
viridis_colours <- viridis(41, option = "H")



ggplot(apicalis_centroids, aes(x = longitude, y = latitude)) +
  geom_path(colour = "grey50", linewidth = 1) +
  geom_point(aes(colour = time_bp), size = 2) +
 scale_colour_gradientn(
  colours = viridis_colours,
  name = "Years BP",
  limits = c(-40000, 0),
  breaks = c(-40000, -30000, -20000, -10000, 0),
  labels = c("40k", "30k", "20k", "10k", "0")
) +
  coord_fixed(xlim = c(129, 131), ylim = c(-28, -24.8)) +  # Consistent extent
  labs(
    title = "Centroid shift of predicted apicalis habitat through time",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
   theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    panel.grid.major = element_line(colour = "#C4C4C4", linewidth = 0.3),
    panel.grid.minor = element_line(colour = "#C4C4C4", linewidth = 0.2),
    panel.border = element_rect(colour = "#C4C4C4", fill = NA, linewidth = 0.3)
  )


# Save output with clean background
ggsave(
  filename = file.path(results_dir, "apicalis_centroid_shift_plot.png"),
  plot = last_plot(),
  width = 6,
  height = 6,
  dpi = 300,
  bg = "white"
)


# Identify the timestep with the northernmost centroid
northernmost <- apicalis_centroids %>% filter(latitude == max(latitude))
print(northernmost)

```


# **Model ends here.**

# Additional steps for validation and analysis



## Load and organise midden data set
```{r load midden data set}
## Load and organise midden data set
#Midden data set is used for additional validation

# Load the independent dataset
dated_middens <- read.csv(file.path(processed_dir, "datedmiddens.csv"))


# Convert to sf object (assuming it contains longitude and latitude columns)
middens_sf <- dated_middens %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # Adjust column names if different

# Reproject to match the CRS of the apicalis_spatrast
middens_sf <- st_transform(middens_sf, crs = crs(apicalis_spatrast))

# Convert sf object to SpatVector for compatibility with terra::extract
middens_vect <- terra::vect(middens_sf)

# --- Time-matched extraction (fix for [,2]) ---
# Model time steps (e.g. -40000 .. 0)
rt <- as.numeric(terra::time(apicalis_spatrast))

# For each midden, find nearest model time and its column index
middens_sf <- middens_sf |>
  dplyr::mutate(
    binned_time_bp = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
    nearest_idx    = match(binned_time_bp, rt),
    lubridate_time = ybp2date(time_bp)
  )

# Extract predictions from ALL layers once (no ID column)
extracted_values <- terra::extract(apicalis_spatrast, middens_vect, ID = FALSE)

# Assign the per-row, time-correct value
vals_mat <- as.matrix(extracted_values)
middens_sf$predicted <- vals_mat[cbind(seq_len(nrow(middens_sf)), middens_sf$nearest_idx)]
# -----------------------------------------------

# Create a data frame with raster timesteps
raster_timesteps <- rt
timesteps_df <- data.frame(raster_timestep = raster_timesteps)

# Count occurrences per timestep (and mean predicted suitability)
organized_data <- timesteps_df %>%
  left_join(
    middens_sf %>%
      group_by(binned_time_bp) %>%
      summarise(
        count = n(),
        mean_value = mean(predicted, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("raster_timestep" = "binned_time_bp")
  ) %>%
  mutate(
    count = replace_na(count, 0),
    mean_value = replace_na(mean_value, NA_real_)
  )

print(organized_data)


```
```{r plot middens}
library(rworldxtra)

# Load world map data and filter for Australia
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")  # Keep only Australia

# Extract raster timesteps
raster_timesteps <- time(apicalis_spatrast)

# Loop through each timestep
walk(seq_along(raster_timesteps), function(i) {
  current_year_bp <- raster_timesteps[i]

  # Filter middens within ±500 years of the timestep
  current_middens <- middens_sf %>%
    filter(abs(time_bp - current_year_bp) <= 500)

  # Get prediction layer
  current_prediction <- apicalis_spatrast[[i]]

  # Create plot
  p <- ggplot() +
    geom_spatraster(data = current_prediction, aes(fill = mean)) +
    scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +
    geom_sf(data = current_middens, shape = 21, fill = "#FF3399", colour = "black", size = 2, stroke = 0.4)  +    coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
    ggtitle(bquote(italic("Leporillus apicalis") * " SDM with Middens — Years BP: " * .(current_year_bp))) +
    theme_minimal(base_size = 11) +
theme(
  plot.background = element_rect(fill = "white", colour = NA),     
  panel.background = element_rect(fill = "#F2F2F2", colour = NA),
  panel.grid.major = element_line(colour = "#C4C4C4"),
  panel.grid.minor = element_line(colour = "#C4C4C4"),
  plot.margin = margin(0, 0, 0, 0, "pt")
    )

  # Define save path
plot_path <- file.path(
  results_dir, "midden maps",
  paste0("apicalis_middens_", current_year_bp, ".png")
)

# Save plot with white background
ggsave(
  filename = plot_path,
  plot = p,
  width = 7,
  height = 7,
  dpi = 300,
  bg = "white"
)

assign("plot_path", plot_path, envir = .GlobalEnv)

# Add metadata
source(here("create_explainer", "create_explainer.R"))


  # Trim and add white border
  img <- magick::image_read(plot_path)
  img_trimmed <- magick::image_trim(img)
  img_bordered <- image_border(img_trimmed, color = "white", geometry = "30x30")
  magick::image_write(img_bordered, path = plot_path)
})

```





#Make 6 panel plot

```{r make 6 panel plot}

# Extract raster time slices
raster_times <- terra::time(apicalis_spatrast)

# Select target time steps
target_times <- c(-28000, -15000, -12000, -7000, -4000, 0)
matched_indices <- match(target_times, raster_times)
panel_tags <- letters[1:length(target_times)]

# Define colours
point_colour <- "#FF3399"
outline_colour <- "black"

# Create individual panels
panel_plots <- pmap(
  list(idx = matched_indices, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- apicalis_spatrast[[idx]]
    current_middens <- middens_sf %>%
      filter(abs(time_bp - year_bp) <= 500) %>%
      mutate(type = "Midden")

    ggplot() +
      geom_spatraster(data = current_prediction, aes(fill = mean)) +
      scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, colour = outline_colour, linewidth = 0.5) +
      geom_sf(data = current_middens,
              aes(colour = type, shape = type),
              fill = point_colour, size = 2, stroke = 0.4, show.legend = TRUE) +
      scale_colour_manual(name = NULL, values = c("Midden" = "black")) +
      scale_shape_manual(name = NULL, values = c("Midden" = 21)) +
      coord_sf(xlim = c(112, 150), ylim = c(-40, -10)) +
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(abs(year_bp)))) +
      theme(
        plot.background = element_rect(fill = "white", colour = NA),
        panel.background = element_rect(fill = "#E8E8E8", colour = NA),
        panel.grid.major = element_line(colour = "#C4C4C4"),
        panel.grid.minor = element_line(colour = "#C4C4C4"),
        axis.text = element_text(size = 8),
        axis.title = element_blank(),
        plot.margin = margin(2, 2, 2, 2),
        legend.position = "bottom",
        legend.box = "horizontal",
        legend.justification = "center",
        legend.spacing.x = unit(1.2, "cm"),
        legend.spacing.y = unit(0.4, "cm"),
        legend.key = element_rect(fill = "transparent", colour = NA),
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 11)
      ) +
      guides(
  fill = guide_colourbar(
    title = "Probability of occurrence",
    title.position = "top",
    title.theme = element_text(size = 11),
    label.theme = element_text(size = 10),
    direction = "horizontal",
    barwidth = unit(5, "cm"),
    barheight = unit(0.4, "cm")
  ),
  colour = guide_legend(
    override.aes = list(size = 2.5),
    title.theme = element_text(size = 11),
    label.theme = element_text(size = 11)
  ),
  shape = guide_legend(
    title.theme = element_text(size = 11),
    label.theme = element_text(size = 11)
  )
)
  }
)

# Combine panels and apply shared legend theme correctly
final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +
  plot_annotation(
    title = "Leporillus apicalis SDM with middens",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )
  ) &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.justification = "center",
    legend.spacing.x = unit(1.2, "cm"),
    legend.spacing.y = unit(0.4, "cm"),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )

# Save figure
ggsave(
  filename = file.path(results_dir, "midden_panel_plot.png"),
  plot = final_panel,
  width = 10,
  height = 9,
  dpi = 300
)


```
# Calculate TSS 
The model output is a continuous suitability gradient. for TSS to work we need to define suitability binary. To do this we need to define a threshold - above which is suitable and below is unsuitable. We use MasSS to calculate and define the optimal threshold.

```{r true skills statistic using midden data MaxSS, message=FALSE, warning=FALSE}

# -----------------------------------------------------------------
# Step 1: Initial Binary Conversion & Pseudo-absence Sampling (initial threshold)
# -----------------------------------------------------------------

# Use an initial threshold (0.27) to create a preliminary binary raster.
threshold_init <- 0.27
apicalis_binary_init <- app(x = apicalis_spatrast, fun = \(i) i > threshold_init)
time(apicalis_binary_init) <- time(apicalis_spatrast) |> pastclim::ybp2date()
apicalis_rs_init <- sds(apicalis_binary_init)
names(apicalis_rs_init) <- "pres_abs"

# Sample pseudo-absences using the initial binary raster.
pseudoabs_init <- sample_pseudoabs_time(
  data          = dated_middens,
  raster        = apicalis_rs_init,
  n_per_presence = 3,                     # 3 pseudo-absences per presence
  time_col      = "time_bp",
  lubridate_fun = pastclim::ybp2date,
  method        = c("dist_min", km2m(200)),
  time_buffer   = 365000 * 2              # ~2 ka buffer in days
)

# -----------------------------------------------------------------
# Step 2: Extract continuous predictions time-matched to each point
# -----------------------------------------------------------------

library(sf)
library(terra)
library(dplyr)

# Model time steps in years BP (e.g., -40000 .. 0)
rt <- as.numeric(terra::time(apicalis_spatrast))

## PRESENCES (dated middens)
presence_sf <- st_as_sf(dated_middens, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs(apicalis_spatrast)) |>
  mutate(
    nearest_time = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
    nearest_idx  = match(nearest_time, rt)
  )

# Extract from all layers once, then pick per-row column by nearest_idx
pres_mat <- terra::extract(apicalis_spatrast, terra::vect(presence_sf), ID = FALSE)
presence_vals <- as.matrix(pres_mat)[cbind(seq_len(nrow(presence_sf)), presence_sf$nearest_idx)]

## PSEUDO-ABSENCES (from pseudoabs_init; time_step is a Date)
pseudoabs_sf <- st_as_sf(pseudoabs_init, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs(apicalis_spatrast)) |>
  mutate(
    time_bp      = pastclim::date2ybp(time_step),
    nearest_time = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
    nearest_idx  = match(nearest_time, rt)
  )

pseudo_mat <- terra::extract(apicalis_spatrast, terra::vect(pseudoabs_sf), ID = FALSE)
pseudoabs_vals <- as.matrix(pseudo_mat)[cbind(seq_len(nrow(pseudoabs_sf)), pseudoabs_sf$nearest_idx)]

## Combine + clean
predicted_values <- c(presence_vals, pseudoabs_vals)
observed <- c(rep(1, length(presence_vals)), rep(0, length(pseudoabs_vals)))
keep <- is.finite(predicted_values) & is.finite(observed)
predicted_values <- predicted_values[keep]
observed <- observed[keep]

# -----------------------------------------------------------------
# Step 3: Calculate the Optimal Threshold Using maxSSS
# -----------------------------------------------------------------

thresholds <- seq(0, 1, by = 0.01)
results <- data.frame(threshold = thresholds, TSS = NA_real_)

for (i in seq_along(thresholds)) {
  thresh <- thresholds[i]
  pred_binary <- ifelse(predicted_values > thresh, 1L, 0L)

  # Confusion matrix components
  a <- sum(pred_binary == 1L & observed == 1L)  # True Positives
  b <- sum(pred_binary == 1L & observed == 0L)  # False Positives
  c <- sum(pred_binary == 0L & observed == 1L)  # False Negatives
  d <- sum(pred_binary == 0L & observed == 0L)  # True Negatives

  # Skip thresholds where either class is missing
  if ((a + c) == 0 || (b + d) == 0) next

  sensitivity <- a / (a + c)
  specificity <- d / (b + d)
  results$TSS[i] <- sensitivity + specificity - 1
}

opt_threshold <- results$threshold[which.max(results$TSS)]
cat("Optimal Threshold (maxSSS):", opt_threshold, "\n")

# -----------------------------------------------------------------
# Step 4: Re-create Binary Raster & Pseudo-Absences Using the Optimal Threshold
# -----------------------------------------------------------------

# Recompute the binary raster using the optimal threshold.
apicalis_binary <- app(x = apicalis_spatrast, fun = \(i) i > opt_threshold)
time(apicalis_binary) <- time(apicalis_spatrast) |> pastclim::ybp2date()
apicalis_rs <- sds(apicalis_binary)
names(apicalis_rs) <- "pres_abs"

# Re-sample pseudo-absences using the updated binary raster.
pseudoabs <- sample_pseudoabs_time(
  data          = dated_middens,
  raster        = apicalis_rs,
  n_per_presence = 3,
  time_col      = "time_bp",
  lubridate_fun = pastclim::ybp2date,
  method        = c("dist_min", km2m(200)),
  time_buffer   = 365000 * 2
)

# Quick visual sanity check (optional)
ggplot2::ggplot() +
  tidyterra::geom_spatraster(data = land_mask_current, ggplot2::aes()) +
  tidyterra::scale_fill_terrain_c() +
  ggplot2::geom_point(data = pseudoabs, ggplot2::aes(col = time_step, x = longitude, y = latitude)) +
  ggplot2::geom_point(data = dated_middens, ggplot2::aes(x = longitude, y = latitude), colour = "red")

# -----------------------------------------------------------------
# Step 5: Evaluate with the Updated Binary Predictions
# -----------------------------------------------------------------

# Create data frame in correct format for location_slice_from_region_series
pres_abs_observed <- pseudoabs %>%
  mutate(time_bp = pastclim::date2ybp(time_step)) %>%
  select(-class, -time_step) %>%
  tibble::rowid_to_column("name")

# Extract presence/absence from SDM at the matched times/locations
# (Note: location_slice_from_region_series is a custom function on your system.)
pres_abs_pred <- location_slice_from_region_series(
  x = pres_abs_observed,
  region_series = apicalis_rs
)

# Restore class labels (order is preserved from sample_pseudoabs_time)
pres_abs_pred$class <- pseudoabs$class

# Confusion matrix components
a <- nrow(dplyr::filter(pres_abs_pred, class == "presence",  pres_abs == 1))
b <- nrow(dplyr::filter(pres_abs_pred, class == "pseudoabs", pres_abs == 1))
c <- nrow(dplyr::filter(pres_abs_pred, class == "presence",  pres_abs == 0))
d <- nrow(dplyr::filter(pres_abs_pred, class == "pseudoabs", pres_abs == 0))

tpr <- a / (a + c)  # sensitivity
tnr <- d / (b + d)  # specificity
tss <- tpr + tnr - 1
sorensen <- 2 * (a / (c + (2 * a) + d))
f2score <- 2 * (((a / (a + b)) * tpr) / ((a / (a + b)) + tpr))

cat(sprintf("True Skill Statistic: %.3f\n", tss))
cat(sprintf("Sorensen: %.3f\n", sorensen))
cat(sprintf("F2 score: %.3f\n", f2score))



```
```{r}
# =========================
# Panels (a) & (b): full + coarse; combined publication figure (vertical, no titles)
# =========================
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(ggplot2)
  library(sf)
  library(terra)
  library(cowplot)
})

# ---------- Inputs ----------
stopifnot(inherits(apicalis_spatrast, "SpatRaster"),
          inherits(middens_sf, "sf"),
          dir.exists(results_dir))

# ---------- Global settings ----------
set.seed(42)
step_years   <- 1000
rt           <- as.numeric(terra::time(apicalis_spatrast))
thr_a <- if (exists("opt_threshold")) opt_threshold else 0.27   # Panel (a)
thr_b <- 0.35                                                   # Panel (b)
lookback_ka     <- 30
include_current <- TRUE
coarse_factor   <- 2
w_single <- 7; h_single <- 5; dpi_out <- 300

# --- Resolutions for titles ---
res_full_deg   <- terra::res(apicalis_spatrast)[1]
res_coarse_deg <- res_full_deg * coarse_factor
res_full_lbl   <- sprintf("%.1f°", res_full_deg)
res_coarse_lbl <- sprintf("%.1f°", res_coarse_deg)

# ---------- Helpers ----------
ensure_time_cols <- function(m_sf, rt) {
  if (!all(c("binned_time_bp","nearest_idx") %in% names(m_sf))) {
    m_sf <- m_sf %>%
      mutate(
        binned_time_bp = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
        nearest_idx    = match(binned_time_bp, rt)
      )
  }
  m_sf
}

fd_bin_count <- function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) return(30L)
  fd_bw <- 2 * IQR(x, na.rm = TRUE) / (length(x)^(1/3))
  max(20, min(60, round(1 / max(min(fd_bw, 0.08), 0.02))))
}

time_matched_background <- function(m_sf, rast){
  rt <- as.numeric(terra::time(rast))
  m_sf <- ensure_time_cols(m_sf, rt)
  split(m_sf, m_sf$binned_time_bp) |>
    lapply(function(df_i){
      tb  <- unique(df_i$binned_time_bp)
      lyr <- match(tb, rt)
      n_i <- nrow(df_i)
      vals <- terra::spatSample(rast[[lyr]], size = n_i, method = "random",
                                values = TRUE, na.rm = TRUE)[,1] |> as.numeric()
      tibble(suitability = vals, group = "Background (time-matched)", bin = tb)
    }) |> bind_rows()
}

compute_persistence_strict <- function(rast, m_sf, thr, include_current, lookback_ka, step_years){
  rt <- as.numeric(terra::time(rast))
  m_sf <- ensure_time_cols(m_sf, rt)
  mat <- terra::extract(rast, terra::vect(m_sf), ID = FALSE) |> as.matrix()
  lookback_steps <- if (is.finite(lookback_ka)) as.integer(lookback_ka) else Inf
  start_idx <- if (!include_current) pmax(m_sf$nearest_idx - 1L, 0L) else m_sf$nearest_idx
  lead_true_count <- function(x){
    if (!length(x)) return(0L)
    n <- 0L; for (v in x) { if (isTRUE(v)) n <- n + 1L else break }; n
  }
  steps <- vapply(seq_len(nrow(mat)), function(i){
    idx <- start_idx[i]; if (idx <= 0L) return(0L)
    from <- idx
    to   <- if (is.finite(lookback_steps)) max(idx - lookback_steps + 1L, 1L) else 1L
    sidx <- seq(from, to, by = -1L)
    above <- mat[i, sidx] > thr
    lead_true_count(above)
  }, integer(1))
  tibble(persist_ka = steps * (step_years/1000))
}

# =====================================================
# Prepare data
# =====================================================
middens_sf <- ensure_time_cols(middens_sf, rt)

# coarse raster
apicalis_spatrast_coarse <- terra::aggregate(
  apicalis_spatrast, fact = coarse_factor, fun = mean, na.rm = TRUE
)
terra::time(apicalis_spatrast_coarse) <- terra::time(apicalis_spatrast)

# =====================================================
# PANEL (a): Habitat suitability at dated midden locations
# =====================================================
# --- Full resolution ---
mid_df_full <- tibble(
  suitability = sf::st_drop_geometry(middens_sf)$predicted,
  group = "Middens",
  bin   = middens_sf$binned_time_bp
)
bg_df_full   <- time_matched_background(middens_sf, apicalis_spatrast)
plot_df_full <- bind_rows(mid_df_full, bg_df_full) |> filter(is.finite(suitability))
bins_a_full  <- fd_bin_count(mid_df_full$suitability)

p_a_full <- ggplot(plot_df_full, aes(x = suitability, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.65, bins = bins_a_full,
                 boundary = 0, colour = "white") +
  geom_vline(xintercept = thr_a, linetype = "dashed", linewidth = 0.6) +
  scale_fill_manual(values = c("Middens" = "#BC43FF",
                               "Background (time-matched)" = "#9AA0A6")) +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,0.1)) +
  labs(title = sprintf("Habitat suitability at dated midden locations (%.1f°)", res_full_deg),
       x = "Modelled suitability (0–1)", y = "Count", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", colour = NA),
        panel.background = element_rect(fill = "#F2F2F2", colour = NA),
        panel.grid.major = element_line(colour = "white"),
        legend.position = "bottom")

print(p_a_full)
ggsave(file.path(results_dir, "panel_a_full_hist.png"), p_a_full,
       width = w_single, height = h_single, dpi = dpi_out, bg = "white")

# --- Coarse resolution ---
mid_vals_c <- terra::extract(apicalis_spatrast_coarse, terra::vect(middens_sf), ID = FALSE) |> as.matrix()
pick_idx   <- cbind(seq_len(nrow(mid_vals_c)), middens_sf$nearest_idx)
mid_suit_c <- mid_vals_c[pick_idx] |> as.numeric()
mid_df_c <- tibble(suitability = mid_suit_c, group = "Middens", bin = middens_sf$binned_time_bp)
bg_df_c  <- time_matched_background(middens_sf, apicalis_spatrast_coarse)
plot_df_c <- bind_rows(mid_df_c, bg_df_c) |> filter(is.finite(suitability))
bins_a_c  <- fd_bin_count(mid_df_c$suitability)

p_a_coarse <- ggplot(plot_df_c, aes(x = suitability, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.65, bins = bins_a_c,
                 boundary = 0, colour = "white") +
  geom_vline(xintercept = thr_a, linetype = "dashed", linewidth = 0.6) +
  scale_fill_manual(values = c("Middens" = "#BC43FF",
                               "Background (time-matched)" = "#9AA0A6")) +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,0.1)) +
  labs(title = sprintf("Habitat suitability at dated midden locations (%.1f°)", res_coarse_deg),
       x = "Modelled suitability (0–1)", y = "Count", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", colour = NA),
        panel.background = element_rect(fill = "#F2F2F2", colour = NA),
        panel.grid.major = element_line(colour = "white"),
        legend.position = "bottom")

print(p_a_coarse)
ggsave(file.path(results_dir, "panel_a_coarse_hist.png"), p_a_coarse,
       width = w_single, height = h_single, dpi = dpi_out, bg = "white")

# =====================================================
# PANEL (b): Duration of suitability prior to midden deposition
# =====================================================
# --- Full resolution ---
persist_full <- compute_persistence_strict(
  rast = apicalis_spatrast, m_sf = middens_sf, thr = thr_b,
  include_current = include_current, lookback_ka = lookback_ka, step_years = step_years
)
x_max_b <- if (is.finite(lookback_ka)) lookback_ka else max(persist_full$persist_ka, na.rm = TRUE)

p_b_full <- ggplot(persist_full, aes(x = persist_ka)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left",
                 colour = "white", fill = "#BC43FF", alpha = 0.9) +
  scale_x_continuous(limits = c(0, x_max_b),
                     breaks = seq(0, x_max_b, by = 10),
                     labels = function(x) paste0(x, " ka")) +
  labs(title = sprintf("Duration of suitability prior to midden deposition (%.1f°)", res_full_deg),
       x = sprintf("Years suitable before deposition (ka; suitability \u2265 %.2f)", thr_b),
       y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", colour = NA),
        panel.background = element_rect(fill = "#F2F2F2", colour = NA),
        panel.grid.major = element_line(colour = "white"))

print(p_b_full)
ggsave(file.path(results_dir, "panel_b_full_hist_thr035.png"), p_b_full,
       width = w_single, height = h_single, dpi = dpi_out, bg = "white")

# --- Coarse resolution ---
persist_coarse <- compute_persistence_strict(
  rast = apicalis_spatrast_coarse, m_sf = middens_sf, thr = thr_b,
  include_current = include_current, lookback_ka = lookback_ka, step_years = step_years
)

p_b_coarse <- ggplot(persist_coarse, aes(x = persist_ka)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left",
                 colour = "white", fill = "#BC43FF", alpha = 0.9) +
  scale_x_continuous(limits = c(0, x_max_b),
                     breaks = seq(0, x_max_b, by = 10),
                     labels = function(x) paste0(x, " ka")) +
  labs(title = sprintf("Duration of suitability prior to midden deposition (%.1f°)", res_coarse_deg),
       x = sprintf("Years suitable before deposition (ka; suitability \u2265 %.2f)", thr_b),
       y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", colour = NA),
        panel.background = element_rect(fill = "#F2F2F2", colour = NA),
        panel.grid.major = element_line(colour = "white"))

print(p_b_coarse)
ggsave(file.path(results_dir, "panel_b_coarse_hist_thr035.png"), p_b_coarse,
       width = w_single, height = h_single, dpi = dpi_out, bg = "white")

# =====================================================
# Combined publication figure (coarse panels only, stacked; NO TITLES)
# =====================================================
p_a_tag <- ggdraw(p_a_coarse + labs(title = NULL)) +
  draw_label("(a)", x = 0.02, y = 0.98, hjust = 0, vjust = 1, fontface = "bold")
p_b_tag <- ggdraw(p_b_coarse + labs(title = NULL)) +
  draw_label("(b)", x = 0.02, y = 0.98, hjust = 0, vjust = 1, fontface = "bold")

fig_ab  <- plot_grid(p_a_tag, p_b_tag, ncol = 1, rel_heights = c(1,1))
print(fig_ab)

ggsave(file.path(results_dir, "figure_histograms_panels_ab_COARSE_thrB_035_VERTICAL.png"),
       fig_ab, width = 7, height = 10, dpi = dpi_out, bg = "white")

```



# Partial Dependence Profiles and Variable Importance

```{r PDP}
# =====================================
# Partial Dependence Profiles and Variable Importance (Table version)
# =====================================
suppressPackageStartupMessages({
  library(DALEX)
  library(tidysdm)
  library(ggplot2)
  library(dplyr)
  library(knitr)
})

# ---- Set up ----
apicalis_explainer <- explain_tidysdm(apicalis_ensemble)
apicalis_outdir <- file.path(results_dir, "variable_importance")
if (!dir.exists(apicalis_outdir)) dir.create(apicalis_outdir, recursive = TRUE)

# ---- Variable importance ----
apicalis_vip <- model_parts(explainer = apicalis_explainer) %>%
  filter(!variable %in% c("_baseline_", "_full_model_"))

# Summarise mean ± SD for each variable
apicalis_vip_table <- apicalis_vip %>%
  group_by(variable) %>%
  summarise(
    mean_dropout_loss = mean(dropout_loss, na.rm = TRUE),
    sd_dropout_loss   = sd(dropout_loss, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  arrange(desc(mean_dropout_loss)) %>%
  mutate(
    mean_dropout_loss = round(mean_dropout_loss, 3),
    sd_dropout_loss   = round(sd_dropout_loss, 3)
  ) %>%
  rename(
    `Environmental variable` = variable,
    `Mean drop-out loss` = mean_dropout_loss,
    `SD` = sd_dropout_loss
  )

# Print table to console
print(apicalis_vip_table)

# Save as CSV
write.csv(
  apicalis_vip_table,
  file.path(apicalis_outdir, "apicalis_variable_importance_table.csv"),
  row.names = FALSE
)

# Save a markdown table (useful for reports)
kable(
  apicalis_vip_table,
  caption = "Variable importance summary (Leporillus apicalis ensemble model)",
  format = "markdown"
)

# =====================================
# Partial Dependence Profiles (PDPs)
# =====================================

apicalis_plot_and_save_pdp <- function(var, label) {
  apicalis_pdp <- model_profile(apicalis_explainer, N = 500, variables = var)

  apicalis_pdp_plot <- ggplot(apicalis_pdp$agr_profiles, aes(x = `_x_`, y = `_yhat_`)) +
    geom_line(colour = "darkblue", linewidth = 1) +
    labs(
      title = paste("Leporillus apicalis – Partial Dependence on", label),
      x = label,
      y = "Predicted suitability"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.background = element_rect(fill = "white", colour = NA),
      panel.background = element_rect(fill = "#F2F2F2", colour = NA),
      panel.grid.major = element_line(colour = "#C4C4C4", linewidth = 0.3),
      panel.border = element_rect(colour = "#C4C4C4", fill = NA, linewidth = 0.3),
      plot.title = element_text(face = "italic", size = 13)
    )

  ggsave(
    filename = file.path(apicalis_outdir, paste0("apicalis_pdp_", var, ".png")),
    plot = apicalis_pdp_plot,
    width = 6, height = 4, dpi = 300, bg = "white"
  )
  print(apicalis_pdp_plot)
}

# Generate and save PDPs for key predictors
apicalis_plot_and_save_pdp("bio05", "Maximum temperature of warmest month (°C)")
apicalis_plot_and_save_pdp("bio06", "Minimum temperature of coldest month (°C)")
apicalis_plot_and_save_pdp("bio12", "Annual precipitation (mm)")

```

##Kernel density estimate 
```{r}
# Load required libraries
# Load necessary libraries
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)

# Load and prepare data
dated_middens_2 <- read.csv(file.path(processed_dir, "datedmiddens.csv"))


# Convert to sf
middens_sf_2 <- st_as_sf(dated_middens_2, coords = c("longitude", "latitude"), crs = 4326)

# Extract coordinates
middens_df_2 <- middens_sf_2 %>%
  mutate(
    longitude = st_coordinates(.)[, 1],
    latitude = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry()

# Australia map
aus_outline <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "Australia")

# Dummy bbox for consistent framing
bbox_sf <- st_as_sfc(st_bbox(c(
  xmin = 108, xmax = 152,
  ymin = -46, ymax = -8
), crs = 4326)) %>% st_sf()

# Check if there are at least a few dozen distinct coordinates
print(head(middens_df_2))
print(nrow(middens_df_2))

# Add dummy points at map edges to force KDE range
pad_points <- data.frame(
  longitude = c(110, 150, 150, 110),
  latitude = c(-10, -10, -45, -45)
)

# Combine with midden data
middens_df_padded <- bind_rows(middens_df_2, pad_points)

# Plot with KDE using padded data
p <- ggplot() +
  geom_sf(data = aus_outline, fill = scales::alpha("#C4C4C4", 0.5), colour = "black") +
  stat_density_2d(
    data = middens_df_padded,
    aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    colour = NA,
    adjust = 0.4,
    alpha = 0.8,
    n = 200
  ) +
  scale_fill_viridis_c(option = "magma", name = "Density", direction = -1) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +
  labs(
    title = "Spatial Density of Dated Midden Samples",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "#F2F2F2", colour = NA),
    panel.grid.major = element_line(colour = "#C4C4C4"),
    panel.grid.minor = element_line(colour = "#C4C4C4"),
    
    # Legend styling for in-panel placement
    legend.position = c(0.03, 0.20),  # ⬅️ top-left corner
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", colour = "grey70", linewidth = 0.3),
    legend.key = element_rect(fill = "white", colour = NA),
    legend.margin = margin(6, 12, 6, 9),  # ⬅️ wider horizontal padding
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
      ) +
    guides(
      fill = guide_colourbar(
        title = "Density",
        title.position = "top",
        title.theme = element_text(size = 10),
        label.theme = element_text(size = 9),
        direction = "horizontal",
        barwidth = unit(4, "cm"),
        barheight = unit(0.4, "cm"),
        ticks = TRUE,
        ticks.colour = "black",
        frame.colour = "grey70",
        frame.linewidth = 0.3,
        label.position = "bottom",
        breaks = c(0, 0.04, 0.08)  # ⬅️ This limits to 3 ticks
      )
    )


ggsave(
  filename = file.path(here("Results/stratified enviromental sampling"), "midden_sampling_density.png"),
  plot = p,
  width = 7,
  height = 6,
  dpi = 300,
  bg = "white"
)



```
```{r endnote library}
# Define Results folder
results_root <- here("Results")

# Run EndNote export script to create citation library in this folder
source(here("create_endnote_library", "create_endnote_library.r"))


```

```{r}
lapply(c("tidysdm", "tidymodels", "workflowsets", "workflows", "tune", "parsnip", "recipes", "yardstick", "rsample", 
         "spatialsample", "dials", "dplyr", "tidyr", "purrr", "tibble", "ggplot2", "cowplot", "scales", "viridis", 
         "viridisLite", "sf", "terra", "tidyterra", "pastclim", "oz", "rworldmap", "DALEX", "ecospat", "broom", 
         "infer", "readr", "fs", "gtools", "RefManageR"), 
       function(pkg) { cat("\n---", pkg, "---\n"); print(citation(pkg)) })

```