---
title: "Krapp conditor"
author: "Annie"
date: "2024-06-20"
output: html_document
---

```{r setup, include=FALSE}
# Set global chunk options to display code by default
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries
```{r libraries}
# Load necessary libraries for data manipulation, spatial analysis, plotting, and modeling
library(broom)
library(car)
library(DALEX)
library(dials)
library(dplyr)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
```

# Set the working directory
```{r Set the working directory}
# Set the working directory to the processed data folder
knitr::opts_knit$set(root.dir =("C:/github/SNR_SDM/Data/processed"))
```

# Download Krapp2021 dataset
```{r download climate}
# Download the Krapp2021 dataset using pastclim package
pastclim::download_dataset("Krapp2021")
```

# Read conditor data
```{r read file}
# Set working directory to the processed data folder
setwd("C:/github/SNR_SDM/Data/processed")

# Read in the 'conditor.csv' file, print its contents, convert to an sf object, and set the CRS
conditor <- read_csv("conditor.csv") %>%
  {print(.); .} %>%  # Print the data to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>%  # Convert to a simple features (sf) object using latitude and longitude
  st_set_crs(4326)  # Set the Coordinate Reference System (CRS) to EPSG:4326 (GDA2020)
```

# Create an extent for Australia
```{r}
# Define an extent object for Australia using specific longitude (X) and latitude (Y) coordinates
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)
```

# Krapp2021 Land Mask
```{r}
# Create a land mask for the current time using the 'Krapp2021' dataset from the 'pastclim' package
# This mask will restrict our area of interest to Australia
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>%  # Crop to the Australian extent
  aggregate(fact = 5, fun = "mean", na.rm = TRUE)  # Aggregate for better resolution (factor of 5), removing NAs
```

# Plot conditor Occurrence Data
```{r Plot the data using ggplot2}
# Plot the conditor occurrence points with the current land mask using ggplot2
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_sf(data = conditor, aes(col = time_bp))  # Plot conditor occurrence points, colored by time_bp
```

# Create Climate Time Series Using Krapp2021
```{r create climate time series}
# Create a time series of palaeoclimate reconstructions using the 'Krapp2021' dataset
# Variables used: bio05 (max temp), bio06 (min temp), bio12 (annual precipitation)
climate_vars <- c('bio05','bio06','bio12')
climate_full <- pastclim::region_series(
  bio_variables = climate_vars,  # Specify the variables we are interested in
  data = "Krapp2021",  # Use the Krapp2021 dataset
  crop = vect(Aust_extent)  # Crop the climate data to the extent of Australia
)


```
# Sample Pseudo-Absences and Plot conditor Occurrence with Land Mask
#```{r}
# Add a 'class' column with 'presence' for all occurrence points
conditor <- conditor %>%
  mutate(class = "presence")  # Create the 'class' column

# Generate pseudo-absences
pseudo_absences <- sample_pseudoabs_time(conditor, 
                                    n_per_presence = 3,  # Generate 3 pseudo-absences per presence point
                                    raster = climate_full[[1]],  # Use the first layer of the climate data
                                    time_col = "time_bp",  # Specify the time column for matching climate data
                                    lubridate_fun = pastclim::ybp2date,  # Convert time_bp to date
                                    method = c("dist_min", km2m(150)))  # Ensure pseudo-absences are at least 150 km away

# Assign 'absence' class to pseudo-absences
pseudo_absences <- pseudo_absences %>%
  mutate(class = "absence")  # Add 'absence' class

# Combine presence and pseudo-absence data
conditor2 <- bind_rows(conditor, pseudo_absences)

# Plot conditor occurrence with the current land mask
ggplot() +
  geom_spatraster(data = land_mask_current) +  # Add the land mask raster
  scale_fill_terrain_c() +  # Use a terrain color scale
  geom_sf(data = conditor, aes(col = class))  # Plot conditor pseudo-absence and occurrence points, colored by class
#```

# Sample Pseudo-Absences and Plot conditor Occurrence with Land Mask
```{r}


# Sample pseudo-absences for conditor occurrence points between 300 and 90 years before present
conditor <- sample_pseudoabs_time(conditor %>% 
                                     filter(time_bp == -500), # Filter apicalis occurrences older than 20,000 years
                                   n_per_presence = 3,  # Generate 3 pseudo-absences per presence point
                                   raster = climate_full[[1]],  # Use the first layer of the climate data for sampling
                                   time_col = "time_bp",  # Specify the time column for matching climate data
                                   lubridate_fun = pastclim::ybp2date,  # Convert time_bp to date using pastclim function
                                   method = c("dist_min", km2m(150)))  # Ensure pseudo-absences are at least 150 km away




# Plot conditor occurrence with the current land mask
ggplot() +
  geom_spatraster(data = land_mask_current) +  # Add the land mask raster
  scale_fill_terrain_c() +  # Use a terrain color scale
  geom_sf(data = conditor2, aes(col = class))  # Plot conditor pseudo-absence and occurrence points, colored by class
```

# Retrieve Climate Data for This Region
```{r extract climate for region}
# Convert conditor to a data frame and extract spatial coordinates (longitude, latitude)
conditor_df <- conditor %>%
  dplyr::bind_cols(sf::st_coordinates(conditor)) %>%  # Bind spatial coordinates to the data
  mutate(time_bp = date2ybp(time_step)) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  select(-geometry)  # Remove geometry column

# Retrieve climate data for the conditor occurrence points from the climate series
conditor_df <- location_slice_from_region_series(conditor_df,
                                                 region_series = climate_full)  # Use the full climate series for extraction
```

# Add Climate Reconstructions to the sf Object
```{r add climate to sf}
# Add the climate reconstructions (bio05, bio06, bio12) to the conditor sf object and remove the time_step column
conditor <- conditor %>%
  bind_cols(conditor_df[, c("bio05", "bio06", "bio12")]) %>%  # Bind the climate data to the sf object
  select(-time_step) %>%  # Remove time_step column as it’s not needed for modeling
  filter(!is.na(.$bio05))  # Filter out any rows where bio05 is missing
```

# Set Up a Recipe for Modeling
```{r make recipe}
# Define a recipe for the model where 'class' is the outcome and other variables are predictors
# Geometry is automatically ignored as a predictor for sf objects
conditor_rec <- recipe(conditor, formula = class ~ .)  # Define the model formula
conditor_rec  # Check the recipe
```

# Build a Workflow Set for Model Tuning
```{r specify models}
# Create a workflow set of different models for tuning
# We will use glm, gam, random forest, and boosted trees as our models
conditor_models <- workflow_set(
  preproc = list(default = conditor_rec),  # Use the recipe for preprocessing
  models = list(
    glm = sdm_spec_glm(),  # Generalized Linear Model
    gam = sdm_spec_gam(),  # Generalized Additive Model
    rf = sdm_spec_rf(),    # Random Forest with tunable hyperparameters
    gbm = sdm_spec_boost_tree()  # Boosted Tree model with tunable hyperparameters
  ),
  cross = TRUE  # Cross all combinations of preprocessing and models
) %>%
  # Update the workflow model for GAMs with a formula specifying smooth terms for continuous predictors
  update_workflow_model("default_gam",
                        spec = sdm_spec_gam(),
                        formula = gam_formula(conditor_rec)) %>%
  # Adjust controls to store information needed for the ensemble model
  option_add(control = control_ensemble_grid())
```

# Spatial Block Cross-Validation for Model Tuning
```{r set up cross validation}
# Set up spatial block cross-validation with 5 folds for tuning and assessing models
set.seed(1005)  # Set seed for reproducibility
conditor_cv <- spatial_block_cv(conditor, v = 5)  # Create spatial block cross-validation with 5 folds
autoplot(conditor_cv)  # Plot the cross-validation folds
```

# Tune and Assess Models
```{r tune models}
# Use block cross-validation folds to tune and assess models
set.seed(123)  # Set seed for reproducibility
conditor_models <- conditor_models %>%
  workflow_map("tune_grid",  # Tune hyperparameters using grid search
               resamples = conditor_cv,  # Use the cross-validation resamples
               grid = 5,  # Use a grid size of 5 for tuning
               metrics = sdm_metric_set(),  # Use the default set of SDM metrics
               verbose = TRUE)  # Print verbose output
```

# Create an Ensemble Model
```{r ensemble}
# Create an ensemble by selecting the best set of parameters for each model
# The Boyce continuous index will be used as the metric for selecting the best random forest and boosted tree models
conditor_ensemble <- simple_ensemble() %>%
  add_member(conditor_models, metric = "boyce_cont")  # Add members to the ensemble using Boyce index as the metric

autoplot(conditor_ensemble)  # Plot the ensemble model performance
```

# Make Predictions for the Last Glacial Maximum (LGM)
```{r create LGM}
# Create a climate raster for the Last Glacial Maximum (LGM) using the correct layer (21) for 20,000 BP
climate_lgm <- rast(list(climate_full[[1]][[779]], climate_full[[2]][[779]], climate_full[[3]][[779]]))  # Extract 20,000 BP layers
names(climate_lgm) <- c("bio05", "bio06", "bio12")  # Rename the layers to match the climate variables


```
# Predict Using the Ensemble Model for LGM
```{r predict lgm}
# Use the ensemble model to predict species occurrence probabilities for the Last Glacial Maximum (LGM)
prediction_lgm <- predict_raster(conditor_ensemble, climate_lgm)  # Predict LGM probabilities

# Plot the predicted probabilities using ggplot2
ggplot() +
  geom_spatraster(data = prediction_lgm, aes(fill = mean)) +  # Add the prediction raster to the plot
  scale_fill_terrain_c(name = "Probability\nof occurrence")  # Use a terrain color scale to show probabilities
```

# Create a Stack of Predictions for All Time Steps
```{r stack o maps}
# Loop over 40 time steps (40ka in 1ka timesteps) to create and save a map of predictions for each time step
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present for each timestep
  
  # Create a climate snapshot for the current timestep
  climate_snapshot <- rast(list(climate_full[[1]][[timestep]], 
                                climate_full[[2]][[timestep]], 
                                climate_full[[3]][[timestep]]))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename layers to match the climate variables

  # Predict species occurrence for the current climate snapshot
  prediction1 <- predict_raster(conditor_ensemble, climate_snapshot)

  # Plot the predictions using ggplot2
  p <- ggplot() +
    geom_spatraster(data = prediction1, aes(fill = mean)) +  # Add the prediction raster
    scale_fill_terrain_c(name = "Probability\nof occurrence") +  # Use a terrain color scale
    ggtitle(paste0("Krapp2021 Model conditor Years BP: ", year_bp))  # Add a title with the year before present

  # Save the plot to a folder with filenames indicating the year BP
  ggsave(path= "C:/github/SNR_SDM/Results/conditorKrapp2021",
         plot = p,
         filename = paste0("conditor_map_", year_bp, ".png"))
})
```

# Partial Dependence Profiles and Variable Importance
## PDP for bio12
```{r PDP}
# Explain the ensemble model to generate partial dependence profiles and variable importance
explainer_conditor_ens <- explain_tidysdm(conditor_ensemble)  # Create an explainer object for the ensemble model

# Compute variable importance using the DALEX package
vip_ensemble <- model_parts(explainer = explainer_conditor_ens)  # Calculate variable importance
plot(vip_ensemble)  # Plot the variable importance results

# Generate a partial dependence profile for bio12
pdp_bio12 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio12")  # Generate PDP for bio12
plot(pdp_bio12)  # Plot the partial dependence profile for bio12
```

## PDP for bio05
```{r PDP bio05}
# Generate and plot a partial dependence profile for bio05
pdp_bio05 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio05")
plot(pdp_bio05)
```

## PDP for bio06
```{r PDP bio06}
# Generate and plot a partial dependence profile for bio06
pdp_bio06 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio06")
plot(pdp_bio06)
```