---
title: "Mastacomys_fuscus"
author: "Annie"
date: "2025-12-16"
output: html_document
---



```{r setup, include=FALSE}
# Set knitr defaults and seed for reproducibility
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

# Install packages (optional)

```{r install-packages, eval=FALSE}
# Convenience installer (prefer renv in real workflows)
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")     # devtools for GitHub/local installs

# Packages used below
packages <- c(
  "DALEX", "RefManageR", "USE", "broom", "car", "cowplot", "devtools", "dials",
  "dplyr", "ecospat", "fs", "ggplot2", "grid", "gtools", "here", "infer",
  "knitr", "magick", "modeldata", "oz", "parsnip", "pastclim", "patchwork",
  "purrr", "readr", "recipes", "rnaturalearth", "rnaturalearthdata", "rsample",
  "rworldmap", "rworldxtra", "scales", "sf", "spatialsample", "terra", "tibble",
  "tidymodels", "tidysdm", "tidyterra", "tidyverse", "tidyr", "tune", "viridis",
  "viridisLite", "workflowsets", "workflows", "yardstick", "ggnewscale"
)

# Install any missing packages (special cases for local/GitHub)
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "pastclim") {
      devtools::install_local("C:/github/pastclim")                                # local install if needed
    } else if (pkg == "USE") {
      devtools::install_github("danddr/USE")                                       # GitHub install
    } else {
      install.packages(pkg)                                                        # CRAN
    }
  }
}
```

# Libraries

```{r libraries}
suppressPackageStartupMessages({                                                  # keep console clean
  library(DALEX); library(RefManageR); library(USE)                               # explainers, refs, utils
  library(broom);  library(car);       library(cowplot)                           # model tidiers, diagnostics, plotting
  library(dials);  library(dplyr);     library(ecospat)                           # tuning, wrangling, ecology
  library(fs);     library(ggplot2);   library(grid)                              # files, plotting, grid
  library(gtools); library(here);      library(infer)                             # helpers, project paths, stats
  library(knitr);  library(magick);    library(modeldata)                         # knit, image ops, example data
  library(oz);     library(parsnip);   library(pastclim)                          # AU map helpers, model specs, palaeoclimate
  library(patchwork); library(purrr);  library(readr)                             # compose plots, mapping, fast IO
  library(recipes); library(rnaturalearth); library(rnaturalearthdata)            # preprocessing, basemaps
  library(rsample); library(rworldmap); library(rworldxtra)                       # resampling, basemaps
  library(scales);  library(sf);       library(spatialsample)                     # scales, SF, spatial CV
  library(terra);   library(tibble);   library(tidymodels)                         # raster, tibbles, ML framework
  library(tidysdm); library(tidyterra); library(tidyverse)                        # SDM tools, terra geoms, tidyverse
  library(tidyr);   library(tune);     library(viridis)                           # reshape, tuning, palettes
  library(viridisLite); library(workflowsets); library(workflows)                 # palettes, model sets, workflows
  library(yardstick)  ; library(ggnewscale)                                                            # metrics
})
```

# Project paths

```{r paths}
here::i_am("Analyses/Mastacomys_fuscus.Rmd")                               # declare project root via here()

results_dir   <- here("Results/stratified environmental sampling", "Mastacomys_fuscus")     # figures/tables output
processed_dir <- here("Data", "processed")                                         # cleaned input data
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)            # ensure results folder exists
```

# Globals (shared objects)

```{r globals}
world_map <- rworldmap::getMap(resolution = "high")                                 # coarse world polygons
australia_outline <- sf::st_as_sf(world_map) |> dplyr::filter(ADMIN == "Australia") # Australia boundary as sf

Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)                                  # study extent (lon/lat)
xlim <- c(112, 150); ylim <- c(-45, -10)                                            # plotting window

theme_map_clean <- theme_minimal(base_size = 12) +                                  # reusable map theme
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank()
  )
```

# Data: download Krapp2021

```{r download-climate}
pastclim::download_dataset("Krapp2021")                                             # make sure climate dataset is available
```

# Read Mastacomys_fuscus occurrences

```{r read-Mastacomys_fuscus}
Mastacomys_fuscus <- readr::read_csv(file.path(processed_dir, "Mastacomys_fuscus.csv"),         # load occurrences
                            show_col_types = FALSE) |>
  (\(x) { print(x); x })() |>                                                      # echo a preview for sanity
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)                    # convert to sf (WGS84)
```

# Land mask (0 ka)

```{r land-mask}
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") |> # 0 ka land/sea mask
  terra::crop(terra::vect(Aust_extent)) |>                                          # crop to Australia extent
  terra::aggregate(fact = 5, fun = "mean", na.rm = TRUE)                            # coarsen for quick plotting
```

# Quick map of occurrences

```{r plot-occurrences}
ggplot() +
  tidyterra::geom_spatraster(data = land_mask_current, aes()) +                     # draw land mask
  tidyterra::scale_fill_terrain_c() +                                               # simple palette
  geom_sf(data = Mastacomys_fuscus, aes(col = time_bp)) +                                    # plot points coloured by time
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)                                                # map window
```

# Climate time series (Krapp2021)

```{r climate-series}
climate_vars <- c("bio05","bio06","bio12")                                         # chosen bioclim variables
climate_full <- pastclim::region_series(                                            # time series for region
  bio_variables = climate_vars,
  data = "Krapp2021",
  crop = terra::vect(Aust_extent)
)
```

# Spatial filtering

```{r spatial-filter}
set.seed(123)                                                                       # reproducible thinning
Mastacomys_fuscus <- thin_by_dist(Mastacomys_fuscus, dist_min = km2m(50))                             # remove points <50 km apart
NROW(Mastacomys_fuscus)                                                                       # report retained count
```
#Grid-resolution-optimisation
NOTE: This is slow but only needs to be run once.
 ```{r grid-resolution-optimisation}
# 1) Extract environmental values from the raster
env_values <- terra::as.data.frame(env_space_1950, xy = TRUE, na.rm = TRUE)

# 2) Perform PCA on the environmental variables
#    (Check that names below match env_values)
pca_res <- prcomp(
  env_values[, c("bio05_0", "bio06_0", "bio12_0")],
  scale. = TRUE
)

# 3) Create an sf object with first two PC scores as coordinates
env_pca_sf <- sf::st_as_sf(
  data.frame(
    PC1 = pca_res$x[, 1],
    PC2 = pca_res$x[, 2],
    x   = env_values$x,
    y   = env_values$y
  ),
  coords = c("PC1", "PC2")
)

# 4) Grid-resolution optimisation
opt_res <- optimRes(
  env_pca_sf,
  grid.res = seq(1, 100, by = 1),
  perc.thr = 10,
  cr       = 1,
  showOpt  = TRUE
)

optimal_grid_res <- opt_res[[2]]
print(optimal_grid_res)


 ```
# Pseudo-absence sampling in environmental space

```{r env-space-setup}
# 1950 slice (index 800 ≈ near-present in Krapp2021 time axis)
env_space_1950 <- c(climate_full[[1]][[800]],                                      # bio05 layer at ~0 BP
                    climate_full[[2]][[800]],                                      # bio06 layer
                    climate_full[[3]][[800]])                                      # bio12 layer
```

```{r pseudo-abs-env}
library(scales)                                                                     # for alpha()
Mastacomys_fuscus3 <- paSampling(                                                            # sample pseudo-absences
  env.rast = env_space_1950,                                                        # env raster stack at ~present
  pres     = Mastacomys_fuscus %>% dplyr::filter(time_bp == 1),                           # use ~1950 AD presences
  thres    = 0.75,                                                                  # env distance threshold
  n.tr     = 9,                                                                     # trials per grid cell
  grid.res = 9                                                                      # grid resolution for sampling
) |>
  sf::st_drop_geometry()                                                            # drop geometry for data.frame ops

# Quick visual check of pseudo-absences vs presences
pseudoabsence_map <- ggplot() +
  geom_sf(data = australia_outline, fill = alpha("#C4C4C4", 0.5), color = "grey50") +
  geom_point(data = Mastacomys_fuscus3, aes(x, y, shape = "Pseudo Absence", fill = "Pseudo Absence"),
             color = "black", size = 2.5, stroke = 0.3) +
  geom_sf(data = Mastacomys_fuscus |> dplyr::filter(time_bp == -300),
          aes(fill = "Presence", shape = "Presence"),
          color = "black", size = 2.5, stroke = 0.3) +
  scale_fill_manual(values = c("Presence" = "#A100FA", "Pseudo Absence" = "#FF7E1D"), name = "Legend") +
  scale_shape_manual(values = c("Presence" = 24, "Pseudo Absence" = 21), name = "Legend") +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +
  scale_x_continuous(breaks = seq(110, 155, by = 5)) +
  scale_y_continuous(breaks = seq(-45, -10, by = 5)) +
  labs(title = "Pseudo Absences in Environmental Space", x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "#C4C4C4", linewidth = 0.3),
    panel.grid.minor = element_line(color = "#C4C4C4", linewidth = 0.2),
    panel.border     = element_rect(color = "#C4C4C4", fill = NA, linewidth = 0.3)
  )

ggsave(file.path(results_dir, "pseudo_absences_envspace.png"), pseudoabsence_map,  # save figure
       width = 6, height = 5, dpi = 300, bg = "white")
```

# Extract climate for points

```{r extract-climate}
Mastacomys_fuscus_df <- Mastacomys_fuscus |>                                                          # start with sf presences
  dplyr::bind_cols(sf::st_coordinates(Mastacomys_fuscus)) |>                                 # add lon/lat columns
  dplyr::rename(x = X, y = Y) |>                                                    # rename to x/y
  dplyr::mutate(percP = "pres") |>                                                  # label as presence
  dplyr::bind_rows(Mastacomys_fuscus3) |>                                                    # append pseudo-absences
  dplyr::mutate(time_bp = 1) |>                                                  # time slice to extract
  as.data.frame() |>                                                                
  dplyr::select(-geometry)                                                          # drop geometry

Mastacomys_fuscus_df <- location_slice_from_region_series(                                   # extract env time slice
  Mastacomys_fuscus_df,
  region_series = climate_full
)
```

# Add climate back and cast to sf

```{r add-climate-sf}
Mastacomys_fuscus4 <- Mastacomys_fuscus |>                                                            # rebuild sf with env vars
  dplyr::bind_cols(sf::st_coordinates(Mastacomys_fuscus)) |>
  dplyr::rename(x = X, y = Y) |>
  dplyr::mutate(percP = "pres") |>
  dplyr::bind_rows(Mastacomys_fuscus3) |>
  dplyr::mutate(time_bp = 1) |>
  as.data.frame() |>
  dplyr::select(-geometry) |>
  dplyr::bind_cols(Mastacomys_fuscus_df[, c("bio05", "bio06", "bio12")]) |>                 # attach env predictors
  dplyr::filter(!is.na(.data$bio05)) |>
  dplyr::mutate(percP = dplyr::if_else(percP == "pabs", "pseudoabsence", "presence")) |>
  sf::st_as_sf(coords = c("x", "y"), crs = 4326)                                   # back to sf
```

# Recipe

```{r recipe}
Mastacomys_fuscus_rec <- recipes::recipe(Mastacomys_fuscus4, formula = percP ~ bio05 + bio06 + bio12) # model formula
Mastacomys_fuscus_rec                                                                        # print recipe summary
```

# Models & workflows

```{r models}
Mastacomys_fuscus_models <- workflowsets::workflow_set(                                      # define model set
  preproc = list(default = Mastacomys_fuscus_rec),
  models  = list(
    glm = sdm_spec_glm(),                                                           # GLM
    gam = sdm_spec_gam(),                                                           # GAM
    rf  = sdm_spec_rf(),                                                            # Random Forest
    gbm = sdm_spec_boost_tree()                                                     # Gradient Boosting
  ),
  cross = TRUE
) |>
  update_workflow_model(                                                            # set GAM smooths
    "default_gam",
    spec = sdm_spec_gam(),
    formula = percP ~ s(bio05, k = 10) + s(bio06, k = 10) + s(bio12, k = 10)
  ) |>
  option_add(control = control_ensemble_grid())                                     # tuning control
```

# Spatial block CV

```{r cv}
set.seed(1005)                                                                      # CV reproducibility
Mastacomys_fuscus_cv <- spatial_block_cv(Mastacomys_fuscus4, v = 5)                                   # spatial CV folds
autoplot(Mastacomys_fuscus_cv)                                                               # quick diagnostics plot
```

# Tune

```{r tune}
set.seed(123)                                                                       # tuning reproducibility
Mastacomys_fuscus_models <- Mastacomys_fuscus_models |>
  workflow_map("tune_grid",
               resamples = Mastacomys_fuscus_cv,                                             # use spatial CV
               grid = 5,                                                            # grid size
               metrics = sdm_metric_set(),                                          # SDM metrics incl. TSS
               verbose = TRUE)
```

# Build ensemble & inspect choices

```{r ensemble}
Mastacomys_fuscus_ensemble <- simple_ensemble() |>                                           # start ensemble
  add_member(Mastacomys_fuscus_models, metric = "tss_max")                                   # pick best by TSS

# Summarise chosen hyperparameters and TSS per workflow
chosen <- Mastacomys_fuscus_models |>
  dplyr::mutate(
    best = purrr::map(result, ~ tune::select_best(.x, metric = "tss_max")),
    tss  = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "tss_max", n = 1)$mean),
    best = purrr::map_chr(best, ~ paste(names(.x), .x, sep = "=", collapse = "; "))
  ) |>
  dplyr::transmute(wflow_id, best, tss)
print(chosen)

# Fold-level performance boxplots
workflow_labels <- c(
  default_gam = "Generalised Additive Model",
  default_gbm = "Boosted Tree",
  default_glm = "Generalised Linear Model",
  default_rf  = "Random Forest"
)
model_order <- c("Generalised Additive Model", "Boosted Tree", "Generalised Linear Model", "Random Forest")
metric_labels <- c(boyce_cont = "Boyce Index", roc_auc = "AUC", tss_max = "TSS")
color_map <- c(
  "Generalised Additive Model" = "#FF61C3",
  "Boosted Tree"               = "#00B140",
  "Generalised Linear Model"   = "#00BFFF",
  "Random Forest"              = "#BC43FF"
)

# Collect fold metrics and draw
perf_data <- Mastacomys_fuscus_models |>
  dplyr::mutate(metrics = purrr::map(result, collect_metrics, summarize = FALSE)) |>
  dplyr::select(wflow_id, metrics) |>
  tidyr::unnest(metrics) |>
  dplyr::filter(.metric %in% names(metric_labels)) |>
  dplyr::mutate(
    model        = factor(workflow_labels[wflow_id], levels = model_order),
    metric_label = factor(metric_labels[.metric], levels = c("Boyce Index", "AUC", "TSS"))
  )

perf_plot <- ggplot(perf_data, aes(x = model, y = .estimate, fill = model)) +
  geom_boxplot(width = 0.6, alpha = 0.9, outlier.shape = NA) +
  geom_jitter(aes(color = model), size = 1.3, width = 0.15, alpha = 0.4, stroke = 0.3, shape = 21) +
  scale_fill_manual(values = color_map, name = "Model Type") +
  scale_color_manual(values = rep("black", length(model_order)), guide = "none") +
  facet_wrap(~ metric_label, scales = "free_y") +
  labs(title = "Model performance across algorithms (fold-level)", x = NULL, y = "Performance metric") +
  theme_minimal(base_size = 12) +
  theme(
    panel.background  = element_rect(fill = "#EEEEEE", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.grid.major  = element_line(color = "white"),
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_blank(),
    axis.ticks.x      = element_blank(),
    legend.position   = "right",
    legend.direction  = "vertical",
    legend.title      = element_text(size = 11),
    legend.text       = element_text(size = 10),
    plot.title        = element_text(face = "bold", size = 13),
    strip.text        = element_text(margin = margin(b = 5))
  )

ggsave(file.path(results_dir, "ensemble", "ensemble_performance_clean.png"), perf_plot,
       width = 8, height = 5, dpi = 300)                                           # save figure
perf_plot                                                                            # print plot

# PRINT STATISTICAL SUMMARY FOR EACH MODEL TYPE
# ─────────────────────────────────────────────────────────────

# Extract mean AUC and Boyce as well as TSS (already computed above)
stats_summary <- Mastacomys_fuscus_models |>
  dplyr::mutate(
    auc    = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "roc_auc",  n = 1)$mean),
    boyce  = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "boyce_cont", n = 1)$mean),
    tss    = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "tss_max",  n = 1)$mean)
  ) |>
  dplyr::transmute(
    Model = workflow_labels[wflow_id],
    AUC   = round(auc,   3),
    Boyce = round(boyce, 3),
    TSS   = round(tss,   3)
  )

cat("\n\n===== Ensemble Model Performance Summary =====\n")
print(stats_summary)
cat("================================================\n\n")
```
```{r ensemble}
# ---- Build ensemble object (placeholder) ----
# We'll fill this with the best-ranked workflow below
# (simple 'best-of-set' ensemble)
# NOTE: Mastacomys_fuscus_models is your workflow_set with tuned results

# Summarise chosen hyperparameters and TSS per workflow (as before)
Mastacomys_fuscus_ensemble_dummy <- simple_ensemble()                                    # keep for interface

chosen <- Mastacomys_fuscus_models |>
  dplyr::mutate(
    best = purrr::map(result, ~ tune::select_best(.x, metric = "tss_max")),
    tss  = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "tss_max", n = 1)$mean),
    best = purrr::map_chr(best, ~ paste(names(.x), .x, sep = "=", collapse = "; "))
  ) |>
  dplyr::transmute(wflow_id, best, tss)
print(chosen)

# ---- Multi-metric summary and ranking (AUC, Boyce, TSS) ----
metrics_summary <- Mastacomys_fuscus_models |>
  dplyr::mutate(
    auc   = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "roc_auc",    n = 1)$mean),
    boyce = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "boyce_cont", n = 1)$mean),
    tss   = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "tss_max",    n = 1)$mean)
  ) |>
  dplyr::transmute(
    wflow_id,
    Model = workflow_labels[wflow_id],
    AUC   = auc,
    Boyce = boyce,
    TSS   = tss
  ) |>
  dplyr::mutate(
    rank_auc   = dplyr::min_rank(dplyr::desc(AUC)),
    rank_boyce = dplyr::min_rank(dplyr::desc(Boyce)),
    rank_tss   = dplyr::min_rank(dplyr::desc(TSS)),
    rank_sum   = rank_auc + rank_boyce + rank_tss
  )

cat("\n\n===== Multi-metric ranking (AUC, Boyce, TSS) =====\n")
print(
  metrics_summary |>
    dplyr::arrange(rank_sum) |>
    dplyr::select(Model, AUC, Boyce, TSS, rank_auc, rank_boyce, rank_tss, rank_sum)
)
cat("===================================================\n\n")

# ---- Select best workflow based on combined rank ----
best_wflow_id <- metrics_summary$wflow_id[which.min(metrics_summary$rank_sum)]
cat("Best workflow (multi-metric rank):", best_wflow_id, "\n")

best_model <- Mastacomys_fuscus_models |>
  dplyr::filter(wflow_id == best_wflow_id)

# ---- Build 'ensemble' from the best-ranked workflow ----
Mastacomys_fuscus_ensemble <- simple_ensemble() |>
  add_member(best_model, metric = "tss_max")    # metric irrelevant here; only one member

# ---- Fold-level performance boxplots (unchanged) ----
workflow_labels <- c(
  default_gam = "Generalised Additive Model",
  default_gbm = "Boosted Tree",
  default_glm = "Generalised Linear Model",
  default_rf  = "Random Forest"
)
model_order <- c("Generalised Linear Model", "Generalised Additive Model",
                 "Random Forest", "Boosted Tree")
metric_labels <- c(boyce_cont = "Boyce Index", roc_auc = "AUC", tss_max = "TSS")
color_map <- c(
  "Generalised Additive Model" = "#FF61C3",
  "Boosted Tree"               = "#00B140",
  "Generalised Linear Model"   = "#00BFFF",
  "Random Forest"              = "#BC43FF"
)

perf_data <- Mastacomys_fuscus_models |>
  dplyr::mutate(metrics = purrr::map(result, collect_metrics, summarize = FALSE)) |>
  dplyr::select(wflow_id, metrics) |>
  tidyr::unnest(metrics) |>
  dplyr::filter(.metric %in% names(metric_labels)) |>
  dplyr::mutate(
    model        = factor(workflow_labels[wflow_id], levels = model_order),
    metric_label = factor(metric_labels[.metric], levels = c("Boyce Index", "AUC", "TSS"))
  )

perf_plot <- ggplot(perf_data, aes(x = model, y = .estimate, fill = model)) +
  geom_boxplot(width = 0.6, alpha = 0.9, outlier.shape = NA) +
  geom_jitter(aes(color = model), size = 1.3, width = 0.15, alpha = 0.4, stroke = 0.3, shape = 21) +
  scale_fill_manual(values = color_map, name = "Model Type") +
  scale_color_manual(values = rep("black", length(model_order)), guide = "none") +
  facet_wrap(~ metric_label, scales = "free_y") +
  labs(title = "Model performance across algorithms (fold-level)", x = NULL, y = "Performance metric") +
  theme_minimal(base_size = 12) +
  theme(
    panel.background  = element_rect(fill = "#EEEEEE", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.grid.major  = element_line(color = "white"),
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_blank(),
    axis.ticks.x      = element_blank(),
    legend.position   = "right",
    legend.direction  = "vertical",
    legend.title      = element_text(size = 11),
    legend.text       = element_text(size = 10),
    plot.title        = element_text(face = "bold", size = 13),
    strip.text        = element_text(margin = margin(b = 5))
  )

ggsave(file.path(results_dir, "ensemble", "ensemble_performance_clean.png"),
       perf_plot, width = 8, height = 5, dpi = 300)

perf_plot
```
# LGM snapshot

```{r lgm}
climate_lgm <- terra::rast(list(climate_full[[1]][[779]],                           # build 3-band env raster
                                climate_full[[2]][[779]],
                                climate_full[[3]][[779]]))
names(climate_lgm) <- c("bio05", "bio06", "bio12")                                  # name bands

prediction_lgm <- predict_raster(Mastacomys_fuscus_ensemble, climate_lgm)                    # project ensemble

ggplot() +
  tidyterra::geom_spatraster(data = prediction_lgm, aes(fill = mean)) +
  tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE)
```

# Save model as SpatRaster stack

```{r save-stack}
Mastacomys_fuscus_spatrast <- NULL                                                           # init empty stack
purrr::walk(760:800, function(timestep) {                                           # iterate timesteps
  year_bp <- time(climate_full[[1]])[timestep] - 1950                               # convert to years BP
  climate_snapshot <- terra::rast(list(                                             # assemble env snapshot
    climate_full[[1]][[timestep]],
    climate_full[[2]][[timestep]],
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")
  prediction_raster <- predict_raster(Mastacomys_fuscus_ensemble, climate_snapshot)          # predict suitability
  Mastacomys_fuscus_spatrast <<- if (is.null(Mastacomys_fuscus_spatrast)) prediction_raster           # stack predictions
                      else c(Mastacomys_fuscus_spatrast, prediction_raster)
})

time(Mastacomys_fuscus_spatrast) <- time(climate_full[[1]])[760:800] - 1950                  # attach BP times

output_dir <- here("Results", "Mastacomys_fuscus Krapp model")                               # output folder
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
writeRaster(Mastacomys_fuscus_spatrast, filename = file.path(output_dir, "predictions_stack.tif"),
            overwrite = TRUE)                                                       # write GeoTIFF
```

# Stack o' maps

```{r stack-o-maps, dependson="globals"}
suppressPackageStartupMessages({ library(magick) })                                 # image post-processing
stopifnot(exists("australia_outline"), exists("results_dir"),                       # guardrails
          exists("climate_full"), exists("Mastacomys_fuscus_ensemble"))

out_dir <- file.path(results_dir, "model"); if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

purrr::walk(760:800, function(timestep) {                                           # loop through timesteps
  year_bp <- time(climate_full[[1]])[timestep] - 1950
  climate_snapshot <- terra::rast(list(
    climate_full[[1]][[timestep]],
    climate_full[[2]][[timestep]],
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")
  prediction_raster <- predict_raster(Mastacomys_fuscus_ensemble, climate_snapshot)

  p <- ggplot() +                                                                   # draw map
    tidyterra::geom_spatraster(data = prediction_raster, aes(fill = mean)) +
    tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +
    coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
    ggtitle(bquote(italic("Leporillus Mastacomys_fuscus") * " — Years BP: " * .(year_bp))) +
    theme_minimal() +
    theme(
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#F2F2F2", color = NA),
      panel.grid.major = element_line(color = "#C4C4C4"),
      panel.grid.minor = element_line(color = "#C4C4C4"),
      plot.margin      = margin(0, 0, 0, 0, unit = "pt")
    )

  plot_path <- file.path(out_dir, paste0("Mastacomys_fuscus_map_", year_bp, ".png"))         # save path
  ggsave(plot_path, p, width = 7, height = 7, dpi = 300, bg = "transparent")       # save plot

  img <- image_read(plot_path)                                                      # trim/border
  img <- image_border(image_trim(img), color = "white", geometry = "30x30")
  image_write(img, path = plot_path)
})
```

# 6-image panel (model only)

```{r six-panel}
raster_times  <- terra::time(Mastacomys_fuscus_spatrast)                                     # all BP times
target_times  <- c(-21000, -17000, -15000, -11000, -6000, 0)                        # selected panels
matched_idx   <- match(target_times, raster_times)                                   # index match
panel_tags    <- letters[seq_along(target_times)]                                    # (a)-(f)

panel_plots <- pmap(
  list(idx = matched_idx, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- Mastacomys_fuscus_spatrast[[idx]]                                  # pick raster layer
    ggplot() +
      tidyterra::geom_spatraster(data = current_prediction, aes(fill = mean)) +
      tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +
      coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
      scale_y_continuous(breaks = seq(-45, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(year_bp))) +
      theme_minimal(base_size = 10) +
      theme(
        panel.background = element_rect(fill = "#DCF3FC", color = NA),
        plot.background  = element_rect(fill = "white", color = NA),
        panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_blank(),
        axis.text        = element_text(size = 8),
        axis.title       = element_blank(),
        plot.margin      = margin(2, 2, 2, 2),
        legend.position  = "right",
        legend.title     = element_text(size = 11),
        legend.text      = element_text(size = 11)
      )
  }
)

final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +              # combine panels
  plot_annotation(title = expression(italic("Leporillus Mastacomys_fuscus") * " SDM model predictions"),
                  theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)))

ggsave(file.path(results_dir, "Mastacomys_fuscus_model_panel.png"), final_panel,             # save figure
       width = 10, height = 9, dpi = 300)
```




# Variable importance table + PDPs

```{r vip-pdp}
suppressPackageStartupMessages({
  library(DALEX); library(tidysdm); library(ggplot2); library(dplyr); library(knitr)
})                                                                              # quiet attaches

# ---- Build explainer & output dir ----
Mastacomys_fuscus_explainer <- explain_tidysdm(Mastacomys_fuscus_ensemble)                        # DALEX explainer
Mastacomys_fuscus_outdir <- file.path(results_dir, "variable_importance")
dir.create(Mastacomys_fuscus_outdir, showWarnings = FALSE, recursive = TRUE)

# ---- Aggregate permutation importance into a neat table ----
Mastacomys_fuscus_vip <- model_parts(explainer = Mastacomys_fuscus_explainer) |>
  dplyr::filter(!variable %in% c("_baseline_", "_full_model_")) |>
  dplyr::group_by(variable) |>
  dplyr::summarise(
    `Mean drop-out loss` = round(mean(dropout_loss, na.rm = TRUE), 3),
    `SD`                 = round(sd(dropout_loss,   na.rm = TRUE), 3),
    .groups = "drop_last"
  ) |>
  dplyr::arrange(desc(`Mean drop-out loss`)) |>
  dplyr::rename(`Environmental variable` = variable)

print(Mastacomys_fuscus_vip)                                                             # print in console
write.csv(Mastacomys_fuscus_vip, file.path(Mastacomys_fuscus_outdir, "Mastacomys_fuscus_variable_importance_table.csv"),
          row.names = FALSE)                                                    # save CSV
knitr::kable(Mastacomys_fuscus_vip,
             caption = "Variable importance summary (Leporillus Mastacomys_fuscus ensemble model)",
             format = "markdown")                                               # show as table in Rmd

# ---- Helper to compute & save PDPs for a single variable ----
plot_and_save_pdp <- function(var, label){
  pdp <- model_profile(Mastacomys_fuscus_explainer, N = 500, variables = var)           # partial dependence
  p <- ggplot(pdp$agr_profiles, aes(x = `_x_`, y = `_yhat_`)) +
    geom_line(linewidth = 1) +
    labs(title = paste("Leporillus Mastacomys_fuscus – Partial dependence on", label),
         x = label, y = "Predicted suitability") +
    theme_minimal(base_size = 12) +
    theme(plot.background  = element_rect(fill = "white", color = NA),
          panel.background = element_rect(fill = "#F2F2F2", color = NA),
          panel.grid.major = element_line(color = "#C4C4C4", linewidth = 0.3),
          panel.border     = element_rect(color = "#C4C4C4", fill = NA, linewidth = 0.3),
          plot.title       = element_text(face = "italic", size = 13))
  ggsave(file.path(Mastacomys_fuscus_outdir, paste0("Mastacomys_fuscus_pdp_", var, ".png")),
         p, width = 6, height = 4, dpi = 300, bg = "white")
  print(p)
}

# ---- Generate PDPs ----
plot_and_save_pdp("bio05", "Maximum temperature of warmest month (°C)")
plot_and_save_pdp("bio06", "Minimum temperature of coldest month (°C)")
plot_and_save_pdp("bio12", "Annual precipitation (mm)")
```


