---
title: "Apicalis cleaned"
author: "Annie"
date: "2025-10-30"
output: html_document
---

```{r setup, include=FALSE}
# Set knitr defaults and seed for reproducibility
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(42)
```

# Install packages (optional)

```{r install-packages, eval=FALSE}
# Convenience installer (prefer renv in real workflows)
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")     # devtools for GitHub/local installs

# Packages used below
packages <- c(
  "DALEX", "RefManageR", "USE", "broom", "car", "cowplot", "devtools", "dials",
  "dplyr", "ecospat", "fs", "ggplot2", "grid", "gtools", "here", "infer",
  "knitr", "magick", "modeldata", "oz", "parsnip", "pastclim", "patchwork",
  "purrr", "readr", "recipes", "rnaturalearth", "rnaturalearthdata", "rsample",
  "rworldmap", "rworldxtra", "scales", "sf", "spatialsample", "terra", "tibble",
  "tidymodels", "tidysdm", "tidyterra", "tidyverse", "tidyr", "tune", "viridis",
  "viridisLite", "workflowsets", "workflows", "yardstick", "ggnewscale"
)

# Install any missing packages (special cases for local/GitHub)
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "pastclim") {
      devtools::install_local("C:/github/pastclim")                                # local install if needed
    } else if (pkg == "USE") {
      devtools::install_github("danddr/USE")                                       # GitHub install
    } else {
      install.packages(pkg)                                                        # CRAN
    }
  }
}
```

# Libraries

```{r libraries}
suppressPackageStartupMessages({                                                  # keep console clean
  library(DALEX); library(RefManageR); library(USE)                               # explainers, refs, utils
  library(broom);  library(car);       library(cowplot)                           # model tidiers, diagnostics, plotting
  library(dials);  library(dplyr);     library(ecospat)                           # tuning, wrangling, ecology
  library(fs);     library(ggplot2);   library(grid)                              # files, plotting, grid
  library(gtools); library(here);      library(infer)                             # helpers, project paths, stats
  library(knitr);  library(magick);    library(modeldata)                         # knit, image ops, example data
  library(oz);     library(parsnip);   library(pastclim)                          # AU map helpers, model specs, palaeoclimate
  library(patchwork); library(purrr);  library(readr)                             # compose plots, mapping, fast IO
  library(recipes); library(rnaturalearth); library(rnaturalearthdata)            # preprocessing, basemaps
  library(rsample); library(rworldmap); library(rworldxtra)                       # resampling, basemaps
  library(scales);  library(sf);       library(spatialsample)                     # scales, SF, spatial CV
  library(terra);   library(tibble);   library(tidymodels)                         # raster, tibbles, ML framework
  library(tidysdm); library(tidyterra); library(tidyverse)                        # SDM tools, terra geoms, tidyverse
  library(tidyr);   library(tune);     library(viridis)                           # reshape, tuning, palettes
  library(viridisLite); library(workflowsets); library(workflows)                 # palettes, model sets, workflows
  library(yardstick)  ; library(ggnewscale)                                                            # metrics
})
```

# Project paths

```{r paths}
here::i_am("Analyses/Krapp apicalis cleaned.Rmd")                               # declare project root via here()

results_dir   <- here("Results/stratified environmental sampling", "apicalis")     # figures/tables output
processed_dir <- here("Data", "processed")                                         # cleaned input data
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)            # ensure results folder exists
```

# Globals (shared objects)

```{r globals}

# --- Load outline from rworldmap ---
world_map <- rworldmap::getMap(resolution = "high")

australia_outline_raw <-
  sf::st_as_sf(world_map) |>
  dplyr::filter(ADMIN == "Australia")

# --- Study extent (lon/lat) ---
Aust_extent <- terra::ext(110, 155, -45, -10)

xlim <- c(110, 155)
ylim <- c(-45, -10)

# --- Crop outline so ggplot can't expand to sub-Antarctic polygons ---
australia_outline <-
  sf::st_crop(
    australia_outline_raw,
    xmin = xlim[1], xmax = xlim[2],
    ymin = ylim[1], ymax = ylim[2]
  )

# --- Clean map theme ---
theme_map_clean <- theme_minimal(base_size = 12) +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank()
  )


```

# Data: download Krapp2021

```{r download-climate}
pastclim::download_dataset("Krapp2021")                                             # make sure climate dataset is available
```

# Read apicalis occurrences

```{r read-apicalis}
apicalis <- readr::read_csv(file.path(processed_dir, "apicalis_flat.csv"),         # load occurrences
                            show_col_types = FALSE) |>
  (\(x) { print(x); x })() |>                                                      # echo a preview for sanity
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)                    # convert to sf (WGS84)
```

# Globals (shared objects)

```{r globals}
world_map <- rworldmap::getMap(resolution = "high")                                 # coarse world polygons
australia_outline <- sf::st_as_sf(world_map) |> dplyr::filter(ADMIN == "Australia") # Australia boundary as sf

Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)                                  # study extent (lon/lat)
xlim <- c(112, 150); ylim <- c(-40, -10)                                            # plotting window

theme_map_clean <- theme_minimal(base_size = 12) +                                  # reusable map theme
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank()
  )
```

# Data: download Krapp2021

```{r download-climate}
pastclim::download_dataset("Krapp2021")                                             # make sure climate dataset is available
```

# Read *apicalis* occurrences

```{r read-apicalis}
apicalis <- readr::read_csv(file.path(processed_dir, "apicalis_flat.csv"),         # load occurrences
                            show_col_types = FALSE) |>
  (\(x) { print(x); x })() |>                                                      # echo a preview for sanity
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)                    # convert to sf (WGS84)
```

# Land mask (0 ka)

```{r land-mask}
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") |> # 0 ka land/sea mask
  terra::crop(terra::vect(Aust_extent)) |>                                          # crop to Australia extent
  terra::aggregate(fact = 5, fun = "mean", na.rm = TRUE)                            # coarsen for quick plotting
```

# Quick map of occurrences

```{r plot-occurrences}
ggplot() +
  tidyterra::geom_spatraster(data = land_mask_current, aes()) +                     # draw land mask
  tidyterra::scale_fill_terrain_c() +                                               # simple palette
  geom_sf(data = apicalis, aes(col = time_bp)) +                                    # plot points coloured by time
  coord_sf(xlim = xlim, ylim = ylim)                                                # map window
```

# Climate time series (Krapp2021)

```{r climate-series}
climate_vars <- c("bio05","bio06","bio12")                                         # chosen bioclim variables
climate_full <- pastclim::region_series(                                            # time series for region
  bio_variables = climate_vars,
  data = "Krapp2021",
  crop = terra::vect(Aust_extent)
)
```

# Spatial filtering

```{r spatial-filter}
set.seed(123)                                                                       # reproducible thinning
apicalis <- thin_by_dist(apicalis, dist_min = km2m(50))                             # remove points <50 km apart
NROW(apicalis)                                                                       # report retained count
```

# Pseudo-absence sampling in environmental space

```{r env-space-setup}
# 1950 slice (index 800 ≈ near-present in Krapp2021 time axis)
env_space_1950 <- c(climate_full[[1]][[800]],                                      # bio05 layer at ~0 BP
                    climate_full[[2]][[800]],                                      # bio06 layer
                    climate_full[[3]][[800]])                                      # bio12 layer
```

```{r pseudo-abs-env}
library(scales)                                                                     # for alpha()
apicalis3 <- paSampling(                                                            # sample pseudo-absences
  env.rast = env_space_1950,                                                        # env raster stack at ~present
  pres     = apicalis %>% dplyr::filter(time_bp == -300),                           # use ~1950 AD presences
  thres    = 0.75,                                                                  # env distance threshold
  n.tr     = 9,                                                                     # trials per grid cell
  grid.res = 9                                                                      # grid resolution for sampling
) |>
  sf::st_drop_geometry()                                                            # drop geometry for data.frame ops

# Quick visual check of pseudo-absences vs presences
pseudoabsence_map <- ggplot() +
  geom_sf(data = australia_outline, fill = alpha("#C4C4C4", 0.5), color = "grey50") +
  geom_point(data = apicalis3, aes(x, y, shape = "Pseudo Absence", fill = "Pseudo Absence"),
             color = "black", size = 2.5, stroke = 0.3) +
  geom_sf(data = apicalis |> dplyr::filter(time_bp == -300),
          aes(fill = "Presence", shape = "Presence"),
          color = "black", size = 2.5, stroke = 0.3) +
  scale_fill_manual(values = c("Presence" = "#A100FA", "Pseudo Absence" = "#FF7E1D"), name = "Legend") +
  scale_shape_manual(values = c("Presence" = 24, "Pseudo Absence" = 21), name = "Legend") +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +
  scale_x_continuous(breaks = seq(110, 155, by = 5)) +
  scale_y_continuous(breaks = seq(-45, -10, by = 5)) +
  labs(title = "Pseudo Absences in Environmental Space", x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "#C4C4C4", linewidth = 0.3),
    panel.grid.minor = element_line(color = "#C4C4C4", linewidth = 0.2),
    panel.border     = element_rect(color = "#C4C4C4", fill = NA, linewidth = 0.3)
  )

ggsave(file.path(results_dir, "pseudo_absences_envspace.png"), pseudoabsence_map,  # save figure
       width = 6, height = 5, dpi = 300, bg = "white")
```

# Extract climate for points

```{r extract-climate}
apicalis_df <- apicalis |>                                                          # start with sf presences
  dplyr::bind_cols(sf::st_coordinates(apicalis)) |>                                 # add lon/lat columns
  dplyr::rename(x = X, y = Y) |>                                                    # rename to x/y
  dplyr::mutate(percP = "pres") |>                                                  # label as presence
  dplyr::bind_rows(apicalis3) |>                                                    # append pseudo-absences
  dplyr::mutate(time_bp = -300) |>                                                  # time slice to extract
  as.data.frame() |>                                                                
  dplyr::select(-geometry)                                                          # drop geometry

apicalis_df <- location_slice_from_region_series(                                   # extract env time slice
  apicalis_df,
  region_series = climate_full
)
```

# Add climate back and cast to sf

```{r add-climate-sf}
apicalis4 <- apicalis |>                                                            # rebuild sf with env vars
  dplyr::bind_cols(sf::st_coordinates(apicalis)) |>
  dplyr::rename(x = X, y = Y) |>
  dplyr::mutate(percP = "pres") |>
  dplyr::bind_rows(apicalis3) |>
  dplyr::mutate(time_bp = -300) |>
  as.data.frame() |>
  dplyr::select(-geometry) |>
  dplyr::bind_cols(apicalis_df[, c("bio05", "bio06", "bio12")]) |>                 # attach env predictors
  dplyr::filter(!is.na(.data$bio05)) |>
  dplyr::mutate(percP = dplyr::if_else(percP == "pabs", "pseudoabsence", "presence")) |>
  sf::st_as_sf(coords = c("x", "y"), crs = 4326)                                   # back to sf
```

# Recipe

```{r recipe}
apicalis_rec <- recipes::recipe(apicalis4, formula = percP ~ bio05 + bio06 + bio12) # model formula
apicalis_rec                                                                        # print recipe summary
```

# Models & workflows

```{r models}
apicalis_models <- workflowsets::workflow_set(                                      # define model set
  preproc = list(default = apicalis_rec),
  models  = list(
    glm = sdm_spec_glm(),                                                           # GLM
    gam = sdm_spec_gam(),                                                           # GAM
    rf  = sdm_spec_rf(),                                                            # Random Forest
    gbm = sdm_spec_boost_tree()                                                     # Gradient Boosting
  ),
  cross = TRUE
) |>
  update_workflow_model(                                                            # set GAM smooths
    "default_gam",
    spec = sdm_spec_gam(),
    formula = percP ~ s(bio05, k = 10) + s(bio06, k = 10) + s(bio12, k = 10)
  ) |>
  option_add(control = control_ensemble_grid())                                     # tuning control
```

# Spatial block CV

```{r cv}
set.seed(1005)                                                                      # CV reproducibility
apicalis_cv <- spatial_block_cv(apicalis4, v = 5)                                   # spatial CV folds
autoplot(apicalis_cv)                                                               # quick diagnostics plot
```

# Tune

```{r tune}
set.seed(123)                                                                       # tuning reproducibility
apicalis_models <- apicalis_models |>
  workflow_map("tune_grid",
               resamples = apicalis_cv,                                             # use spatial CV
               grid = 5,                                                            # grid size
               metrics = sdm_metric_set(),                                          # SDM metrics incl. TSS
               verbose = TRUE)
```

# Build ensemble & inspect choices

```{r ensemble}
apicalis_ensemble <- simple_ensemble() |>                                           # start ensemble
  add_member(apicalis_models, metric = "tss_max")                                   # pick best by TSS

# Summarise chosen hyperparameters and TSS per workflow
chosen <- apicalis_models |>
  dplyr::mutate(
    best = purrr::map(result, ~ tune::select_best(.x, metric = "tss_max")),
    tss  = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "tss_max", n = 1)$mean),
    best = purrr::map_chr(best, ~ paste(names(.x), .x, sep = "=", collapse = "; "))
  ) |>
  dplyr::transmute(wflow_id, best, tss)
print(chosen)

# Fold-level performance boxplots
workflow_labels <- c(
  default_gam = "Generalised Additive Model",
  default_gbm = "Boosted Tree",
  default_glm = "Generalised Linear Model",
  default_rf  = "Random Forest"
)
model_order <- c("Generalised Additive Model", "Boosted Tree", "Generalised Linear Model", "Random Forest")
metric_labels <- c(boyce_cont = "Boyce Index", roc_auc = "AUC", tss_max = "TSS")
color_map <- c(
  "Generalised Additive Model" = "#FF61C3",
  "Boosted Tree"               = "#00B140",
  "Generalised Linear Model"   = "#00BFFF",
  "Random Forest"              = "#BC43FF"
)

# Collect fold metrics and draw
perf_data <- apicalis_models |>
  dplyr::mutate(metrics = purrr::map(result, collect_metrics, summarize = FALSE)) |>
  dplyr::select(wflow_id, metrics) |>
  tidyr::unnest(metrics) |>
  dplyr::filter(.metric %in% names(metric_labels)) |>
  dplyr::mutate(
    model        = factor(workflow_labels[wflow_id], levels = model_order),
    metric_label = factor(metric_labels[.metric], levels = c("Boyce Index", "AUC", "TSS"))
  )

perf_plot <- ggplot(perf_data, aes(x = model, y = .estimate, fill = model)) +
  geom_boxplot(width = 0.6, alpha = 0.9, outlier.shape = NA) +
  geom_jitter(aes(color = model), size = 1.3, width = 0.15, alpha = 0.4, stroke = 0.3, shape = 21) +
  scale_fill_manual(values = color_map, name = "Model Type") +
  scale_color_manual(values = rep("black", length(model_order)), guide = "none") +
  facet_wrap(~ metric_label, scales = "free_y") +
  labs(title = "Model performance across algorithms (fold-level)", x = NULL, y = "Performance metric") +
  theme_minimal(base_size = 12) +
  theme(
    panel.background  = element_rect(fill = "#EEEEEE", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.grid.major  = element_line(color = "white"),
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_blank(),
    axis.ticks.x      = element_blank(),
    legend.position   = "right",
    legend.direction  = "vertical",
    legend.title      = element_text(size = 11),
    legend.text       = element_text(size = 10),
    plot.title        = element_text(face = "bold", size = 13),
    strip.text        = element_text(margin = margin(b = 5))
  )

ggsave(file.path(results_dir, "ensemble", "ensemble_performance_clean.png"), perf_plot,
       width = 8, height = 5, dpi = 300)                                           # save figure
perf_plot                                                                            # print plot
```

# LGM snapshot

```{r lgm}
climate_lgm <- terra::rast(list(climate_full[[1]][[779]],                           # build 3-band env raster
                                climate_full[[2]][[779]],
                                climate_full[[3]][[779]]))
names(climate_lgm) <- c("bio05", "bio06", "bio12")                                  # name bands

prediction_lgm <- predict_raster(apicalis_ensemble, climate_lgm)                    # project ensemble

ggplot() +
  tidyterra::geom_spatraster(data = prediction_lgm, aes(fill = mean)) +
  tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
  coord_sf(xlim = xlim, ylim = ylim)
```

# Save model as SpatRaster stack

```{r save-stack}
apicalis_spatrast <- NULL                                                           # init empty stack
purrr::walk(760:800, function(timestep) {                                           # iterate timesteps
  year_bp <- time(climate_full[[1]])[timestep] - 1950                               # convert to years BP
  climate_snapshot <- terra::rast(list(                                             # assemble env snapshot
    climate_full[[1]][[timestep]],
    climate_full[[2]][[timestep]],
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")
  prediction_raster <- predict_raster(apicalis_ensemble, climate_snapshot)          # predict suitability
  apicalis_spatrast <<- if (is.null(apicalis_spatrast)) prediction_raster           # stack predictions
                      else c(apicalis_spatrast, prediction_raster)
})

time(apicalis_spatrast) <- time(climate_full[[1]])[760:800] - 1950                  # attach BP times

output_dir <- here("Results", "apicalis Krapp model")                               # output folder
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
writeRaster(apicalis_spatrast, filename = file.path(output_dir, "predictions_stack.tif"),
            overwrite = TRUE)                                                       # write GeoTIFF
```

# Stack o' maps

```{r stack-o-maps, dependson="globals"}
suppressPackageStartupMessages({ library(magick) })                                 # image post-processing
stopifnot(exists("australia_outline"), exists("results_dir"),                       # guardrails
          exists("climate_full"), exists("apicalis_ensemble"))

out_dir <- file.path(results_dir, "model"); if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

purrr::walk(760:800, function(timestep) {                                           # loop through timesteps
  year_bp <- time(climate_full[[1]])[timestep] - 1950
  climate_snapshot <- terra::rast(list(
    climate_full[[1]][[timestep]],
    climate_full[[2]][[timestep]],
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")
  prediction_raster <- predict_raster(apicalis_ensemble, climate_snapshot)

  p <- ggplot() +                                                                   # draw map
    tidyterra::geom_spatraster(data = prediction_raster, aes(fill = mean)) +
    tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +
    coord_sf(xlim = xlim, ylim = ylim) +
    ggtitle(bquote(italic("Leporillus apicalis") * " — Years BP: " * .(year_bp))) +
    theme_minimal() +
    theme(
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#F2F2F2", color = NA),
      panel.grid.major = element_line(color = "#C4C4C4"),
      panel.grid.minor = element_line(color = "#C4C4C4"),
      plot.margin      = margin(0, 0, 0, 0, unit = "pt")
    )

  plot_path <- file.path(out_dir, paste0("apicalis_map_", year_bp, ".png"))         # save path
  ggsave(plot_path, p, width = 7, height = 7, dpi = 300, bg = "transparent")       # save plot

  img <- image_read(plot_path)                                                      # trim/border
  img <- image_border(image_trim(img), color = "white", geometry = "30x30")
  image_write(img, path = plot_path)
})
```

# 6-image panel (model only)

```{r six-panel}
raster_times  <- terra::time(apicalis_spatrast)                                     # all BP times
target_times  <- c(-40000, -26000, -21000, -12000, -6000, 0)                        # selected panels
matched_idx   <- match(target_times, raster_times)                                   # index match
panel_tags    <- letters[seq_along(target_times)]                                    # (a)-(f)

panel_plots <- pmap(
  list(idx = matched_idx, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- apicalis_spatrast[[idx]]                                  # pick raster layer
    ggplot() +
      tidyterra::geom_spatraster(data = current_prediction, aes(fill = mean)) +
      tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +
      coord_sf(xlim = xlim, ylim = ylim) +
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(year_bp))) +
      theme_minimal(base_size = 10) +
      theme(
        panel.background = element_rect(fill = "#DCF3FC", color = NA),
        plot.background  = element_rect(fill = "white", color = NA),
        panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_blank(),
        axis.text        = element_text(size = 8),
        axis.title       = element_blank(),
        plot.margin      = margin(2, 2, 2, 2),
        legend.position  = "right",
        legend.title     = element_text(size = 11),
        legend.text      = element_text(size = 11)
      )
  }
)

final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +              # combine panels
  plot_annotation(title = expression(italic("Leporillus apicalis") * " SDM model predictions"),
                  theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)))

ggsave(file.path(results_dir, "apicalis_model_panel.png"), final_panel,             # save figure
       width = 10, height = 9, dpi = 300)
```

# Centroid through time & diff map (0 vs 20 ka)

```{r centroid-diff, dependson="globals"}
# Compute weighted centroids over time
apicalis_centroids <- lapply(1:nlyr(apicalis_spatrast), function(i) {
  r <- apicalis_spatrast[[i]]
  vals <- terra::values(r)[, 1]                                                     # suitability values
  coords <- terra::crds(r, na.rm = FALSE)                                           # cell centroids
  valid <- !is.na(vals) & vals > 0
  if (!any(valid)) return(NULL)
  w <- vals[valid]; cxy <- coords[valid, ]
  data.frame(time_bp = time(apicalis_spatrast)[i],
             longitude = sum(cxy[,1] * w) / sum(w),
             latitude  = sum(cxy[,2] * w) / sum(w))
}) |>
  dplyr::bind_rows()

# Difference map: present minus 20 ka
r_times <- as.numeric(terra::time(apicalis_spatrast))
idx_20  <- which.min(abs(r_times - (-20000)))
idx_0   <- which.min(abs(r_times - 0))
r_diff  <- apicalis_spatrast[[idx_0]] - apicalis_spatrast[[idx_20]]

# Symmetric colour scale based on robust limits
vals <- terra::values(r_diff, mat = FALSE)
lims <- quantile(abs(vals), 0.98, na.rm = TRUE) |> as.numeric()
lims <- max(0.1, min(1, lims))

fill_scale <- scale_fill_gradient2(
  name = "Change in suitability\nsince 20 ka",
  low = "#3B4CC0", mid = "white", high = "#B40426",
  midpoint = 0, limits = c(-lims, lims)
)

# Endpoints for legend markers
pt_20 <- apicalis_centroids[which.min(abs(apicalis_centroids$time_bp + 20000)), , drop = FALSE]
pt_0  <- apicalis_centroids[which.min(abs(apicalis_centroids$time_bp - 0)),       , drop = FALSE]

# Plot single-panel diff + trajectory
p_single <- ggplot() +
  tidyterra::geom_spatraster(data = r_diff, aes(fill = mean)) +
  fill_scale +
  geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.4) +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  geom_path(data = apicalis_centroids, aes(longitude, latitude), color = "black", linewidth = 0.7, alpha = 0.85) +
  geom_point(data = apicalis_centroids, aes(longitude, latitude, color = time_bp), size = 2) +
  geom_point(data = pt_20, aes(longitude, latitude, shape = "20 ka centroid"), size = 3, fill = "white", color = "black", stroke = 0.6) +
  geom_point(data = pt_0,  aes(longitude, latitude, shape = "Present centroid"), size = 3, fill = "yellow", color = "black", stroke = 0.6) +
  scale_color_viridis_c(name = "Years BP", option = "H",
                        limits = c(-40000, 0), breaks = c(-40000, -30000, -20000, -10000, 0),
                        labels = c("40 ka", "30 ka", "20 ka", "10 ka", "0")) +
  scale_shape_manual(name = "Centroid positions", values = c("20 ka centroid" = 24, "Present centroid" = 21)) +
  labs(x = "Longitude", y = "Latitude") +
  theme_map_clean +
  guides(fill = guide_colorbar(title.position = "top", order = 1),
         color = guide_colorbar(title.position = "top", order = 2),
         shape  = guide_legend(title.position  = "top", order = 3))

ggsave(file.path(results_dir, "figure_range_shift_single.png"), p_single,          # save figure
       width = 7.5, height = 6.5, dpi = 300, bg = "white")
```

# Middens validation prep

```{r middens-load}
dated_middens <- read.csv(file.path(processed_dir, "datedmiddens.csv"))             # load midden metadata

middens_sf <- dated_middens |>                                                      # to sf, WGS84
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  sf::st_transform(crs = terra::crs(apicalis_spatrast))                             # match raster CRS

middens_vect <- terra::vect(middens_sf)                                             # to terra vect for extract
rt <- as.numeric(terra::time(apicalis_spatrast))                                    # numeric BP timesteps

# Bin each midden to nearest raster timestep and convert to calendar date
middens_sf <- middens_sf |>
  dplyr::mutate(
    binned_time_bp = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
    nearest_idx    = match(binned_time_bp, rt),
    lubridate_time = pastclim::ybp2date(time_bp)
  )

# Extract predicted suitability at each midden’s nearest timestep
extracted_values <- terra::extract(apicalis_spatrast, middens_vect, ID = FALSE)
vals_mat <- as.matrix(extracted_values)
middens_sf$predicted <- vals_mat[cbind(seq_len(nrow(middens_sf)), middens_sf$nearest_idx)]

# Summarise counts/means per timestep for quick QA
raster_timesteps <- rt
timesteps_df <- data.frame(raster_timestep = raster_timesteps)

organized_data <- timesteps_df |>
  dplyr::left_join(
    middens_sf |>
      dplyr::group_by(binned_time_bp) |>
      dplyr::summarise(count = dplyr::n(), mean_value = mean(predicted, na.rm = TRUE), .groups = "drop"),
    by = c("raster_timestep" = "binned_time_bp")
  ) |>
  dplyr::mutate(count = tidyr::replace_na(count, 0),
                mean_value = tidyr::replace_na(mean_value, NA_real_))
print(organized_data)
```

# Middens-overlaid maps (loop)

```{r middens-maps, dependson="globals"}
# Load `magick` for image trimming / borders
suppressPackageStartupMessages({ library(magick) })

# Get Australia outline
world_map <- rworldmap::getMap(resolution = "high")
australia_outline <- sf::st_as_sf(world_map) %>%
  dplyr::filter(ADMIN == "Australia")

# Extract the time stamps from the prediction stack
raster_timesteps <- time(apicalis_spatrast)

# Loop through each time slice and build a map
walk(seq_along(raster_timesteps), function(i) {

  # Current slice in years BP
  current_year_bp <- raster_timesteps[i]

  # Middens close in time (± 500 yrs)
  current_middens <- middens_sf %>%
    dplyr::filter(abs(time_bp - current_year_bp) <= 500)

  # Pull the raster for this timestep
  current_prediction <- apicalis_spatrast[[i]]

  # Build plot
  p <- ggplot() +
    tidyterra::geom_spatraster(data = current_prediction, aes(fill = mean)) +
    tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +
    geom_sf(data = current_middens, shape = 21,
            fill = "#FF3399", color = "black",
            size = 2, stroke = 0.4) +
    coord_sf(xlim = xlim, ylim = ylim) +
    ggtitle(bquote(italic("Leporillus apicalis") *
                   " SDM with Middens — Years BP: " *
                   .(current_year_bp))) +
    theme_minimal(base_size = 11) +
    theme(
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#F2F2F2", color = NA),
      panel.grid.major = element_line(color = "#C4C4C4"),
      panel.grid.minor = element_line(color = "#C4C4C4"),
      plot.margin      = margin(0, 0, 0, 0, "pt")
    )

  # Path for saving output
  plot_path <- file.path(results_dir,
                         "midden maps",
                         paste0("apicalis_middens_", current_year_bp, ".png"))

  # Ensure output directory exists
  dir.create(dirname(plot_path), showWarnings = FALSE, recursive = TRUE)

  # Save PNG
  ggsave(plot_path, p,
         width = 7, height = 7, dpi = 300,
         bg = "white")

  # Trim and add border for visual neatness
  img <- magick::image_read(plot_path) %>%
    magick::image_trim() %>%
    image_border(color = "white", geometry = "30x30")

  magick::image_write(img, path = plot_path)
})
```
# 6-panel with middens

```{r six-panel-middens}
# Extract all raster timesteps
raster_times <- terra::time(apicalis_spatrast)

# Target slices to display (Years BP)
target_times <- c(-28000, -15000, -12000, -7000, -4000, 0)

# Match targets to nearest raster indices
matched_indices <- match(target_times, raster_times)

# Panel tag labels (a–f)
panel_tags <- letters[seq_along(target_times)]

# Build individual panels
panel_plots <- pmap(
  list(idx = matched_indices, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {

    # Prediction raster at that timestep
    current_prediction <- apicalis_spatrast[[idx]]

    # Middens near that timestep (±500 yrs)
    current_middens <- middens_sf %>%
      dplyr::filter(abs(time_bp - year_bp) <= 500) %>%
      dplyr::mutate(type = "Midden")

    ggplot() +
      tidyterra::geom_spatraster(
        data = current_prediction,
        aes(fill = mean)
      ) +
      tidyterra::scale_fill_terrain_c(
        name   = "Probability\nof occurrence",
        limits = c(0, 1)
      ) +
      geom_sf(
        data   = australia_outline,
        fill   = NA,
        color  = "black",
        linewidth = 0.5
      ) +
      geom_sf(
        data = current_middens,
        aes(color = type, shape = type),
        fill  = "#FF3399",
        size  = 2,
        stroke = 0.4,
        show.legend = TRUE
      ) +
      scale_color_manual(name = NULL, values = c("Midden" = "black")) +
      scale_shape_manual(name = NULL, values = c("Midden" = 21)) +
      coord_sf(xlim = xlim, ylim = ylim) +
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(
        bquote("(" * .(tag_label) * ")  Years BP: " * .(abs(year_bp)))
      ) +
      theme(
        plot.background   = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "#E8E8E8", color = NA),
        panel.grid.major  = element_line(color = "#C4C4C4"),
        panel.grid.minor  = element_line(color = "#C4C4C4"),
        axis.text         = element_text(size = 8),
        axis.title        = element_blank(),
        plot.margin       = margin(2, 2, 2, 2),
        legend.position   = "bottom",
        legend.box        = "horizontal",
        legend.justification = "center",
        legend.spacing.x  = unit(1.2, "cm"),
        legend.spacing.y  = unit(0.4, "cm"),
        legend.key        = element_rect(fill = "transparent", color = NA),
        legend.title      = element_text(size = 11),
        legend.text       = element_text(size = 11)
      ) +
      guides(
        fill = guide_colorbar(
          title          = "Probability of occurrence",
          title.position = "top",
          title.theme    = element_text(size = 11),
          label.theme    = element_text(size = 11),
          direction      = "horizontal",
          barwidth       = unit(5, "cm"),
          barheight      = unit(0.4, "cm")
        ),
        color = guide_legend(
          override.aes = list(size = 2.5),
          title.theme  = element_text(size = 11),
          label.theme  = element_text(size = 11)
        ),
        shape = guide_legend(
          title.theme = element_text(size = 11),
          label.theme = element_text(size = 11)
        )
      )
  }
)

# Combine panels + shared legend + title
final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +
  plot_annotation(
    title = "Leporillus apicalis SDM with middens",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )
  ) &
  theme(
    legend.position   = "bottom",
    legend.box        = "horizontal",
    legend.justification = "center",
    legend.spacing.x  = unit(1.2, "cm"),
    legend.spacing.y  = unit(0.4, "cm"),
    legend.key        = element_rect(fill = "transparent", color = NA)
  )

# Save final panel
ggsave(
  file.path(results_dir, "midden_panel_plot.png"),
  final_panel,
  width = 10, height = 9, dpi = 300
)
```

# Threshold, pseudo-absences (time-aware), and evaluation

```{r tss-maxsss}
# ---- Bootstrap a starting threshold for time-aware pseudo-absences ----
threshold_init <- 0.27                                                        # initial cut-off
apicalis_binary_init <- terra::app(x = apicalis_spatrast, fun = \(i) i > threshold_init)  # binarise stack
terra::time(apicalis_binary_init) <- terra::time(apicalis_spatrast) |> pastclim::ybp2date() # set dates
apicalis_rs_init <- terra::sds(apicalis_binary_init); names(apicalis_rs_init) <- "pres_abs" # wrap as sds

# ---- Sample pseudo-absences with temporal & spatial rules ----
pseudoabs_init <- sample_pseudoabs_time(
  data            = dated_middens,                                            # dated midden points
  raster          = apicalis_rs_init,                                         # binary model through time
  n_per_presence  = 3,                                                        # 3 PAs per presence
  time_col        = "time_bp",                                                # input time column (BP)
  lubridate_fun   = pastclim::ybp2date,                                       # BP -> calendar converter
  method          = c("dist_min", km2m(200)),                                 # min distance in space
  time_buffer     = 365000 * 2                                                # ±2 ka temporal buffer (days)
)

# ---- Extract continuous predictions matched to each row’s nearest timestep ----
rt <- as.numeric(terra::time(apicalis_spatrast))                              # all model timesteps (BP)

# Presence values (match each midden to nearest model time slice)
presence_sf <- sf::st_as_sf(dated_middens, coords = c("longitude", "latitude"), crs = 4326) |>
  sf::st_transform(crs(apicalis_spatrast)) |>
  dplyr::mutate(
    nearest_time = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),   # nearest model time
    nearest_idx  = match(nearest_time, rt)                                    # index for extraction
  )
pres_mat <- terra::extract(apicalis_spatrast, terra::vect(presence_sf), ID = FALSE)
presence_vals <- as.matrix(pres_mat)[cbind(seq_len(nrow(presence_sf)), presence_sf$nearest_idx)]

# Pseudo-absence values (same time-matching as above)
pseudoabs_sf <- sf::st_as_sf(pseudoabs_init, coords = c("longitude", "latitude"), crs = 4326) |>
  sf::st_transform(crs(apicalis_spatrast)) |>
  dplyr::mutate(
    time_bp      = pastclim::date2ybp(time_step),                             # convert to BP
    nearest_time = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
    nearest_idx  = match(nearest_time, rt)
  )
pseudo_mat <- terra::extract(apicalis_spatrast, terra::vect(pseudoabs_sf), ID = FALSE)
pseudoabs_vals <- as.matrix(pseudo_mat)[cbind(seq_len(nrow(pseudoabs_sf)), pseudoabs_sf$nearest_idx)]

# ---- Grid search for the MaxSSS threshold (using TSS = sens + spec - 1) ----
predicted_values <- c(presence_vals, pseudoabs_vals)                          # all scores
observed <- c(rep(1, length(presence_vals)), rep(0, length(pseudoabs_vals)))  # labels
keep <- is.finite(predicted_values) & is.finite(observed)                     # guard against NA/Inf
predicted_values <- predicted_values[keep]; observed <- observed[keep]

thresholds <- seq(0, 1, by = 0.01)                                            # search grid
results <- data.frame(threshold = thresholds, TSS = NA_real_)                 # holder
for (i in seq_along(thresholds)) {                                            # loop grid
  thr <- thresholds[i]
  pred_binary <- ifelse(predicted_values > thr, 1L, 0L)
  a <- sum(pred_binary == 1L & observed == 1L)                                 # TP
  b <- sum(pred_binary == 1L & observed == 0L)                                 # FP
  c <- sum(pred_binary == 0L & observed == 1L)                                 # FN
  d <- sum(pred_binary == 0L & observed == 0L)                                 # TN
  if ((a + c) == 0 || (b + d) == 0) next                                       # skip degenerate cases
  sens <- a / (a + c); spec <- d / (b + d)                                     # sensitivity & specificity
  results$TSS[i] <- sens + spec - 1                                            # TSS
}
opt_threshold <- results$threshold[which.max(results$TSS)]                     # best threshold
cat("Optimal Threshold (maxSSS):", opt_threshold, "\n")

# ---- Rebuild binary stack at optimal threshold ----
apicalis_binary <- terra::app(x = apicalis_spatrast, fun = \(i) i > opt_threshold) # binarise with optimum
terra::time(apicalis_binary) <- terra::time(apicalis_spatrast) |> pastclim::ybp2date() # set dates
apicalis_rs <- terra::sds(apicalis_binary); names(apicalis_rs) <- "pres_abs"         # sds wrapper

# ---- Resample pseudo-absences with final thresholded stack ----
pseudoabs <- sample_pseudoabs_time(
  data            = dated_middens,
  raster          = apicalis_rs,
  n_per_presence  = 3,
  time_col        = "time_bp",
  lubridate_fun   = pastclim::ybp2date,
  method          = c("dist_min", km2m(200)),
  time_buffer     = 365000 * 2
)

# ---- Compute evaluation metrics (TSS, Sørensen, F2) ----
pres_abs_observed <- pseudoabs |>
  dplyr::mutate(time_bp = pastclim::date2ybp(time_step)) |>                   # add BP
  dplyr::select(-class, -time_step) |>
  tibble::rowid_to_column("name")                                             # ensure row id

pres_abs_pred <- location_slice_from_region_series(                           # extract model class
  x = pres_abs_observed, region_series = apicalis_rs
)
pres_abs_pred$class <- pseudoabs$class                                        # add true class

# Confusion components
a <- nrow(dplyr::filter(pres_abs_pred, class == "presence",  pres_abs == 1))  # TP
b <- nrow(dplyr::filter(pres_abs_pred, class == "pseudoabs", pres_abs == 1))  # FP
c <- nrow(dplyr::filter(pres_abs_pred, class == "presence",  pres_abs == 0))  # FN
d <- nrow(dplyr::filter(pres_abs_pred, class == "pseudoabs", pres_abs == 0))  # TN

tpr <- a / (a + c)                                                            # sensitivity
tnr <- d / (b + d)                                                            # specificity
tss <- tpr + tnr - 1                                                          # TSS
sorensen <- 2 * (a / (c + (2 * a) + d))                                       # Sørensen/Dice
precision <- a / (a + b)                                                      # precision
f2score <- 2 * ((precision * tpr) / (precision + tpr))                        # F2 (β=2)

cat(sprintf("True Skill Statistic: %.3f\n", tss))
cat(sprintf("Sorensen: %.3f\n", sorensen))
cat(sprintf("F2 score: %.3f\n", f2score))
```

# Panels (a) & (b): histograms (full vs coarse) and persistence

```{r hist-panels}
suppressPackageStartupMessages({ library(cowplot) })                           # for panel labelling
stopifnot(inherits(apicalis_spatrast, "SpatRaster"),
          inherits(middens_sf, "sf"),
          dir.exists(results_dir))                                             # sanity checks

# ---- Figure parameters & thresholds ----
step_years <- 1000                                                             # temporal step (years)
rt <- as.numeric(terra::time(apicalis_spatrast))                               # BP vector
thr_a <- if (exists("opt_threshold")) opt_threshold else 0.27                  # vertical line on hist
thr_b <- 0.35                                                                  # persistence threshold
lookback_ka <- 30                                                              # look back 30 ka
include_current <- TRUE                                                        # count current slice
coarse_factor <- 2                                                             # 2x coarser than base
w_single <- 7; h_single <- 5; dpi_out <- 300                                   # output size

# ---- Useful helpers ----
res_full_deg   <- terra::res(apicalis_spatrast)[1]                             # native res (deg)
res_coarse_deg <- res_full_deg * coarse_factor                                 # coarse res (deg)

ensure_time_cols <- function(m_sf, rt) {                                       # ensure binned time cols
  if (!all(c("binned_time_bp","nearest_idx") %in% names(m_sf))) {
    m_sf <- m_sf %>% dplyr::mutate(
      binned_time_bp = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
      nearest_idx    = match(binned_time_bp, rt)
    )
  }
  m_sf
}

fd_bin_count <- function(x) {                                                  # Freedman–Diaconis-ish bins
  x <- x[is.finite(x)]; if (!length(x)) return(30L)
  fd_bw <- 2 * IQR(x, na.rm = TRUE) / (length(x)^(1/3))
  max(20, min(60, round(1 / max(min(fd_bw, 0.08), 0.02))))
}

time_matched_background <- function(m_sf, rast){                               # background matched in time
  rt <- as.numeric(terra::time(rast))
  m_sf <- ensure_time_cols(m_sf, rt)
  split(m_sf, m_sf$binned_time_bp) |>
    lapply(function(df_i){
      tb  <- unique(df_i$binned_time_bp)
      lyr <- match(tb, rt)
      n_i <- nrow(df_i)
      vals <- terra::spatSample(rast[[lyr]], size = n_i, method = "random", values = TRUE, na.rm = TRUE)[,1] |> as.numeric()
      tibble::tibble(suitability = vals, group = "Background (time-matched)", bin = tb)
    }) |>
    dplyr::bind_rows()
}

compute_persistence_strict <- function(rast, m_sf, thr, include_current, lookback_ka, step_years){
  rt <- as.numeric(terra::time(rast))
  m_sf <- ensure_time_cols(m_sf, rt)
  mat <- terra::extract(rast, terra::vect(m_sf), ID = FALSE) |> as.matrix()    # suitability time series
  lookback_steps <- if (is.finite(lookback_ka)) as.integer(lookback_ka) else Inf
  start_idx <- if (!include_current) pmax(m_sf$nearest_idx - 1L, 0L) else m_sf$nearest_idx
  lead_true_count <- function(x){ n <- 0L; for (v in x) { if (isTRUE(v)) n <- n + 1L else break }; n }
  steps <- vapply(seq_len(nrow(mat)), function(i){
    idx <- start_idx[i]; if (idx <= 0L) return(0L)
    from <- idx; to <- if (is.finite(lookback_steps)) max(idx - lookback_steps + 1L, 1L) else 1L
    sidx <- seq(from, to, by = -1L)                                            # walk backwards in time
    above <- mat[i, sidx] > thr
    lead_true_count(above)                                                     # consecutive TRUEs
  }, integer(1))
  tibble::tibble(persist_ka = steps * (step_years/1000))                       # convert to ka
}

# ---- Prepare data at native resolution ----
middens_sf <- ensure_time_cols(middens_sf, rt)
mid_df_full <- tibble::tibble(
  suitability = sf::st_drop_geometry(middens_sf)$predicted,                    # midden scores
  group = "Middens",
  bin = middens_sf$binned_time_bp
)
bg_df_full  <- time_matched_background(middens_sf, apicalis_spatrast)          # matched background
plot_df_full <- dplyr::bind_rows(mid_df_full, bg_df_full) |>
  dplyr::filter(is.finite(suitability))
bins_a_full <- fd_bin_count(mid_df_full$suitability)

# ---- Histogram (native resolution) ----
p_a_full <- ggplot(plot_df_full, aes(x = suitability, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.65, bins = bins_a_full, boundary = 0, color = "white") +
  geom_vline(xintercept = thr_a, linetype = "dashed", linewidth = 0.6) +
  scale_fill_manual(values = c("Middens" = "#BC43FF", "Background (time-matched)" = "#9AA0A6")) +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,0.1)) +
  labs(title = sprintf("Habitat suitability at dated midden locations (%.1f°)", terra::res(apicalis_spatrast)[1]),
       x = "Modelled suitability (0–1)", y = "Count", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "#F2F2F2", color = NA),
        panel.grid.major = element_line(color = "white"),
        legend.position = "bottom")
ggsave(file.path(results_dir, "panel_a_full_hist.png"), p_a_full, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

# ---- Build coarse version (2× cell size) and repeat ----
apicalis_spatrast_coarse <- terra::aggregate(apicalis_spatrast, fact = coarse_factor, fun = mean, na.rm = TRUE)
terra::time(apicalis_spatrast_coarse) <- terra::time(apicalis_spatrast)        # keep time vector

mid_vals_c <- terra::extract(apicalis_spatrast_coarse, terra::vect(middens_sf), ID = FALSE) |> as.matrix()
pick_idx   <- cbind(seq_len(nrow(mid_vals_c)), middens_sf$nearest_idx)
mid_suit_c <- mid_vals_c[pick_idx] |> as.numeric()
mid_df_c <- tibble::tibble(suitability = mid_suit_c, group = "Middens", bin = middens_sf$binned_time_bp)
bg_df_c  <- time_matched_background(middens_sf, apicalis_spatrast_coarse)
plot_df_c <- dplyr::bind_rows(mid_df_c, bg_df_c) |> dplyr::filter(is.finite(suitability))
bins_a_c  <- fd_bin_count(mid_df_c$suitability)

p_a_coarse <- ggplot(plot_df_c, aes(x = suitability, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.65, bins = bins_a_c, boundary = 0, color = "white") +
  geom_vline(xintercept = thr_a, linetype = "dashed", linewidth = 0.6) +
  scale_fill_manual(values = c("Middens" = "#BC43FF", "Background (time-matched)" = "#9AA0A6")) +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,0.1)) +
  labs(title = sprintf("Habitat suitability at dated midden locations (%.1f°)", terra::res(apicalis_spatrast_coarse)[1]),
       x = "Modelled suitability (0–1)", y = "Count", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "#F2F2F2", color = NA),
        panel.grid.major = element_line(color = "white"),
        legend.position = "bottom")
ggsave(file.path(results_dir, "panel_a_coarse_hist.png"), p_a_coarse, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

# ---- Persistence histograms (strict consecutive suitability) ----
persist_full <- compute_persistence_strict(apicalis_spatrast, middens_sf, thr = thr_b, include_current = include_current, lookback_ka = lookback_ka, step_years = step_years)
x_max_b <- if (is.finite(lookback_ka)) lookback_ka else max(persist_full$persist_ka, na.rm = TRUE)

p_b_full <- ggplot(persist_full, aes(x = persist_ka)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left", color = "white", fill = "#BC43FF", alpha = 0.9) +
  scale_x_continuous(limits = c(0, x_max_b), breaks = seq(0, x_max_b, by = 10), labels = function(x) paste0(x, " ka")) +
  labs(title = sprintf("Duration of suitability prior to midden deposition (%.1f°)", terra::res(apicalis_spatrast)[1]),
       x = sprintf("Years suitable before deposition (ka; suitability \u2265 %.2f)", thr_b), y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "#F2F2F2", color = NA),
        panel.grid.major = element_line(color = "white"))
ggsave(file.path(results_dir, "panel_b_full_hist_thr035.png"), p_b_full, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

persist_coarse <- compute_persistence_strict(apicalis_spatrast_coarse, middens_sf, thr = thr_b, include_current = include_current, lookback_ka = lookback_ka, step_years = step_years)

p_b_coarse <- ggplot(persist_coarse, aes(x = persist_ka)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left", color = "white", fill = "#BC43FF", alpha = 0.9) +
  scale_x_continuous(limits = c(0, x_max_b), breaks = seq(0, x_max_b, by = 10), labels = function(x) paste0(x, " ka")) +
  labs(title = sprintf("Duration of suitability prior to midden deposition (%.1f°)", terra::res(apicalis_spatrast_coarse)[1]),
       x = sprintf("Years suitable before deposition (ka; suitability \u2265 %.2f)", thr_b), y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "#F2F2F2", color = NA),
        panel.grid.major = element_line(color = "white"))
ggsave(file.path(results_dir, "panel_b_coarse_hist_thr035.png"), p_b_coarse, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

# ---- Compose a vertical figure with (a) + (b) ----
p_a_tag <- cowplot::ggdraw(p_a_coarse + labs(title = NULL)) +
  cowplot::draw_label("(a)", x = 0.02, y = 0.98, hjust = 0, vjust = 1, fontface = "bold")
p_b_tag <- cowplot::ggdraw(p_b_coarse + labs(title = NULL)) +
  cowplot::draw_label("(b)", x = 0.02, y = 0.98, hjust = 0, vjust = 1, fontface = "bold")

fig_ab <- cowplot::plot_grid(p_a_tag, p_b_tag, ncol = 1, rel_heights = c(1,1))
print(fig_ab)

ggsave(file.path(results_dir, "figure_histograms_panels_ab_COARSE_thrB_035_VERTICAL.png"),
       fig_ab, width = 7, height = 10, dpi = dpi_out, bg = "white")
```

# Variable importance table + PDPs

```{r vip-pdp}
suppressPackageStartupMessages({
  library(DALEX); library(tidysdm); library(ggplot2); library(dplyr); library(knitr)
})                                                                              # quiet attaches

# ---- Build explainer & output dir ----
apicalis_explainer <- explain_tidysdm(apicalis_ensemble)                        # DALEX explainer
apicalis_outdir <- file.path(results_dir, "variable_importance")
dir.create(apicalis_outdir, showWarnings = FALSE, recursive = TRUE)

# ---- Aggregate permutation importance into a neat table ----
apicalis_vip <- model_parts(explainer = apicalis_explainer) |>
  dplyr::filter(!variable %in% c("_baseline_", "_full_model_")) |>
  dplyr::group_by(variable) |>
  dplyr::summarise(
    `Mean drop-out loss` = round(mean(dropout_loss, na.rm = TRUE), 3),
    `SD`                 = round(sd(dropout_loss,   na.rm = TRUE), 3),
    .groups = "drop_last"
  ) |>
  dplyr::arrange(desc(`Mean drop-out loss`)) |>
  dplyr::rename(`Environmental variable` = variable)

print(apicalis_vip)                                                             # print in console
write.csv(apicalis_vip, file.path(apicalis_outdir, "apicalis_variable_importance_table.csv"),
          row.names = FALSE)                                                    # save CSV
knitr::kable(apicalis_vip,
             caption = "Variable importance summary (Leporillus apicalis ensemble model)",
             format = "markdown")                                               # show as table in Rmd

# ---- Helper to compute & save PDPs for a single variable ----
plot_and_save_pdp <- function(var, label){
  pdp <- model_profile(apicalis_explainer, N = 500, variables = var)           # partial dependence
  p <- ggplot(pdp$agr_profiles, aes(x = `_x_`, y = `_yhat_`)) +
    geom_line(linewidth = 1) +
    labs(title = paste("Leporillus apicalis – Partial dependence on", label),
         x = label, y = "Predicted suitability") +
    theme_minimal(base_size = 12) +
    theme(plot.background  = element_rect(fill = "white", color = NA),
          panel.background = element_rect(fill = "#F2F2F2", color = NA),
          panel.grid.major = element_line(color = "#C4C4C4", linewidth = 0.3),
          panel.border     = element_rect(color = "#C4C4C4", fill = NA, linewidth = 0.3),
          plot.title       = element_text(face = "italic", size = 13))
  ggsave(file.path(apicalis_outdir, paste0("apicalis_pdp_", var, ".png")),
         p, width = 6, height = 4, dpi = 300, bg = "white")
  print(p)
}

# ---- Generate PDPs ----
plot_and_save_pdp("bio05", "Maximum temperature of warmest month (°C)")
plot_and_save_pdp("bio06", "Minimum temperature of coldest month (°C)")
plot_and_save_pdp("bio12", "Annual precipitation (mm)")
```

# KDE of midden sampling density (optional viz)

```{r kde-middens}
suppressPackageStartupMessages({ library(rnaturalearth); library(rnaturalearthdata) })  # basemaps

# ---- Build a simple df of midden XYs ----
middens_df_2 <- middens_sf_2 %>%                                               # assumes middens_sf_2 exists
  dplyr::mutate(
    longitude = sf::st_coordinates(.)[, 1],
    latitude  = sf::st_coordinates(.)[, 2]
  ) %>%
  sf::st_drop_geometry()

# ---- Basemap & padding points to stabilise KDE near edges ----
australia_outline_ne <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  dplyr::filter(admin == "Australia")

pad_points <- data.frame(                                                       # corners of plotting box
  longitude = c(110, 150, 150, 110),
  latitude  = c(-10, -10, -45, -45)
)
middens_df_padded <- dplyr::bind_rows(middens_df_2, pad_points)

# ---- Plot KDE of midden density ----
p <- ggplot() +
  geom_sf(data = australia_outline_ne, fill = scales::alpha("#C4C4C4", 0.5), color = "black") +
  stat_density_2d(
    data   = middens_df_padded,
    aes(x = longitude, y = latitude, fill = after_stat(level)),
    geom   = "polygon",
    contour = TRUE,
    color  = NA,
    adjust = 0.4,
    alpha  = 0.8,
    n      = 200
  ) +
  scale_fill_viridis_c(option = "magma", name = "Density", direction = -1) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +
  labs(title = "Spatial density of dated midden samples",
       x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 12) +
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "#C4C4C4"),
    panel.grid.minor = element_line(color = "#C4C4C4"),
    legend.position  = c(0.03, 0.20),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color = "grey70", linewidth = 0.3),
    legend.key       = element_rect(fill = "white", color = NA),
    legend.margin    = margin(6, 12, 6, 9),
    legend.title     = element_text(size = 10),
    legend.text      = element_text(size = 9)
  ) +
  guides(fill = guide_colorbar(
    title = "Density", title.position = "top",
    title.theme = element_text(size = 10),
    label.theme = element_text(size = 9),
    direction  = "horizontal",
    barwidth   = unit(4, "cm"),
    barheight  = unit(0.4, "cm"),
    ticks = TRUE, ticks.colour = "black"
  ))

# ---- Save KDE figure ----
ggsave(file.path(here("Results/stratified environmental sampling"), "midden_sampling_density.png"),
       p, width = 7, height = 6, dpi = 300, bg = "white")
```


# Citations dump (optional)

```{r citations, eval=FALSE}
# Print citations for packages used (handy for thesis/manuscript methods sections)
lapply(
  c("tidysdm", "tidymodels", "workflowsets", "workflows", "tune",
    "parsnip", "recipes", "yardstick", "rsample", "spatialsample",
    "dials", "dplyr", "tidyr", "purrr", "tibble", "ggplot2",
    "cowplot", "scales", "viridis", "viridisLite", "sf", "terra",
    "tidyterra", "pastclim", "oz", "rworldmap", "DALEX", "ecospat",
    "broom", "infer", "readr", "fs", "gtools", "RefManageR"),
  function(pkg) {
    cat("\n---", pkg, "---\n")
    print(citation(pkg))
  }
)
```