---
title: "Krapp conditor"
author: "Annie"
date: "2024-06-20"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

Using code written for paleoview

```{r setup, include=FALSE}
# Set global chunk options to display code by default
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r install_libraries, message=FALSE, warning=FALSE}

# List of packages to check, install if needed, and load
packages <- c(
  "broom", "car", "DALEX", "dials", "dplyr", "ggplot2", "infer", 
  "modeldata", "oz", "parsnip", "pastclim", "purrr", "readr", 
  "recipes", "rsample", "rworldmap", "sf", "tidyr", "tidymodels", 
  "tidysdm", "tidyterra", "tibble", "tune", "terra", "workflows", 
  "workflowsets", "gtools", "fs"
)

# Function to display a cat message
cat_message <- function(pkg, message_type) {
  if (message_type == "installed") {
    cat(paste0(pkg, " has been installed successfully!\n"))
  } else if (message_type == "loading") {
    cat(paste0(pkg, " is already installed and has been loaded!\n"))
  }
}

# Install missing packages and load them
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
    cat_message(pkg, "installed")
  } else {
    cat_message(pkg, "loading")
  }
  library(pkg, character.only = TRUE)
}

# If you are using renv, you can snapshot the renv after loading all the packages.

# renv::snapshot()
```

```{r libraries}
# Load necessary libraries for data manipulation, spatial analysis, plotting, and modeling
library(broom)
library(car)
library(DALEX)
library(dials)
library(ggplot2)
library(infer)
library(modeldata)
library(oz)
library(parsnip)
library(pastclim)
library(purrr)
library(readr)
library(recipes)
library(rsample)
library(rworldmap)
library(sf)
library(tidyr)
library(tidymodels)
library(tidysdm)
library(tidyterra)
library(tibble)
library(tune)
library(terra)
library(workflows)
library(workflowsets)
library(gtools)
library(fs)
library(dplyr)
```

# Set the working directory

```{r Set the working directory}
# Set the working directory to the processed data folder
knitr::opts_knit$set(root.dir =("C:/github/SNR_SDM/Data/processed"))
```

# Preparing your data

## Download Krapp2021 dataset

```{r download climate}
# Download the Krapp2021 dataset using pastclim package
pastclim::download_dataset("Krapp2021")
```

## Read conditor data

```{r read file}
# Set working directory to the processed data folder
setwd("C:/github/SNR_SDM/Data/processed")

# Read in the 'conditor.csv' file, print its contents, convert to an sf object, and set the CRS
conditor <- read_csv("conditor.csv") %>%
  {print(.); .} %>%  # Print the data to check its contents
  st_as_sf(coords = c("longitude", "latitude")) %>%  # Convert to a simple features (sf) object using latitude and longitude
  st_set_crs(4326)  # Set the Coordinate Reference System (CRS) to EPSG:4326 (GDA2020)
```

## Land Mask

### Create an extent for Australia

```{r}
# Define an extent object for Australia using specific longitude (X) and latitude (Y) coordinates
Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)
```

### Krapp2021 Land Mask

```{r}
# Create a land mask for the current time using the 'Krapp2021' dataset from the 'pastclim' package
# This mask will restrict our area of interest to Australia
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") %>%
  crop(vect(Aust_extent)) %>%  # Crop to the Australian extent
  aggregate(fact = 5, fun = "mean", na.rm = TRUE)  # Aggregate for better resolution (factor of 5), removing NAs
```

## Plot conditor Occurrence Data

```{r Plot the data using ggplot2}
# Plot the conditor occurrence points with the current land mask using ggplot2
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_sf(data = conditor, aes(col = time_bp))  # Plot conditor occurrence points, colored by time_bp
```

## Create Climate Time Series Using Krapp2021

```{r create climate time series}
# Create a time series of palaeoclimate reconstructions using the 'Krapp2021' dataset
# Variables used: bio05 (max temp), bio06 (min temp), bio12 (annual precipitation)
climate_vars <- c('bio05','bio06','bio12')
climate_full <- pastclim::region_series(
  bio_variables = climate_vars,  # Specify the variables we are interested in
  data = "Krapp2021",  # Use the Krapp2021 dataset
  crop = vect(Aust_extent)  # Crop the climate data to the extent of Australia
)


```

## Spatial Filtering

```{r spatial filtering}
set.seed(123)
conditor <- thin_by_dist(conditor,
  dist_min = km2m(50),

)
nrow(conditor)
```

## Pseudo absence

Sample Pseudo-Absences and Plot conditor Occurrence with Land Mask

Stratified sampling in in physical space. Environmental space is less suitable for palaeo, because its easier to apply temporal depth to physical space stratification- lat, long, time

```{r pseudo absence }
# Sample pseudo-absences for conditor occurrence points between 300 and 90 years before present
conditor3 <- sample_pseudoabs_time(conditor %>%
                                     filter(time_bp == -300),
                                   n_per_presence = 3, # Generate 3 pseudo-absences per presence point
                                   raster = climate_full[[1]], # Use the first layer of climate data as the reference raster
                                   time_col = "time_bp", # Specify the time column
                                   lubridate_fun = pastclim::ybp2date, # Convert years before present (ybp) to date
                                   method = c("dist_min", km2m(70))) # Ensure pseudo-absences are at least 70 km away




# Plot conditor occurrence with the current land mask
ggplot() +
  geom_spatraster(data = land_mask_current) +  # Add the land mask raster
  scale_fill_terrain_c() +  # Use a terrain color scale
  geom_sf(data = conditor3, aes(col = class))  # Plot conditor pseudo-absence and occurrence points, colored by class
```

## Retrieve Climate Data for This Region

```{r extract climate for region}
# Convert conditor to a data frame and extract spatial coordinates (longitude, latitude)
conditor_df <- conditor3 %>%
  dplyr::bind_cols(sf::st_coordinates(conditor3)) %>%  # Bind spatial coordinates to the data
  mutate(time_bp = date2ybp(time_step)) %>%  # Convert time_step to years before present
  as.data.frame() %>%  # Convert to a data frame
  dplyr::select(-geometry)  # Remove geometry column

# Retrieve climate data for the conditor occurrence points from the climate series
conditor_df <- location_slice_from_region_series(conditor_df,
                                                 region_series = climate_full)  # Use the full climate series for extraction
```

## Add Climate Reconstructions to the sf Object

```{r add climate to sf}
# Add the climate reconstructions (bio05, bio06, bio12) to the conditor sf object and remove the time_step column
conditor4 <- conditor3 %>%
  bind_cols(conditor_df[, c("bio05", "bio06", "bio12")]) %>%  # Bind the climate data to the sf object
  # dplyr::select(-time_step) %>%  # Remove time_step column as it’s not needed for modeling
  filter(!is.na(.$bio05))  # Filter out any rows where bio05 is missing
```

# Fit the model by cross-validation

## Set Up a Recipe for Modeling

```{r make recipe}
# Define a recipe for the model where 'class' is the outcome and other variables are predictors
# Geometry is automatically ignored as a predictor for sf objects
conditor_rec <- recipe(conditor4, formula = class ~ bio05 + bio06 + bio12)  # Define the model formula
conditor_rec  # Check the recipe
```

## Build a Workflow Set for Model Tuning

```{r specify models}
# Create a workflow set of different models for tuning
# We will use glm, gam, random forest, and boosted trees as our models
conditor_models <- workflow_set(
  preproc = list(default = conditor_rec),  # Use the recipe for preprocessing
  models = list(
    glm = sdm_spec_glm(),  # Generalized Linear Model
    gam = sdm_spec_gam(),  # Generalized Additive Model
    rf = sdm_spec_rf(),    # Random Forest with tunable hyperparameters
    gbm = sdm_spec_boost_tree()  # Boosted Tree model with tunable hyperparameters
  ),
  cross = TRUE  # Cross all combinations of preprocessing and models
) %>%
  # Update the workflow model for GAMs with a formula specifying smooth terms for continuous predictors
  update_workflow_model("default_gam",
                        spec = sdm_spec_gam(),
                        formula = class ~ s(bio05, k = 10) + s(bio06, k = 10) + s(bio12, k = 10)) %>%
  # Adjust controls to store information needed for the ensemble model
  option_add(control = control_ensemble_grid())
```

## Spatial Block Cross-Validation for Model Tuning

```{r set up cross validation}
# Set up spatial block cross-validation with 5 folds for tuning and assessing models
set.seed(1005)  # Set seed for reproducibility
conditor_cv <- spatial_block_cv(conditor4, v = 5)  # Create spatial block cross-validation with 5 folds
autoplot(conditor_cv)  # Plot the cross-validation folds
```

## Tune and Assess Models

```{r tune models}
# Use block cross-validation folds to tune and assess models
set.seed(123)  # Set seed for reproducibility
conditor_models <- conditor_models %>%
  workflow_map("tune_grid",  # Tune hyperparameters using grid search
               resamples = conditor_cv,  # Use the cross-validation resamples
               grid = 5,  # Use a grid size of 5 for tuning
               metrics = sdm_metric_set(),  # Use the default set of SDM metrics
               verbose = TRUE)  # Print verbose output
```

## Create an Ensemble Model

```{r ensemble}
# Create an ensemble by selecting the best set of parameters for each model
# The Boyce continuous index will be used as the metric for selecting the best random forest and boosted tree models
conditor_ensemble <- simple_ensemble() %>%
  add_member(conditor_models, metric = "boyce_cont")  # Add members to the ensemble using Boyce index as the metric

autoplot(conditor_ensemble)  # Plot the ensemble model performance
```

# Projecting to other times

## Make Predictions for the Last Glacial Maximum (LGM)

```{r create LGM}
# Create a climate raster for the Last Glacial Maximum (LGM) using the correct layer (21) for 20,000 BP
climate_lgm <- rast(list(climate_full[[1]][[779]], climate_full[[2]][[779]], climate_full[[3]][[779]]))  # Extract 20,000 BP layers
names(climate_lgm) <- c("bio05", "bio06", "bio12")  # Rename the layers to match the climate variables


```

## Predict Using the Ensemble Model for LGM

```{r predict lgm}
# Use the ensemble model to predict species occurrence probabilities for the Last Glacial Maximum (LGM)
prediction_lgm <- predict_raster(conditor_ensemble, climate_lgm)  # Predict LGM probabilities

# Plot the predicted probabilities using ggplot2
ggplot() +
  geom_spatraster(data = prediction_lgm, aes(fill = mean)) +  # Add the prediction raster to the plot
  scale_fill_terrain_c(name = "Probability\nof occurrence")  # Use a terrain color scale to show probabilities
```

## save model as spatraster

```{r save_model_spatraster}

# Initialize an empty SpatRaster stack
conditor_spatrast <- NULL

# Loop through 40 time steps (40ka in 1ka timesteps) to create and stack rasters
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present for each timestep
  
  # Create a climate snapshot for the current timestep
  climate_snapshot <- rast(list(
    climate_full[[1]][[timestep]], 
    climate_full[[2]][[timestep]], 
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename layers to match the climate variables

  # Predict species occurrence for the current climate snapshot
  prediction_raster <- predict_raster(conditor_ensemble, climate_snapshot)
  
  # Stack the prediction raster along the z-axis
  conditor_spatrast <<- if (is.null(conditor_spatrast)) {
    prediction_raster
  } else {
    c(conditor_spatrast, prediction_raster)
  }
})

# Assign time steps to the z-axis of the stacked raster
time(conditor_spatrast) <- time(climate_full[[1]])[760:800] - 1950

# Save the stacked SpatRaster to a file
output_file <- "C:/github/SNR_SDM/Results/conditor Krapp model/predictions_stack.tif"
writeRaster(conditor_spatrast, filename = output_file, overwrite = TRUE)
```

## Create a Stack of Predictions and Plot for All Time Steps

```{r stack o maps}
# Loop over 40 time steps (40ka in 1ka timesteps) to create and save a map of predictions for each time step
walk(760:800, function(timestep) {
  year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present for each timestep
  
  # Create a climate snapshot for the current timestep
  climate_snapshot <- rast(list(climate_full[[1]][[timestep]], 
                                climate_full[[2]][[timestep]], 
                                climate_full[[3]][[timestep]]))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")  # Rename layers to match the climate variables

  # Predict species occurrence for the current climate snapshot
  prediction1 <- predict_raster(conditor_ensemble, climate_snapshot)

  # Plot the predictions using ggplot2
  p <- ggplot() +
    geom_spatraster(data = prediction1, aes(fill = mean)) +  # Add the prediction raster
    scale_fill_terrain_c(name = "Probability\nof occurrence") +  # Use a terrain color scale
    ggtitle(paste0("Krapp 70km conditor Years BP: ", year_bp))  # Add a title with the year before present

  # Save the plot to a folder with filenames indicating the year BP
  ggsave(filename = paste0("C:/github/SNR_SDM/Results/conditor Krapp model/stratified-random sampling/70km/conditor_map_", 
                         year_bp, " Krapp70km", ".png"),
       plot = p,
       width = 7, height = 7)

})
```

# **Model ends here.**

# Additional steps for validation and analysis

## Load midden data

```{r load midden data set}
#Midden data set is used for additional validation

# Load the independent dataset
dated_middens <- read.csv("C:/github/SNR_SDM/Data/processed/datedmiddens.csv")

# Convert to sf object (assuming it contains longitude and latitude columns)
middens_sf <- dated_middens %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)  # Adjust column names if different

# Reproject to match the CRS of the conditor_spatrast
middens_sf <- st_transform(middens_sf, crs = crs(conditor_spatrast))

# Convert sf object to SpatVector for compatibility with terra::extract
middens_vect <- terra::vect(middens_sf)

# Extract predicted values from the raster stack for presence locations
extracted_values <- terra::extract(conditor_spatrast, middens_vect)

# Bind the extracted predicted values back to the original sf object
middens_sf$predicted <- extracted_values[, 2]  # Assuming the second column contains predictions

# Extract timesteps from the raster stack
raster_timesteps <- time(conditor_spatrast)

# Round time_bp to the nearest raster timestep
middens_sf <- middens_sf %>%
  mutate(
    binned_time_bp = sapply(time_bp, function(x) {
      raster_timesteps[which.min(abs(raster_timesteps - x))]
    }),
    lubridate_time = ybp2date(time_bp)
  )

# Create a data frame with raster timesteps
timesteps_df <- data.frame(
  raster_timestep = raster_timesteps  # Timesteps from the raster
)

# Count occurrences per timestep
organized_data <- timesteps_df %>%
  left_join(
    middens_sf %>%
      group_by(binned_time_bp) %>%
      summarise(
        count = n(),                        # Number of occurrences
        mean_value = mean(predicted, na.rm = TRUE)  # Mean predicted suitability (example)
      ),
    by = c("raster_timestep" = "binned_time_bp")
  )

# Replace NAs with 0 for counts (if no data in some bins)
organized_data <- organized_data %>%
  mutate(
    count = replace_na(count, 0),
    mean_value = replace_na(mean_value, NA)  # Leave NA for mean_value where no data exists
  )

# View the organized data
print(organized_data)



```

# True skills statistic using midden data

This step is an additional validation step using the midden data set.

The previous output is a continuous suitability gradient. for TSS to work we need to define suitability binary, so an area is classed as either suitable or unsuitable. to do this we need to define a threshold - above which is suitable and below is unsuitable. 50% is standard.

```{r true skills statistic using midden data}

threshold <- 0.5 # we are defining the binary suitability threshold at 50%. 

conditor_binary <- app(x = conditor_spatrast, fun = \(i) i > threshold) #making the binary

time(conditor_binary) <- time(conditor_spatrast) |> pastclim::ybp2date()
conditor_rs <- sds(conditor_binary)
names(conditor_rs) <- "pres_abs"  
pseudoabs <- sample_pseudoabs_time(
  data = dated_middens,
  raster = conditor_rs,
  n_per_presence = 3, # 3 is standard absence per presence
  time_col = "time_bp",
  lubridate_fun = pastclim::ybp2date,
  method = c("dist_min", km2m(200)),
  time_buffer = 3650000/2
) 

# Plot the pseudo absence + presence points with the current land mask using ggplot2
ggplot() +
  geom_spatraster(data = land_mask_current, aes()) +  # Add the land mask as a spatial raster layer
  scale_fill_terrain_c() +  # Use a terrain color scale for the land mask
  geom_point(data = pseudoabs, aes(col = time_step, x = longitude, y = latitude)) + 
  geom_point(data = dated_middens, aes(x = longitude, y = latitude), color = "red")
# Create data frame in correct format
pres_abs_observed <- pseudoabs |> mutate(time_bp = date2ybp(time_step)) |>
  select(-class, -time_step) |> rowid_to_column("name")
#extract presence and absence points from SDM 
#location_slice_from_region_series comes from function on my computer, not from package
pres_abs_pred <- location_slice_from_region_series(
  x = pres_abs_observed,
  region_series = conditor_rs
)
pres_abs_pred$class <- pseudoabs$class


# Confusion matrix
a <- nrow(filter(pres_abs_pred, class == "presence", pres_abs == 1))
b <- nrow(filter(pres_abs_pred, class == "pseudoabs", pres_abs == 1))
c <- nrow(filter(pres_abs_pred, class == "presence", pres_abs == 0))
d <- nrow(filter(pres_abs_pred, class == "pseudoabs", pres_abs == 0))
tpr <- a/(a+c)
tnr <- d/(b+d)
tss <- tpr + tnr - 1
sorensen <- 2*(a/(c+(2*a)+d))
f2score <- 2*(((a/(a+b))*tpr)/((a/(a+b))+tpr)) 
```

# Partial Dependence Profiles and Variable Importance

```{r PDP}
# PDP for bio12
# Explain the ensemble model to generate partial dependence profiles and variable importance
explainer_conditor_ens <- explain_tidysdm(conditor_ensemble)  # Create an explainer object for the ensemble model

# Compute variable importance using the DALEX package
vip_ensemble <- model_parts(explainer = explainer_conditor_ens)  # Calculate variable importance
plot(vip_ensemble)  # Plot the variable importance results

# Generate a partial dependence profile for bio12
pdp_bio12 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio12")  # Generate PDP for bio12
plot(pdp_bio12)  # Plot the partial dependence profile for bio12


#PDP for bio05

# Generate and plot a partial dependence profile for bio05
pdp_bio05 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio05")
plot(pdp_bio05)


#$ PDP for bio06
 
# Generate and plot a partial dependence profile for bio06
pdp_bio06 <- model_profile(explainer_conditor_ens, N = 500, variables = "bio06")
plot(pdp_bio06)
```

# Climate variable maps

Previously we made a stack of maps that shows the model predictions at each time step. This step creates 3 stacks of maps, one for each climate variable, at the same time steps as we used for the distribution maps. This helps us to understand the variables driving the changes in distribution.

```{r Climate variable maps}
# Create a Stack of Maps for Each Climate Variable
walk2(c("bio05", "bio06", "bio12"), 1:3, ~{
  var_name <- .x
  var_idx <- .y
  
  # Create a folder for the variable if it doesn't exist
  dir.create(paste0("C:/github/SNR_SDM/Results/Climate_Maps/", var_name), showWarnings = FALSE)

  # Loop over 41 time steps (40ka in 1ka timesteps starting at 0)
  walk(760:800, function(timestep) {
    year_bp <- time(climate_full[[1]])[timestep] - 1950  # Calculate years before present
    
    # Extract climate data for the variable and timestep
    climate_snapshot <- climate_full[[var_idx]][[timestep]]
    names(climate_snapshot) <- var_name  # Assign name to layer
    
    # Plot the map
    p <- ggplot() +
      geom_spatraster(data = climate_snapshot, aes(fill = !!sym(var_name))) +  # Map variable
      scale_fill_terrain_c(name = var_name) +  # Add terrain scale
      ggtitle(paste0("Krapp Climate Variable: ", var_name, " - Years BP: ", year_bp))
    
    # Save the plot
    ggsave(filename = paste0("C:/github/SNR_SDM/Results/Climate_Maps/", var_name, "/",
                             var_name, "_map_", year_bp, ".png"),
           plot = p,
           width = 7, height = 7)
  })
})

```
