---
title: "Krapp apicalis finished 2 (1° version)"
author: "Annie"
date: "2025-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)  # set knitr defaults
set.seed(42)                                                          # reproducibility
```

# Install packages (optional)
```{r install-packages, eval=FALSE}
# Convenience installer; prefer renv in practice.                                      # explain purpose
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")        # ensure devtools

packages <- c(                                                                          # list packages to check/install
  "DALEX", "RefManageR", "USE", "broom", "car", "cowplot", "devtools", "dials",
  "dplyr", "ecospat", "fs", "ggplot2", "grid", "gtools", "here", "infer",
  "knitr", "magick", "modeldata", "oz", "parsnip", "pastclim", "patchwork",
  "purrr", "readr", "recipes", "rnaturalearth", "rnaturalearthdata", "rsample",
  "rworldmap", "rworldxtra", "scales", "sf", "spatialsample", "terra", "tibble",
  "tidymodels", "tidysdm", "tidyterra", "tidyverse", "tidyr", "tune", "viridis",
  "viridisLite", "workflowsets", "workflows", "yardstick"
)

for (pkg in packages) {                                                                  # loop over packages
  if (!requireNamespace(pkg, quietly = TRUE)) {                                          # if missing, install
    if (pkg == "pastclim") {
      devtools::install_local("C:/github/pastclim")                                      # local install for pastclim
    } else if (pkg == "USE") {
      devtools::install_github("danddr/USE")                                             # install USE from GitHub
    } else {
      install.packages(pkg)                                                              # CRAN install
    }
  }
}
```

# Libraries
```{r libraries}
suppressPackageStartupMessages({                                              # quiet library load
  library(DALEX); library(RefManageR); library(USE)
  library(broom);  library(car);       library(cowplot)
  library(dials);  library(dplyr);     library(ecospat)
  library(fs);     library(ggplot2);   library(grid)
  library(gtools); library(here);      library(infer)
  library(knitr);  library(magick);    library(modeldata)
  library(oz);     library(parsnip);   library(pastclim)
  library(patchwork); library(purrr);  library(readr)
  library(recipes); library(rnaturalearth); library(rnaturalearthdata)
  library(rsample); library(rworldmap); library(rworldxtra)
  library(scales);  library(sf);       library(spatialsample)
  library(terra);   library(tibble);   library(tidymodels)
  library(tidysdm); library(tidyterra); library(tidyverse)
  library(tidyr);   library(tune);     library(viridis)
  library(viridisLite); library(workflowsets); library(workflows)
  library(yardstick)
})
```

# Project paths
```{r paths}
here::i_am("Analyses/Krapp apicalis finished 2 - 1 degree version.Rmd")                       # anchor this file for here()

results_dir   <- here("Results/stratified environmental sampling", "apicalis")  # results out dir
processed_dir <- here("Data", "processed")                                     # processed data dir
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)        # create results dir if needed
```

# Globals (shared objects)
```{r globals}
world_map <- rworldmap::getMap(resolution = "high")                             # world polygons
australia_outline <- sf::st_as_sf(world_map) |> dplyr::filter(ADMIN == "Australia")  # Australia polygon

Aust_extent <- terra::ext(110, 152.5, -42.5, -7.5)                              # modelling extent (lon/lat)
xlim <- c(112, 150); ylim <- c(-40, -10)                                       # plot window

theme_map_clean <- theme_minimal(base_size = 12) +                              # reusable clean map theme
  theme(
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_blank()
  )
```

# Data: download Krapp2021
```{r download-climate}
pastclim::download_dataset("Krapp2021")                                        # ensure Krapp2021 is available
```

# Read *apicalis* occurrences
```{r read-apicalis}
apicalis <- readr::read_csv(file.path(processed_dir, "apicalis_flat.csv"),     # load occurrences
                            show_col_types = FALSE) |>
  (\(x) { print(x); x })() |>                                                  # print head/sanity then pass through
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)                # cast to sf in WGS84
```

# Land mask (0 ka) at 1°
```{r land-mask}
land_mask_current <- pastclim::get_land_mask(time_bp = 0, dataset = "Krapp2021") |>  # 0 ka land mask
  terra::crop(terra::vect(Aust_extent)) |>                                           # crop to Australia extent
  terra::aggregate(fact = 2, fun = "mean", na.rm = TRUE)                             # **1°** aggregate (0.5° * 2)
```

# Quick map of occurrences
```{r plot-occurrences}
ggplot() +
  tidyterra::geom_spatraster(data = land_mask_current, aes()) +              # draw land mask background
  tidyterra::scale_fill_terrain_c() +                                        # terrain fill
  geom_sf(data = apicalis, aes(col = time_bp)) +                             # plot occurrences coloured by time
  coord_sf(xlim = xlim, ylim = ylim)                                         # map window
```

# Climate time series (Krapp2021) → aggregated to 1°
```{r climate-series}
climate_vars <- c("bio05","bio06","bio12")                          # variables to extract

# Pull native 0.5° region series as a SpatRasterDataset (sds)
climate_full_raw <- pastclim::region_series(
  bio_variables = climate_vars,
  data = "Krapp2021",
  crop = terra::vect(Aust_extent)
)

# Aggregate each SpatRaster inside the sds to 1° and rewrap as sds
climate_full <- terra::sds(
  lapply(terra::as.list(climate_full_raw), function(r) {
    terra::aggregate(r, fact = 2, fun = mean, na.rm = TRUE)          # 0.5° * 2 => 1°
  })
)

# --- Restore BP time axis on the aggregated SpatRasterDataset ---
# Get the BP vector from the original (pre-aggregation) region series
bp_vec <- pastclim::time_bp(climate_full_raw)      # years BP (e.g., 0, -1000, ...)

# Try to set at the dataset level; if that method isn't available, set per layer
ok <- TRUE
try({
  pastclim::time_bp(climate_full) <- bp_vec        # preferred: set once on the sds
}, silent = TRUE) -> res

if (inherits(res, "try-error")) {
  ok <- FALSE
}

if (!ok) {
  # Fallback: set on each SpatRaster in the sds
  for (i in seq_along(terra::as.list(climate_full))) {
    pastclim::time_bp(climate_full[[i]]) <- bp_vec
  }
}

# Quick check
stopifnot(identical(pastclim::time_bp(climate_full[[1]]), bp_vec))

# --- Now the original extract should work ---
apicalis_df <- location_slice_from_region_series(
  apicalis_df,
  region_series = climate_full
)

# Quick sanity checks (safe)
stopifnot(inherits(climate_full, "SpatRasterDataset"))               # still an sds
stopifnot(all(vapply(terra::as.list(climate_full),                   # all members are SpatRaster
                     function(x) inherits(x, "SpatRaster"), logical(1))))
# Optional: confirm spatial resolution is ~1° (lon)
print(terra::res(climate_full[[1]]))                                 # should be close to c(1, 1)
```

# Spatial filtering
```{r spatial-filter}
set.seed(123)                                                                 # reproducibility
apicalis <- thin_by_dist(apicalis, dist_min = km2m(50))                       # thin presences by 50 km
NROW(apicalis)                                                                # report retained n
```

# Pseudo-absence sampling in environmental space
```{r env-space-setup}
# 1950 slice (index ~800 ≈ 0 BP in Krapp timeline) assembled into a 3-layer SpatRaster at 1°.
env_space_1950 <- c(climate_full[[1]][[800]], climate_full[[2]][[800]], climate_full[[3]][[800]])  # 1° env slice
```

```{r pseudo-abs-env}
library(scales)                                                               # alpha() for fills
apicalis3 <- paSampling(                                                      # environmental-space pseudoabsences
  env.rast = env_space_1950,                                                  # 1° climate slice
  pres     = apicalis %>% dplyr::filter(time_bp == -300),                     # presence subset (example time)
  thres    = 0.75,                                                            # occupancy threshold in env-space grid
  n.tr     = 9,                                                               # (method parameter)
  grid.res = 9                                                                # (method parameter)
) |>
  sf::st_drop_geometry()                                                      # drop geometry for tabular ops

# Map presences vs pseudoabsences
pseudoabsence_map <- ggplot() +
  geom_sf(data = australia_outline, fill = alpha("#C4C4C4", 0.5), color = "grey50") +  # base map
  geom_point(data = apicalis3,                                                # pseudoabsence points
             aes(x, y, shape = "Pseudo Absence", fill = "Pseudo Absence"),
             color = "black", size = 2.5, stroke = 0.3) +
  geom_sf(data = apicalis |> dplyr::filter(time_bp == -300),                  # presence points
          aes(fill = "Presence", shape = "Presence"),
          color = "black", size = 2.5, stroke = 0.3) +
  scale_fill_manual(values = c("Presence" = "#A100FA", "Pseudo Absence" = "#FF7E1D"), name = "Legend") +
  scale_shape_manual(values = c("Presence" = 24, "Pseudo Absence" = 21), name = "Legend") +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +          # map window
  scale_x_continuous(breaks = seq(110, 155, by = 5)) +                        # x ticks
  scale_y_continuous(breaks = seq(-45, -10, by = 5)) +                        # y ticks
  labs(title = "Pseudo Absences in Environmental Space", x = "Longitude", y = "Latitude") +  # labels
  theme_minimal(base_size = 12) +                                             # theme
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "#C4C4C4", linewidth = 0.3),
    panel.border     = element_rect(color = "#C4C4C4", fill = NA, linewidth = 0.3)
  )

ggsave(file.path(results_dir, "pseudo_absences_envspace.png"), pseudoabsence_map,  # save map
       width = 6, height = 5, dpi = 300, bg = "white")
```

# Extract climate for points
```{r extract-climate}
apicalis_df <- apicalis |>
  dplyr::bind_cols(sf::st_coordinates(apicalis)) |>                           # add XY columns
  dplyr::rename(x = X, y = Y) |>                                              # rename coords
  dplyr::mutate(percP = "pres") |>                                            # mark presences
  dplyr::bind_rows(apicalis3) |>                                              # append pseudoabsences
  dplyr::mutate(time_bp = -300) |>                                            # set time slice for extraction
  as.data.frame() |>                                                          # ensure data.frame
  dplyr::select(-geometry)                                                    # drop geometry column

apicalis_df <- location_slice_from_region_series(apicalis_df,                 # extract env vals at 1°
                                                 region_series = climate_full)
```

# Add climate back and cast to sf
```{r add-climate-sf}
apicalis4 <- apicalis |>
  dplyr::bind_cols(sf::st_coordinates(apicalis)) |>                           # add coords
  dplyr::rename(x = X, y = Y) |>                                              # rename coords
  dplyr::mutate(percP = "pres") |>                                            # label presences
  dplyr::bind_rows(apicalis3) |>                                              # add pseudoabsences
  dplyr::mutate(time_bp = -300) |>                                            # time slice
  as.data.frame() |>                                                          # data.frame
  dplyr::select(-geometry) |>                                                 # drop geometry
  dplyr::bind_cols(apicalis_df[, c("bio05", "bio06", "bio12")]) |>            # bind extracted env variables
  dplyr::filter(!is.na(.data$bio05)) |>                                       # remove NAs
  dplyr::mutate(percP = dplyr::if_else(percP == "pabs", "pseudoabsence", "presence")) |>  # tidy class labels
  sf::st_as_sf(coords = c("x", "y"), crs = 4326)                              # recast to sf
```

# Recipe
```{r recipe}
apicalis_rec <- recipes::recipe(apicalis4, formula = percP ~ bio05 + bio06 + bio12)  # define recipe
apicalis_rec                                                                        # print recipe
```

# Models & workflows
```{r models}
apicalis_models <- workflowsets::workflow_set(                                      # build model set
  preproc = list(default = apicalis_rec),                                            # shared recipe
  models  = list(
    glm = sdm_spec_glm(),                                                            # GLM spec
    gam = sdm_spec_gam(),                                                            # GAM spec
    rf  = sdm_spec_rf(),                                                             # RF spec
    gbm = sdm_spec_boost_tree()                                                      # GBM spec
  ),
  cross = TRUE                                                                       # all combinations
) |>
  update_workflow_model(                                                             # supply GAM formula
    "default_gam",
    spec = sdm_spec_gam(),
    formula = percP ~ s(bio05, k = 10) + s(bio06, k = 10) + s(bio12, k = 10)
  ) |>
  option_add(control = control_ensemble_grid())                                      # tuning controls
```

# Spatial block CV
```{r cv}
set.seed(1005)                                                                       # reproducibility
apicalis_cv <- spatial_block_cv(apicalis4, v = 5)                                    # spatial CV folds
autoplot(apicalis_cv)                                                                # visualise blocks
```

# Tune
```{r tune}
set.seed(123)                                                                        # reproducibility
apicalis_models <- apicalis_models |>
  workflow_map("tune_grid",                                                          # tune grids
               resamples = apicalis_cv,                                              # CV object
               grid = 5,                                                             # grid size
               metrics = sdm_metric_set(),                                           # SDM metrics
               verbose = TRUE)                                                       # verbose progress
```

# Build ensemble & inspect choices
```{r ensemble}
apicalis_ensemble <- simple_ensemble() |>                                            # init ensemble
  add_member(apicalis_models, metric = "tss_max")                                    # add best per model

chosen <- apicalis_models |>                                                         # collect best settings
  dplyr::mutate(
    best = purrr::map(result, ~ tune::select_best(.x, metric = "tss_max")),          # best params
    tss  = purrr::map_dbl(result, ~ tune::show_best(.x, metric = "tss_max", n = 1)$mean), # best TSS
    best = purrr::map_chr(best, ~ paste(names(.x), .x, sep = "=", collapse = "; "))  # as string
  ) |>
  dplyr::transmute(wflow_id, best, tss)                                              # tidy cols
print(chosen)                                                                        # print summary

# Fold-level performance boxplots
workflow_labels <- c(                                                                # label mapping
  default_gam = "Generalised Additive Model",
  default_gbm = "Boosted Tree",
  default_glm = "Generalised Linear Model",
  default_rf  = "Random Forest"
)
model_order <- c("Generalised Additive Model", "Boosted Tree", "Generalised Linear Model", "Random Forest")   # factor order
metric_labels <- c(boyce_cont = "Boyce Index", roc_auc = "AUC", tss_max = "TSS")     # facet labels
color_map <- c(                                                                       # fill colours
  "Generalised Additive Model" = "#FF61C3",
  "Boosted Tree"               = "#00B140",
  "Generalised Linear Model"   = "#00BFFF",
  "Random Forest"              = "#BC43FF"
)

perf_data <- apicalis_models |>                                                       # gather fold-level metrics
  dplyr::mutate(metrics = purrr::map(result, collect_metrics, summarize = FALSE)) |>
  dplyr::select(wflow_id, metrics) |>
  tidyr::unnest(metrics) |>
  dplyr::filter(.metric %in% names(metric_labels)) |>
  dplyr::mutate(
    model        = factor(workflow_labels[wflow_id], levels = model_order),           # map model names
    metric_label = factor(metric_labels[.metric], levels = c("Boyce Index", "AUC", "TSS"))  # facet factor
  )

perf_plot <- ggplot(perf_data, aes(x = model, y = .estimate, fill = model)) +         # build plot
  geom_boxplot(width = 0.6, alpha = 0.9, outlier.shape = NA) +
  geom_jitter(aes(color = model), size = 1.3, width = 0.15, alpha = 0.4, stroke = 0.3, shape = 21) +
  scale_fill_manual(values = color_map, name = "Model Type") +
  scale_color_manual(values = rep("black", length(model_order)), guide = "none") +
  facet_wrap(~ metric_label, scales = "free_y") +
  labs(title = "Model performance across algorithms (fold-level)", x = NULL, y = "Performance metric") +
  theme_minimal(base_size = 12) +
  theme(
    panel.background  = element_rect(fill = "#EEEEEE", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.grid.major  = element_line(color = "white"),
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_blank(),
    axis.ticks.x      = element_blank(),
    legend.position   = "right",
    legend.direction  = "vertical",
    legend.title      = element_text(size = 11),
    legend.text       = element_text(size = 10),
    plot.title        = element_text(face = "bold", size = 13),
    strip.text        = element_text(margin = margin(b = 5))
  )

dir.create(file.path(results_dir, "ensemble"), showWarnings = FALSE, recursive = TRUE)   # ensure out dir
ggsave(file.path(results_dir, "ensemble", "ensemble_performance_clean.png"), perf_plot,  # save figure
       width = 8, height = 5, dpi = 300)
perf_plot                                                                             # print plot
```

# LGM snapshot
```{r lgm}
climate_lgm <- terra::rast(list(climate_full[[1]][[779]], climate_full[[2]][[779]], climate_full[[3]][[779]]))  # 1° LGM layers
names(climate_lgm) <- c("bio05", "bio06", "bio12")                                                               # name bands

prediction_lgm <- predict_raster(apicalis_ensemble, climate_lgm)                                                 # predict at LGM

ggplot() +
  tidyterra::geom_spatraster(data = prediction_lgm, aes(fill = mean)) +                                          # plot mean layer
  tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
  coord_sf(xlim = xlim, ylim = ylim)                                                                             # map window
```

# Save model as SpatRaster stack (1° through time)
```{r save-stack}
apicalis_spatrast <- NULL                                                                 # init empty stack
purrr::walk(760:800, function(timestep) {                                                 # iterate time steps
  year_bp <- time(climate_full[[1]])[timestep] - 1950                                     # convert to y BP
  climate_snapshot <- terra::rast(list(                                                   # snapshot at 1°
    climate_full[[1]][[timestep]],
    climate_full[[2]][[timestep]],
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")                                 # name bands
  prediction_raster <- predict_raster(apicalis_ensemble, climate_snapshot)                # predict suitability
  apicalis_spatrast <<- if (is.null(apicalis_spatrast)) prediction_raster else c(apicalis_spatrast, prediction_raster)  # append
})

time(apicalis_spatrast) <- time(climate_full[[1]])[760:800] - 1950                        # attach time axis (y BP)

output_dir <- here("Results", "apicalis Krapp model")                                     # output dir
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)                     # ensure exists
writeRaster(apicalis_spatrast, filename = file.path(output_dir, "predictions_stack.tif"), # write GeoTIFF stack
            overwrite = TRUE)
```

# Stack o' maps
```{r stack-o-maps, dependson="globals"}
suppressPackageStartupMessages({ library(magick) })                                       # image ops
stopifnot(exists("australia_outline"), exists("results_dir"), exists("climate_full"), exists("apicalis_ensemble"))  # sanity checks

out_dir <- file.path(results_dir, "model"); if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)  # ensure dir

purrr::walk(760:800, function(timestep) {                                                # loop time steps
  year_bp <- time(climate_full[[1]])[timestep] - 1950                                    # year BP
  climate_snapshot <- terra::rast(list(                                                  # 1° snapshot
    climate_full[[1]][[timestep]],
    climate_full[[2]][[timestep]],
    climate_full[[3]][[timestep]]
  ))
  names(climate_snapshot) <- c("bio05", "bio06", "bio12")                                # band names
  prediction_raster <- predict_raster(apicalis_ensemble, climate_snapshot)               # predict

  p <- ggplot() +
    tidyterra::geom_spatraster(data = prediction_raster, aes(fill = mean)) +             # map prediction
    tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +    # outline
    coord_sf(xlim = xlim, ylim = ylim) +                                                 # window
    ggtitle(bquote(italic("Leporillus apicalis") * " — Years BP: " * .(year_bp))) +      # title
    theme_minimal() +
    theme(
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#F2F2F2", color = NA),
      panel.grid.major = element_line(color = "#C4C4C4"),
      panel.grid.minor = element_line(color = "#C4C4C4"),
      plot.margin      = margin(0, 0, 0, 0, unit = "pt")
    )

  plot_path <- file.path(out_dir, paste0("apicalis_map_", year_bp, ".png"))              # path
  ggsave(plot_path, p, width = 7, height = 7, dpi = 300, bg = "transparent")             # save png

  img <- image_read(plot_path)                                                           # read image
  img <- image_border(image_trim(img), color = "white", geometry = "30x30")              # trim + pad
  image_write(img, path = plot_path)                                                     # overwrite
})
```

# 6-image panel (model only)
```{r six-panel}
raster_times  <- terra::time(apicalis_spatrast)                                          # time vector
target_times  <- c(-40000, -26000, -21000, -12000, -6000, 0)                             # chosen years
matched_idx   <- match(target_times, raster_times)                                       # indices
panel_tags    <- letters[seq_along(target_times)]                                        # panel letters

panel_plots <- pmap(                                                                     # build small multiples
  list(idx = matched_idx, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- apicalis_spatrast[[idx]]                                       # select layer
    ggplot() +
      tidyterra::geom_spatraster(data = current_prediction, aes(fill = mean)) +          # plot raster
      tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.3) +  # outline
      coord_sf(xlim = xlim, ylim = ylim) +                                               # window
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +                               # y ticks
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +                               # x ticks
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(year_bp))) +               # title
      theme_minimal(base_size = 10) +
      theme(
        panel.background = element_rect(fill = "#DCF3FC", color = NA),
        plot.background  = element_rect(fill = "white", color = NA),
        panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_blank(),
        axis.text        = element_text(size = 8),
        axis.title       = element_blank(),
        plot.margin      = margin(2, 2, 2, 2),
        legend.position  = "right",
        legend.title     = element_text(size = 11),
        legend.text      = element_text(size = 11)
      )
  }
)

final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +                   # arrange panels
  plot_annotation(title = expression(italic("Leporillus apicalis") * " SDM model predictions"),
                  theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)))

ggsave(file.path(results_dir, "apicalis_model_panel.png"), final_panel, width = 10, height = 9, dpi = 300)  # save panel
```

# Centroid through time & diff map (0 vs 20 ka)
```{r centroid-diff, dependson="globals"}
# Compute weighted centroids per timestep
apicalis_centroids <- lapply(1:nlyr(apicalis_spatrast), function(i) {         # loop layers
  r <- apicalis_spatrast[[i]]                                                 # layer i
  vals <- terra::values(r)[, 1]                                               # suitability vector
  coords <- terra::crds(r, na.rm = FALSE)                                     # cell centroids
  valid <- !is.na(vals) & vals > 0                                            # mask invalid/zero
  if (!any(valid)) return(NULL)                                               # skip empty
  w <- vals[valid]; cxy <- coords[valid, ]                                    # weights and coords
  data.frame(time_bp = time(apicalis_spatrast)[i],                            # time stamp
             longitude = sum(cxy[,1] * w) / sum(w),                           # weighted lon
             latitude  = sum(cxy[,2] * w) / sum(w))                           # weighted lat
}) |>
  dplyr::bind_rows()                                                          # bind rows

# Difference map present - 20 ka
r_times <- as.numeric(terra::time(apicalis_spatrast))                          # numeric times
idx_20  <- which.min(abs(r_times - (-20000)))                                  # index near 20 ka
idx_0   <- which.min(abs(r_times - 0))                                         # index near 0 ka
r_diff  <- apicalis_spatrast[[idx_0]] - apicalis_spatrast[[idx_20]]            # present minus 20 ka
names(r_diff) <- "diff"                                                        # name layer to avoid aes(fill=mean) issues

vals <- terra::values(r_diff, mat = FALSE)                                     # diff values
lims <- quantile(abs(vals), 0.98, na.rm = TRUE) |> as.numeric()                # robust max for scale
lims <- max(0.1, min(1, lims))                                                 # clamp limits

fill_scale <- scale_fill_gradient2(                                            # diverging scale
  name = "Change in suitability\nsince 20 ka",
  low = "#3B4CC0", mid = "white", high = "#B40426",
  midpoint = 0, limits = c(-lims, lims)
)

pt_20 <- apicalis_centroids[which.min(abs(apicalis_centroids$time_bp + 20000)), , drop = FALSE]  # 20 ka centroid
pt_0  <- apicalis_centroids[which.min(abs(apicalis_centroids$time_bp - 0)),       , drop = FALSE]  # 0 ka centroid

p_single <- ggplot() +
  tidyterra::geom_spatraster(data = r_diff, aes(fill = diff)) +               # plot diff layer by its name
  fill_scale +                                                                # add colour scale
  geom_sf(data = australia_outline, fill = NA, color = "grey20", linewidth = 0.4) +  # outline
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +                        # window
  geom_path(data = apicalis_centroids, aes(longitude, latitude), color = "black", linewidth = 0.7, alpha = 0.85) +  # path
  geom_point(data = apicalis_centroids, aes(longitude, latitude, color = time_bp), size = 2) +                      # time points
  geom_point(data = pt_20, aes(longitude, latitude, shape = "20 ka centroid"), size = 3, fill = "white", color = "black", stroke = 0.6) +  # 20 ka mark
  geom_point(data = pt_0,  aes(longitude, latitude, shape = "Present centroid"), size = 3, fill = "yellow", color = "black", stroke = 0.6) + # 0 ka mark
  scale_color_viridis_c(name = "Years BP", option = "H",
                        limits = c(-40000, 0), breaks = c(-40000, -30000, -20000, -10000, 0),
                        labels = c("40 ka", "30 ka", "20 ka", "10 ka", "0")) +                                     # colourbar for time
  scale_shape_manual(name = "Centroid positions", values = c("20 ka centroid" = 24, "Present centroid" = 21)) +    # shapes
  labs(x = "Longitude", y = "Latitude") +                                     # axis labels
  theme_map_clean +                                                           # map theme
  guides(fill = guide_colorbar(title.position = "top", order = 1),            # tidy guides
         color = guide_colorbar(title.position = "top", order = 2),
         shape  = guide_legend(title.position  = "top", order = 3))

ggsave(file.path(results_dir, "figure_range_shift_single.png"), p_single, width = 7.5, height = 6.5, dpi = 300, bg = "white")  # save
```

# Middens validation prep
```{r middens-load}
dated_middens <- read.csv(file.path(processed_dir, "datedmiddens.csv"))                 # load middens table

middens_sf <- dated_middens |>
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>                      # to sf
  sf::st_transform(crs = terra::crs(apicalis_spatrast))                                  # match raster crs

middens_vect <- terra::vect(middens_sf)                                                  # to terra vector
rt <- as.numeric(terra::time(apicalis_spatrast))                                         # raster times (y BP)

middens_sf <- middens_sf |>
  dplyr::mutate(
    binned_time_bp = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),            # bin to nearest raster time
    nearest_idx    = match(binned_time_bp, rt),                                          # index of that bin
    lubridate_time = pastclim::ybp2date(time_bp)                                         # calendar date for ref
  )

extracted_values <- terra::extract(apicalis_spatrast, middens_vect, ID = FALSE)          # extract suitability
vals_mat <- as.matrix(extracted_values)                                                   # to matrix
middens_sf$predicted <- vals_mat[cbind(seq_len(nrow(middens_sf)), middens_sf$nearest_idx)]  # assign predicted value

raster_timesteps <- rt                                                                   # alias
timesteps_df <- data.frame(raster_timestep = raster_timesteps)                           # for join

organized_data <- timesteps_df |>                                                        # summarise counts/means
  dplyr::left_join(
    middens_sf |>
      dplyr::group_by(binned_time_bp) |>
      dplyr::summarise(count = dplyr::n(), mean_value = mean(predicted, na.rm = TRUE), .groups = "drop"),
    by = c("raster_timestep" = "binned_time_bp")
  ) |>
  dplyr::mutate(count = tidyr::replace_na(count, 0),
                mean_value = tidyr::replace_na(mean_value, NA_real_))
print(organized_data)                                                                     # print summary
```

# Middens-overlaid maps (loop)
```{r middens-maps, dependson="globals"}
suppressPackageStartupMessages({ library(magick) })                                      # image ops
world_map <- rworldmap::getMap(resolution = "high")                                      # world map
australia_outline <- sf::st_as_sf(world_map) %>% dplyr::filter(ADMIN == "Australia")     # outline

raster_timesteps <- time(apicalis_spatrast)                                              # time vector
walk(seq_along(raster_timesteps), function(i) {                                          # iterate timesteps
  current_year_bp <- raster_timesteps[i]                                                 # year BP
  current_middens <- middens_sf %>% dplyr::filter(abs(time_bp - current_year_bp) <= 500) # ±500y middens
  current_prediction <- apicalis_spatrast[[i]]                                           # suitability raster

  p <- ggplot() +
    tidyterra::geom_spatraster(data = current_prediction, aes(fill = mean)) +            # draw suitability
    tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence") +
    geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.6) +     # outline
    geom_sf(data = current_middens, shape = 21, fill = "#FF3399", color = "black", size = 2, stroke = 0.4) +  # middens
    coord_sf(xlim = xlim, ylim = ylim) +                                                 # window
    ggtitle(bquote(italic("Leporillus apicalis") * " SDM with Middens — Years BP: " * .(current_year_bp))) +  # title
    theme_minimal(base_size = 11) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "#F2F2F2", color = NA),
      panel.grid.major = element_line(color = "#C4C4C4"),
      panel.grid.minor = element_line(color = "#C4C4C4"),
      plot.margin = margin(0, 0, 0, 0, "pt")
    )

  plot_path <- file.path(results_dir, "midden maps", paste0("apicalis_middens_", current_year_bp, ".png"))  # out path
  dir.create(dirname(plot_path), showWarnings = FALSE, recursive = TRUE)                                     # ensure dir
  ggsave(plot_path, p, width = 7, height = 7, dpi = 300, bg = "white")                                      # save png

  img <- magick::image_read(plot_path) |> magick::image_trim() |> image_border(color = "white", geometry = "30x30")  # tidy img
  magick::image_write(img, path = plot_path)                                                                     # overwrite
})
```

# 6-panel with middens
```{r six-panel-middens}
raster_times <- terra::time(apicalis_spatrast)                                           # time values
target_times <- c(-28000, -15000, -12000, -7000, -4000, 0)                               # selected years
matched_indices <- match(target_times, raster_times)                                     # matching indices
panel_tags <- letters[seq_along(target_times)]                                           # labels

panel_plots <- pmap(                                                                      # loop panels
  list(idx = matched_indices, year = target_times, tag = panel_tags),
  function(idx, year_bp, tag_label) {
    current_prediction <- apicalis_spatrast[[idx]]                                       # prediction layer
    current_middens <- middens_sf %>% dplyr::filter(abs(time_bp - year_bp) <= 500) %>%   # nearby middens
      dplyr::mutate(type = "Midden")

    ggplot() +
      tidyterra::geom_spatraster(data = current_prediction, aes(fill = mean)) +          # raster
      tidyterra::scale_fill_terrain_c(name = "Probability\nof occurrence", limits = c(0, 1)) +
      geom_sf(data = australia_outline, fill = NA, color = "black", linewidth = 0.5) +   # outline
      geom_sf(data = current_middens, aes(color = type, shape = type), fill = "#FF3399", size = 2, stroke = 0.4, show.legend = TRUE) +
      scale_color_manual(name = NULL, values = c("Midden" = "black")) +
      scale_shape_manual(name = NULL, values = c("Midden" = 21)) +
      coord_sf(xlim = xlim, ylim = ylim) +                                               # window
      scale_y_continuous(breaks = seq(-40, -10, by = 4)) +
      scale_x_continuous(breaks = seq(112, 150, by = 8)) +
      ggtitle(bquote("(" * .(tag_label) * ")  Years BP: " * .(abs(year_bp)))) +         # title
      theme(
        plot.background = element_rect(fill = "white", color = NA),
        panel.background  = element_rect(fill = "#E8E8E8", color = NA),
        panel.grid.major  = element_line(color = "#C4C4C4"),
        panel.grid.minor  = element_line(color = "#C4C4C4"),
        axis.text         = element_text(size = 8),
        axis.title        = element_blank(),
        plot.margin       = margin(2, 2, 2, 2),
        legend.position   = "bottom",
        legend.box        = "horizontal",
        legend.justification = "center",
        legend.spacing.x  = unit(1.2, "cm"),
        legend.spacing.y  = unit(0.4, "cm"),
        legend.key        = element_rect(fill = "transparent", color = NA),
        legend.title      = element_text(size = 11),
        legend.text       = element_text(size = 11)
      ) +
      guides(
        fill = guide_colorbar(title = "Probability of occurrence", title.position = "top",
                              title.theme = element_text(size = 11), label.theme = element_text(size = 11),
                              direction = "horizontal", barwidth = unit(5, "cm"), barheight = unit(0.4, "cm")),
        color = guide_legend(override.aes = list(size = 2.5), title.theme = element_text(size = 11), label.theme = element_text(size = 11)),
        shape  = guide_legend(title.theme = element_text(size = 11), label.theme = element_text(size = 11))
      )
  }
)

final_panel <- wrap_plots(panel_plots, ncol = 2, guides = "collect") +
  plot_annotation(title = "Leporillus apicalis SDM with middens",
                  theme = theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))) &
  theme(legend.position = "bottom", legend.box = "horizontal", legend.justification = "center",
        legend.spacing.x = unit(1.2, "cm"), legend.spacing.y = unit(0.4, "cm"), legend.key = element_rect(fill = "transparent", color = NA))

ggsave(file.path(results_dir, "midden_panel_plot.png"), final_panel, width = 10, height = 9, dpi = 300)
```

# Threshold, pseudo-absences (time-aware), and evaluation
```{r tss-maxsss}
# Initial threshold for bootstrapping pseudo-absences
threshold_init <- 0.27
apicalis_binary_init <- terra::app(x = apicalis_spatrast, fun = \(i) i > threshold_init)     # binary stack @ init thr
terra::time(apicalis_binary_init) <- terra::time(apicalis_spatrast) |> pastclim::ybp2date()  # time axis as Date
apicalis_rs_init <- terra::sds(apicalis_binary_init); names(apicalis_rs_init) <- "pres_abs"  # sds for extraction

# Time-aware pseudoabsence sampling around dated middens
pseudoabs_init <- sample_pseudoabs_time(
  data            = dated_middens,
  raster          = apicalis_rs_init,
  n_per_presence  = 3,
  time_col        = "time_bp",
  lubridate_fun   = pastclim::ybp2date,
  method          = c("dist_min", km2m(200)),
  time_buffer     = 365000 * 2
)

# Extract continuous predictions matched to each row's nearest timestep
rt <- as.numeric(terra::time(apicalis_spatrast))

presence_sf <- sf::st_as_sf(dated_middens, coords = c("longitude", "latitude"), crs = 4326) |>
  sf::st_transform(crs(apicalis_spatrast)) |>
  dplyr::mutate(nearest_time = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
                nearest_idx  = match(nearest_time, rt))

pres_mat <- terra::extract(apicalis_spatrast, terra::vect(presence_sf), ID = FALSE)
presence_vals <- as.matrix(pres_mat)[cbind(seq_len(nrow(presence_sf)), presence_sf$nearest_idx)]

pseudoabs_sf <- sf::st_as_sf(pseudoabs_init, coords = c("longitude", "latitude"), crs = 4326) |>
  sf::st_transform(crs(apicalis_spatrast)) |>
  dplyr::mutate(time_bp = pastclim::date2ybp(time_step),
                nearest_time = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
                nearest_idx  = match(nearest_time, rt))

pseudo_mat <- terra::extract(apicalis_spatrast, terra::vect(pseudoabs_sf), ID = FALSE)
pseudoabs_vals <- as.matrix(pseudo_mat)[cbind(seq_len(nrow(pseudoabs_sf)), pseudoabs_sf$nearest_idx)]

predicted_values <- c(presence_vals, pseudoabs_vals)                                    # continuous predictions
observed <- c(rep(1, length(presence_vals)), rep(0, length(pseudoabs_vals)))            # labels
keep <- is.finite(predicted_values) & is.finite(observed)                                # keep finite
predicted_values <- predicted_values[keep]; observed <- observed[keep]

# MaxSSS grid search
thresholds <- seq(0, 1, by = 0.01)
results <- data.frame(threshold = thresholds, TSS = NA_real_)
for (i in seq_along(thresholds)) {
  thr <- thresholds[i]
  pred_binary <- ifelse(predicted_values > thr, 1L, 0L)
  a <- sum(pred_binary == 1L & observed == 1L)
  b <- sum(pred_binary == 1L & observed == 0L)
  c <- sum(pred_binary == 0L & observed == 1L)
  d <- sum(pred_binary == 0L & observed == 0L)
  if ((a + c) == 0 || (b + d) == 0) next
  sens <- a / (a + c); spec <- d / (b + d)
  results$TSS[i] <- sens + spec - 1
}
opt_threshold <- results$threshold[which.max(results$TSS)]
cat("Optimal Threshold (maxSSS):", opt_threshold, "\n")

# Rebuild binary stack at optimal threshold
apicalis_binary <- terra::app(x = apicalis_spatrast, fun = \(i) i > opt_threshold)
terra::time(apicalis_binary) <- terra::time(apicalis_spatrast) |> pastclim::ybp2date()
apicalis_rs <- terra::sds(apicalis_binary); names(apicalis_rs) <- "pres_abs"

# Re-sample time-aware pseudoabsences at the optimised threshold backdrop
pseudoabs <- sample_pseudoabs_time(
  data            = dated_middens,
  raster          = apicalis_rs,
  n_per_presence  = 3,
  time_col        = "time_bp",
  lubridate_fun   = pastclim::ybp2date,
  method          = c("dist_min", km2m(200)),
  time_buffer     = 365000 * 2
)

# Evaluate (TSS, Sørensen, F2)
pres_abs_observed <- pseudoabs |>
  dplyr::mutate(time_bp = pastclim::date2ybp(time_step)) |>
  dplyr::select(-class, -time_step) |>
  tibble::rowid_to_column("name")

pres_abs_pred <- location_slice_from_region_series(x = pres_abs_observed, region_series = apicalis_rs)
pres_abs_pred$class <- pseudoabs$class

a <- nrow(dplyr::filter(pres_abs_pred, class == "presence",  pres_abs == 1))
b <- nrow(dplyr::filter(pres_abs_pred, class == "pseudoabs", pres_abs == 1))
c <- nrow(dplyr::filter(pres_abs_pred, class == "presence",  pres_abs == 0))
d <- nrow(dplyr::filter(pres_abs_pred, class == "pseudoabs", pres_abs == 0))

tpr <- a / (a + c); tnr <- d / (b + d)
tss <- tpr + tnr - 1
sorensen <- 2 * (a / (c + (2 * a) + d))
precision <- a / (a + b)
f2score <- 2 * ((precision * tpr) / (precision + tpr))

cat(sprintf("True Skill Statistic: %.3f\n", tss))
cat(sprintf("Sorensen: %.3f\n", sorensen))
cat(sprintf("F2 score: %.3f\n", f2score))
```

# Panels (a) & (b): histograms (full vs coarse) and persistence
```{r hist-panels}
suppressPackageStartupMessages({ library(cowplot) })
stopifnot(inherits(apicalis_spatrast, "SpatRaster"), inherits(middens_sf, "sf"), dir.exists(results_dir))

step_years <- 1000
rt <- as.numeric(terra::time(apicalis_spatrast))
thr_a <- if (exists("opt_threshold")) opt_threshold else 0.27
thr_b <- 0.35
lookback_ka <- 30
include_current <- TRUE
coarse_factor <- 2                    # with 1° base, coarse = 2°
w_single <- 7; h_single <- 5; dpi_out <- 300

res_full_deg   <- terra::res(apicalis_spatrast)[1]
res_coarse_deg <- res_full_deg * coarse_factor

ensure_time_cols <- function(m_sf, rt) {
  if (!all(c("binned_time_bp","nearest_idx") %in% names(m_sf))) {
    m_sf <- m_sf %>% dplyr::mutate(
      binned_time_bp = sapply(time_bp, function(x) rt[which.min(abs(rt - x))]),
      nearest_idx    = match(binned_time_bp, rt)
    )
  }
  m_sf
}

fd_bin_count <- function(x) {
  x <- x[is.finite(x)]; if (!length(x)) return(30L)
  fd_bw <- 2 * IQR(x, na.rm = TRUE) / (length(x)^(1/3))
  max(20, min(60, round(1 / max(min(fd_bw, 0.08), 0.02))))
}

time_matched_background <- function(m_sf, rast){
  rt <- as.numeric(terra::time(rast))
  m_sf <- ensure_time_cols(m_sf, rt)
  split(m_sf, m_sf$binned_time_bp) |>
    lapply(function(df_i){
      tb  <- unique(df_i$binned_time_bp)
      lyr <- match(tb, rt)
      n_i <- nrow(df_i)
      vals <- terra::spatSample(rast[[lyr]], size = n_i, method = "random", values = TRUE, na.rm = TRUE)[,1] |> as.numeric()
      tibble::tibble(suitability = vals, group = "Background (time-matched)", bin = tb)
    }) |>
    dplyr::bind_rows()
}

compute_persistence_strict <- function(rast, m_sf, thr, include_current, lookback_ka, step_years){
  rt <- as.numeric(terra::time(rast))
  m_sf <- ensure_time_cols(m_sf, rt)
  mat <- terra::extract(rast, terra::vect(m_sf), ID = FALSE) |> as.matrix()
  lookback_steps <- if (is.finite(lookback_ka)) as.integer(lookback_ka) else Inf
  start_idx <- if (!include_current) pmax(m_sf$nearest_idx - 1L, 0L) else m_sf$nearest_idx
  lead_true_count <- function(x){ n <- 0L; for (v in x) { if (isTRUE(v)) n <- n + 1L else break }; n }
  steps <- vapply(seq_len(nrow(mat)), function(i){
    idx <- start_idx[i]; if (idx <= 0L) return(0L)
    from <- idx; to <- if (is.finite(lookback_steps)) max(idx - lookback_steps + 1L, 1L) else 1L
    sidx <- seq(from, to, by = -1L)
    above <- mat[i, sidx] > thr
    lead_true_count(above)
  }, integer(1))
  tibble::tibble(persist_ka = steps * (step_years/1000))
}

middens_sf <- ensure_time_cols(middens_sf, rt)
apicalis_spatrast_coarse <- terra::aggregate(apicalis_spatrast, fact = coarse_factor, fun = mean, na.rm = TRUE)
terra::time(apicalis_spatrast_coarse) <- terra::time(apicalis_spatrast)

mid_df_full <- tibble::tibble(suitability = sf::st_drop_geometry(middens_sf)$predicted, group = "Middens", bin = middens_sf$binned_time_bp)
bg_df_full  <- time_matched_background(middens_sf, apicalis_spatrast)
plot_df_full <- dplyr::bind_rows(mid_df_full, bg_df_full) |> dplyr::filter(is.finite(suitability))
bins_a_full <- fd_bin_count(mid_df_full$suitability)

p_a_full <- ggplot(plot_df_full, aes(x = suitability, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.65, bins = bins_a_full, boundary = 0, color = "white") +
  geom_vline(xintercept = thr_a, linetype = "dashed", linewidth = 0.6) +
  scale_fill_manual(values = c("Middens" = "#BC43FF", "Background (time-matched)" = "#9AA0A6")) +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,0.1)) +
  labs(title = sprintf("Habitat suitability at dated midden locations (%.1f°)", terra::res(apicalis_spatrast)[1]), x = "Modelled suitability (0–1)", y = "Count", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA), panel.background = element_rect(fill = "#F2F2F2", color = NA), panel.grid.major = element_line(color = "white"), legend.position = "bottom")

ggsave(file.path(results_dir, "panel_a_full_hist.png"), p_a_full, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

mid_vals_c <- terra::extract(apicalis_spatrast_coarse, terra::vect(middens_sf), ID = FALSE) |> as.matrix()
pick_idx   <- cbind(seq_len(nrow(mid_vals_c)), middens_sf$nearest_idx)
mid_suit_c <- mid_vals_c[pick_idx] |> as.numeric()
mid_df_c <- tibble::tibble(suitability = mid_suit_c, group = "Middens", bin = middens_sf$binned_time_bp)
bg_df_c  <- time_matched_background(middens_sf, apicalis_spatrast_coarse)
plot_df_c <- dplyr::bind_rows(mid_df_c, bg_df_c) |> dplyr::filter(is.finite(suitability))
bins_a_c  <- fd_bin_count(mid_df_c$suitability)

p_a_coarse <- ggplot(plot_df_c, aes(x = suitability, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.65, bins = bins_a_c, boundary = 0, color = "white") +
  geom_vline(xintercept = thr_a, linetype = "dashed", linewidth = 0.6) +
  scale_fill_manual(values = c("Middens" = "#BC43FF", "Background (time-matched)" = "#9AA0A6")) +
  scale_x_continuous(limits = c(0,1), breaks = seq(0,1,0.1)) +
  labs(title = sprintf("Habitat suitability at dated midden locations (%.1f°)", terra::res(apicalis_spatrast_coarse)[1]), x = "Modelled suitability (0–1)", y = "Count", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA), panel.background = element_rect(fill = "#F2F2F2", color = NA), panel.grid.major = element_line(color = "white"), legend.position = "bottom")

ggsave(file.path(results_dir, "panel_a_coarse_hist.png"), p_a_coarse, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

persist_full <- compute_persistence_strict(apicalis_spatrast, middens_sf, thr = thr_b, include_current = include_current, lookback_ka = lookback_ka, step_years = step_years)
x_max_b <- if (is.finite(lookback_ka)) lookback_ka else max(persist_full$persist_ka, na.rm = TRUE)

p_b_full <- ggplot(persist_full, aes(x = persist_ka)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left", color = "white", fill = "#BC43FF", alpha = 0.9) +
  scale_x_continuous(limits = c(0, x_max_b), breaks = seq(0, x_max_b, by = 10), labels = function(x) paste0(x, " ka")) +
  labs(title = sprintf("Duration of suitability prior to midden deposition (%.1f°)", terra::res(apicalis_spatrast)[1]), x = sprintf("Years suitable before deposition (ka; suitability \u2265 %.2f)", thr_b), y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA), panel.background = element_rect(fill = "#F2F2F2", color = NA), panel.grid.major = element_line(color = "white"))

ggsave(file.path(results_dir, "panel_b_full_hist_thr035.png"), p_b_full, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

persist_coarse <- compute_persistence_strict(apicalis_spatrast_coarse, middens_sf, thr = thr_b, include_current = include_current, lookback_ka = lookback_ka, step_years = step_years)

p_b_coarse <- ggplot(persist_coarse, aes(x = persist_ka)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left", color = "white", fill = "#BC43FF", alpha = 0.9) +
  scale_x_continuous(limits = c(0, x_max_b), breaks = seq(0, x_max_b, by = 10), labels = function(x) paste0(x, " ka")) +
  labs(title = sprintf("Duration of suitability prior to midden deposition (%.1f°)", terra::res(apicalis_spatrast_coarse)[1]), x = sprintf("Years suitable before deposition (ka; suitability \u2265 %.2f)", thr_b), y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.background = element_rect(fill = "white", color = NA), panel.background = element_rect(fill = "#F2F2F2", color = NA), panel.grid.major = element_line(color = "white"))

ggsave(file.path(results_dir, "panel_b_coarse_hist_thr035.png"), p_b_coarse, width = w_single, height = h_single, dpi = dpi_out, bg = "white")

p_a_tag <- cowplot::ggdraw(p_a_coarse + labs(title = NULL)) +
  cowplot::draw_label("(a)", x = 0.02, y = 0.98, hjust = 0, vjust = 1, fontface = "bold")
p_b_tag <- cowplot::ggdraw(p_b_coarse + labs(title = NULL)) +
  cowplot::draw_label("(b)", x = 0.02, y = 0.98, hjust = 0, vjust = 1, fontface = "bold")

fig_ab <- cowplot::plot_grid(p_a_tag, p_b_tag, ncol = 1, rel_heights = c(1,1))
print(fig_ab)

ggsave(file.path(results_dir, "figure_histograms_panels_ab_COARSE_thrB_035_VERTICAL.png"), fig_ab,
       width = 7, height = 10, dpi = dpi_out, bg = "white")
```

# Variable importance table + PDPs
```{r vip-pdp}
suppressPackageStartupMessages({ library(DALEX); library(tidysdm); library(ggplot2); library(dplyr); library(knitr) })
apicalis_explainer <- explain_tidysdm(apicalis_ensemble)                                                           # explainer
apicalis_outdir <- file.path(results_dir, "variable_importance"); dir.create(apicalis_outdir, showWarnings = FALSE, recursive = TRUE)

apicalis_vip <- model_parts(explainer = apicalis_explainer) |>                                                     # permutation VI
  dplyr::filter(!variable %in% c("_baseline_", "_full_model_")) |>
  dplyr::group_by(variable) |>
  dplyr::summarise(mean_dropout_loss = round(mean(dropout_loss, na.rm = TRUE), 3),
                   sd_dropout_loss   = round(sd(dropout_loss,   na.rm = TRUE), 3), .groups = "drop_last") |>
  dplyr::arrange(desc(mean_dropout_loss)) |>
  dplyr::rename(`Environmental variable` = variable, `Mean drop-out loss` = mean_dropout_loss, `SD` = sd_dropout_loss)

print(apicalis_vip)                                                                                                 # preview table
write.csv(apicalis_vip, file.path(apicalis_outdir, "apicalis_variable_importance_table.csv"), row.names = FALSE)   # save CSV
knitr::kable(apicalis_vip, caption = "Variable importance summary (Leporillus apicalis ensemble model)", format = "markdown")

plot_and_save_pdp <- function(var, label){
  pdp <- model_profile(apicalis_explainer, N = 500, variables = var)                                              # PDP
  p <- ggplot(pdp$agr_profiles, aes(x = `_x_`, y = `_yhat_`)) +
    geom_line(color = "darkblue", linewidth = 1) +
    labs(title = paste("Leporillus apicalis – Partial Dependence on", label), x = label, y = "Predicted suitability") +
    theme_minimal(base_size = 12) +
    theme(plot.background = element_rect(fill = "white", color = NA), panel.background = element_rect(fill = "#F2F2F2", color = NA), panel.grid.major = element_line(color = "#C4C4C4", linewidth = 0.3), panel.border = element_rect(color = "#C4C4C4", fill = NA, linewidth = 0.3), plot.title = element_text(face = "italic", size = 13))
  ggsave(file.path(apicalis_outdir, paste0("apicalis_pdp_", var, ".png")), p, width = 6, height = 4, dpi = 300, bg = "white")
  print(p)
}
plot_and_save_pdp("bio05", "Maximum temperature of warmest month (°C)")
plot_and_save_pdp("bio06", "Minimum temperature of coldest month (°C)")
plot_and_save_pdp("bio12", "Annual precipitation (mm)")
```

# KDE of midden sampling density (optional viz)
```{r kde-middens}
suppressPackageStartupMessages({ library(rnaturalearth); library(rnaturalearthdata) })
# Use middens_sf created above (not middens_sf_2) to avoid undefined object.
middens_df_2 <- middens_sf %>%
  dplyr::mutate(
    longitude = sf::st_coordinates(.)[, 1],
    latitude  = sf::st_coordinates(.)[, 2]
  ) %>%
  sf::st_drop_geometry()

australia_outline_ne <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>% dplyr::filter(admin == "Australia")

pad_points <- data.frame(longitude = c(110, 150, 150, 110), latitude = c(-10, -10, -45, -45))
middens_df_padded <- dplyr::bind_rows(middens_df_2, pad_points)

p <- ggplot() +
  geom_sf(data = australia_outline_ne, fill = scales::alpha("#C4C4C4", 0.5), color = "black") +
  stat_density_2d(data = middens_df_padded, aes(x = longitude, y = latitude, fill = after_stat(level)),
                  geom = "polygon", contour = TRUE, color = NA, adjust = 0.4, alpha = 0.8, n = 200) +
  scale_fill_viridis_c(option = "magma", name = "Density", direction = -1) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10), expand = FALSE) +
  labs(title = "Spatial Density of Dated Midden Samples", x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "#F2F2F2", color = NA),
    panel.grid.major = element_line(color = "#C4C4C4"),
    panel.grid.minor = element_line(color = "#C4C4C4"),
    legend.position  = c(0.03, 0.20),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color = "grey70", linewidth = 0.3),
    legend.key = element_rect(fill = "white", color = NA),
    legend.margin = margin(6, 12, 6, 9),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9)
  ) +
  guides(fill = guide_colorbar(title = "Density", title.position = "top", title.theme = element_text(size = 10),
                               label.theme = element_text(size = 9), direction = "horizontal",
                               barwidth = unit(4, "cm"), barheight = unit(0.4, "cm"),
                               ticks = TRUE, ticks.color = "black", frame.color = "grey70", frame.linewidth = 0.3,
                               label.position = "bottom"))

ggsave(file.path(here("Results/stratified environmental sampling"), "midden_sampling_density.png"), p, width = 7, height = 6, dpi = 300, bg = "white")
```

# Citations dump (optional)
```{r citations, eval=FALSE}
lapply(c("tidysdm", "tidymodels", "workflowsets", "workflows", "tune", "parsnip", "recipes", "yardstick", "rsample",
         "spatialsample", "dials", "dplyr", "tidyr", "purrr", "tibble", "ggplot2", "cowplot", "scales", "viridis",
         "viridisLite", "sf", "terra", "tidyterra", "pastclim", "oz", "rworldmap", "DALEX", "ecospat", "broom",
         "infer", "readr", "fs", "gtools", "RefManageR"),
       function(pkg) { cat("\n---", pkg, "---\n"); print(citation(pkg)) })
```
